---
title: "Bad Controls: Peer Bias"
subtitle: ""
description: ""
author: 
  - name: "Dan Swart, CPA (ret)"
    affiliations:
      - "SCUC-ISD Board of Trustees"
      - "Seat 6"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    code-copy: true
    code-link: true
    include-in-header: header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    df-print: paged
    code-overflow: wrap
    toc: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 14pt
    mainfont: "Cabin"
    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false

---





```{r}
#| label: setup
#| include: false

# Prevent scientific notation globally
options(scipen = 999)

# load libraries
library(readxl)
library(writexl) 
library(tidyr) 
library(readr) 
library(purrr) 
library(stringr) 
library(forcats) 
library(lubridate) 
library(janitor) 
library(ggtext) 
library(paletteer) 
library(viridis) 
library(RColorBrewer) 
library(wesanderson) 
library(dutchmasters) 
library(gghighlight) 
library(monochromeR) 
library(ggforce) 
library(ggthemes)
library(gt)
library(gtExtras)
library(tidyverse)
library(plotly)
library(ggplot2)
library(tibble)
library(patchwork)
library(ppcor)
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(car)
library(WRS2)
library(boot)
library(BayesFactor)
library(pwr)
library(qgraph)
library(scales)
library(here)
library(ggdag)      # For plotting DAGs
library(dagitty)    # For working with DAG logic
library(DiagrammeR) # For complete control of the layout
library(knitr)      # For controlling rendering
library(kableExtra) # For tables summarizing results
library(DT)         # For rendering content that kableExtra cannot (symbols)



# Set global theme for consistent plots
theme_set(theme_minimal(base_size = 20) + 
          theme(
    plot.title = element_text(face = "bold", size = 26),    # adjust title size
    plot.subtitle = element_text(face = "bold", size = 24), # adjust subtitle size
    axis.title.x = element_text(face = "bold", size = 22),
    axis.title.y = element_text(face = "bold", size = 22),
    axis.text.x = element_text(face = "bold", size = 22, angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.spacing.x = unit(1.5, "cm"),  # Horizontal spacing only
    panel.spacing.y = unit(1.5, "cm"),   # Vertical spacing only
    plot.margin = margin(20, 20, 20, 20, "pt")
    )
)


# Set seed for reproducibility
set.seed(123)

```


[Watch YouTube video](https://www.youtube.com/watch?v=NSuTaeW6Orc)


<br>


## DAG RENDERING USING DiagrammeR.

(There is no analysis with DiagrammeR, but analysis follows below.)

```{r peer-bias-dag8}
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10


library(DiagrammeR)

grViz("
  digraph DAG {
    # Graph settings
    graph [layout=neato, margin=\"1.0, 1.0, 2.0, 1.0\", rankdir=TB, size=\"14,12\"]  
    
    # Add a title using a simple label approach
    labelloc=\"t\"
    label=\"Bad Controls:  Peer Bias\\n \\n\"      fontname=\"Cabin\" fontcolor=\"darkgreen\"
    fontsize=26
    
    # Node settings - make nodes larger with fontsize
    node [shape=plaintext, fontsize=26, fontname=\"Cabin\"] # Increase
    
    # Edge settings - make edges thicker and arrows larger
    edge [penwidth=4.0, color=\"darkblue\", arrowsize=1.5] # Increase
    
    # Nodes with exact coordinates
    X [label=\"X\", pos=\"1.0, 1.0!\", fontcolor=\"dodgerblue\"]
    Y [label=\"Y\", pos=\"4.0, 1.0!\", fontcolor=\"dodgerblue\"]
    E [label=\"E\", pos=\"2.5, 3.0!\", fontcolor=\"black\"]
    Q [label=\"Q\", pos=\"4.0, 3.0!\", fontcolor=\"darkpurple\"]
    
    
    # Edges
    X -> Y
    X -> E
    E -> Y
    Q -> Y
    Q -> E
    
    # Caption as a separate node at the bottom
    Caption [shape=plaintext, label=\"Cinelli, Forney, Pearl 2021 A Crash\\nCourse in Good and Bad Controls\", 
             fontsize=20, pos=\"2.5,0.0!\"]
  }
  ")

```

<br>


### DAG Visualization using ggdag and dagitty

```{r complex-structure-dag1}
#| fig-cap: "Cinelli, Forney, Pearl 2021 A Crash Course in Good and Bad Controls"
#| fig-width: 12
#| fig-height: 10


# Define the DAG
peer_bias_dag8 <- ggdag::dagify(
  Y ~ X,   # Y is influenced by X
  E ~ X,
  Y ~ E,  
  E ~ Q,
  Y ~ Q,
  exposure = "X",
  outcome = "Y",
  # Add labels here:
  labels = c(X = "X", 
             Y = "Y", 
             E = "E",
             Q = "Q"),
  coords = list(x = c(X = 1.0, Y = 4.0, E = 2.5, Q = 4.0),  
                y = c(X = 1.0, Y = 1.0, E = 3.0, Q = 3.0))
)

# Create a nice visualization of the DAG
ggdag_status(peer_bias_dag8) + 
  theme_dag(base_size = 18) +
  labs(title = "Bad Controls: Peer Bias")

```

## Executive Summary: Peer Bias as a Bad Control

::: {.callout-note collapse="true" title="<span style='font-size: 20px;'>What is Peer Bias?</span> <span style='color: darkblue; font-size: 22px;'>(click to open and close)</span>"}


Peer bias occurs when we adjust for a variable E that is influenced by both the exposure X and an unmeasured confounder Q, which also affects the outcome Y. In this DAG structure, E is a mediator between X and Y, but is also affected by the confounder Q that directly affects Y.

#### Why is it a "Bad Control"?

Controlling for E in this structure is harmful because:

1. **It blocks part of the causal effect**: By conditioning on E, we're blocking the indirect effect of X on Y that flows through E.

2. **It opens a collider path**: Conditioning on E opens a non-causal path between X and Y through Q (X → E ← Q → Y), potentially creating bias.

3. **It can distort the total effect**: The adjustment might lead to estimates that don't reflect the true causal relationship between X and Y.

#### Real-World Example

A researcher is studying the effect of a new teaching method (X) on student final exam scores (Y):

- The teaching method (X) affects student engagement (E).
- Student engagement (E) affects final exam scores (Y).
- Student natural ability (Q) affects both engagement (E) and exam scores (Y).
- The teaching method (X) also has a direct effect on exam scores (Y).

If the researcher controls for student engagement (E), they block the indirect effect of the teaching method (X) through engagement (E) and potentially introduce bias through the opened collider path.

#### How to Avoid Peer Bias

1. **Consider total effects carefully**: Determine whether you're interested in the total effect (direct + indirect) or just the direct effect of X on Y.

2. **Be cautious with mediators**: Think carefully before adjusting for variables that lie on the causal pathway between exposure and outcome.

3. **Account for unmeasured confounders**: Consider the possibility of unmeasured variables that might affect both your mediators and outcomes.

4. **Use appropriate causal inference methods**: Methods like mediation analysis can help decompose direct and indirect effects properly.

Peer bias demonstrates the importance of carefully considering the causal structure before deciding which variables to control for in your analysis.

:::

```{r}
#| message: false
#| warning: false
#| code-fold: false
#| echo: false

# Create a function to display DAG analysis results as a table
display_dag_analysis <- function(dag) {
  # Initialize results list
  results <- list()
  
  # 1. Get the implied conditional independencies
  results$independencies <- tryCatch({
    dagitty::impliedConditionalIndependencies(dag)
  }, error = function(e) {
    "None found"
  })
  
  # 2. Find all valid adjustment sets
  results$adjustment_sets <- tryCatch({
    dagitty::adjustmentSets(dag)
  }, error = function(e) {
    list()
  })
  
  # 3. Find minimal sufficient adjustment sets
  results$minimal_adjustment_sets <- tryCatch({
    dagitty::adjustmentSets(dag, type = "minimal")
  }, error = function(e) {
    list()
  })
  
  # 4. Identify paths between exposure and outcome
results$paths <- tryCatch({
  # Change this line to specify directed=TRUE
  dagitty::paths(dag, from = "X", to = "Y", directed = TRUE)
}, error = function(e) {
  data.frame(paths = character(0), length = numeric(0))
})
  
  # 5. Find instrumental variables
  results$instruments <- tryCatch({
    dagitty::instrumentalVariables(dag, exposure = "X", outcome = "Y")
  }, error = function(e) {
    NULL
  })
  
  # 6. Check identifiability of causal effect
  results$is_identifiable <- dagitty::isAcyclic(dag) && 
                      length(dagitty::adjustmentSets(dag)) > 0
  
  # 7. Find ancestors and descendants
  results$X_ancestors <- dagitty::ancestors(dag, "X")
  results$X_descendants <- dagitty::descendants(dag, "X")
  results$Y_ancestors <- dagitty::ancestors(dag, "Y")
  results$Y_descendants <- dagitty::descendants(dag, "Y")
  results$E_ancestors <- dagitty::ancestors(dag, "E")
  results$E_descendants <- dagitty::descendants(dag, "E")
  results$Q_ancestors <- dagitty::ancestors(dag, "Q")
  results$Q_descendants <- dagitty::descendants(dag, "Q")
  
  # 8. Check backdoor paths
  results$backdoor_paths <- character(0)
  if(is.data.frame(results$paths) && nrow(results$paths) > 0) {
    for(i in 1:nrow(results$paths)) {
      path_str <- results$paths$paths[i]
      path_elements <- strsplit(path_str, " ")[[1]]
      
      # A backdoor path has an arrow pointing into the exposure
      if(length(path_elements) >= 3) {
        second_element <- path_elements[2]
        if(second_element == "<-") {
          results$backdoor_paths <- c(results$backdoor_paths, path_str)
        }
      }
    }
  }
  
  # 9. Find directed paths (potential mediation)
  results$directed_paths <- tryCatch({
    dagitty::paths(dag, from = "X", to = "Y", directed = TRUE)
  }, error = function(e) {
    data.frame(paths = character(0), length = numeric(0))
  })
  
  results$mediators <- character(0)
  if(is.data.frame(results$directed_paths) && nrow(results$directed_paths) > 0) {
    for(i in 1:nrow(results$directed_paths)) {
      path_str <- results$directed_paths$paths[i]
      path_elements <- strsplit(path_str, " ")[[1]]
      
      # Extract variables (every other element)
      path_vars <- path_elements[seq(1, length(path_elements), by = 2)]
      
      # Variables between X and Y are mediators
      if(length(path_vars) > 2) {
        potential_mediators <- path_vars[-c(1, length(path_vars))]
        results$mediators <- c(results$mediators, potential_mediators)
      }
    }
    results$mediators <- unique(results$mediators)
  }
  
  # 10. Test d-separation
  results$d_sep_results <- list(
    XY_given_nothing = dagitty::dseparated(dag, "X", "Y", c()),
    XY_given_E = dagitty::dseparated(dag, "X", "Y", c("E")),
    XY_given_Q = dagitty::dseparated(dag, "X", "Y", c("Q")),
    XY_given_EQ = dagitty::dseparated(dag, "X", "Y", c("E", "Q"))
  )
  
  # 11. Check paths under different adjustments
  results$adjustment_effects <- list()
  adjustment_sets_to_check <- list(
    "None" = c(),
    "E" = c("E"),
    "Q" = c("Q"),
    "E and Q" = c("E", "Q")
  )
  
  for(adj_name in names(adjustment_sets_to_check)) {
    adj_set <- adjustment_sets_to_check[[adj_name]]
    paths <- tryCatch({
      dagitty::paths(dag, from = "X", to = "Y")
    }, error = function(e) {
      data.frame(paths = character(0), length = numeric(0))
    })
    
    if(is.data.frame(paths) && nrow(paths) > 0) {
      open_paths <- tryCatch({
        dagitty::paths(dag, from = "X", to = "Y", Z = adj_set)
      }, error = function(e) {
        data.frame(paths = character(0), length = numeric(0))
      })
      
      results$adjustment_effects[[adj_name]] <- list(
        "total_paths" = nrow(paths),
        "open_paths" = if(is.data.frame(open_paths)) nrow(open_paths) else 0
      )
    } else {
      results$adjustment_effects[[adj_name]] <- list(
        "total_paths" = 0,
        "open_paths" = 0
      )
    }
  }
  
  return(results)
}

```

```{r run-the-analysis}
#| include: true
#| echo: false
#| results: 'hide'
#| code-fold: false


# Run the analysis
dag_results <- display_dag_analysis(peer_bias_dag8)

# Create tables for presentation, but don't print them

# Table 1: Key DAG Properties
properties_df <- data.frame(
  Property = c(
    "Acyclic DAG", 
    "Causal effect identifiable",
    "Number of paths from X to Y",
    "Number of backdoor paths",
    "Direct effect of X on Y exists",
    "Potential mediators",
    "Number of adjustment sets",
    "Minimal adjustment sets"
  ),
  Value = c(
    ifelse(dagitty::isAcyclic(peer_bias_dag8), "Yes", "No"),
    ifelse(dag_results$is_identifiable, "Yes", "No"),
    if(is.data.frame(dag_results$paths)) nrow(dag_results$paths) else 0,
    length(dag_results$backdoor_paths),
    ifelse("X" %in% dagitty::parents(peer_bias_dag8, "Y"), "Yes", "No"),
    ifelse(length(dag_results$mediators) > 0, paste(dag_results$mediators, collapse=", "), "None"),
    length(dag_results$adjustment_sets),
    ifelse(length(dag_results$minimal_adjustment_sets) > 0, 
           paste(sapply(dag_results$minimal_adjustment_sets, function(x) paste(x, collapse=", ")), collapse="; "), 
           "None")
  )
)

```

```{r}
#| label: independencies-df
#| tbl-cap: "Implied Conditional Independencies"
#| results: 'asis'
#| code-fold: false
#| echo: false

# this chunk only creates a data frame but doesn't display it

# Table 2: Conditional Independencies

if(length(dag_results$independencies) > 0) {
  independencies_df <- data.frame(
    Index = 1:length(dag_results$independencies),
    Independencies = sapply(dag_results$independencies, function(x) paste(x, collapse=" "))
  )
} else {
  independencies_df <- data.frame(
    Index = 1,
    Independencies = "No conditional independencies found"
  )
}

```


```{r}
#| label: create-paths-df
#| echo: false
#| include: true
#| results: 'hide'

# this chunk only creates a data frame but doesn't display it

# Table 3: Paths Analysis
if(is.data.frame(dag_results$paths) && nrow(dag_results$paths) > 0) {
  paths_df <- data.frame(
    Path = dag_results$paths$paths,
    Length = dag_results$paths$length,
    IsBackdoor = sapply(dag_results$paths$paths, function(p) {
      elements <- strsplit(p, " ")[[1]]
      if(length(elements) >= 3) {
        return(elements[2] == "<-")
      }
      return(FALSE)
    }),
    IsDirected = sapply(dag_results$paths$paths, function(p) {
      elements <- strsplit(p, " ")[[1]]
      all_forward <- TRUE
      for(i in seq(2, length(elements), by=2)) {
        if(elements[i] != "->") {
          all_forward <- FALSE
          break
        }
      }
      return(all_forward)
    })
  )
} else {
  paths_df <- data.frame(
    Path = "No paths found",
    Length = NA,
    IsBackdoor = NA,
    IsDirected = NA
  )
}
```


```{r}
#| label: create-ancestors-descendants-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 4: Ancestors and Descendants

# this chunk only creates a data frame but doesn't display it

ancestors_descendants_df <- data.frame(
  Variable = c("X", "Y", "E", "Q"),
  Ancestors = c(
    paste(dag_results$X_ancestors, collapse=", "),
    paste(dag_results$Y_ancestors, collapse=", "),
    paste(dag_results$E_ancestors, collapse=", "),
    paste(dag_results$Q_ancestors, collapse=", ")
  ),
  Descendants = c(
    paste(dag_results$X_descendants, collapse=", "),
    paste(dag_results$Y_descendants, collapse=", "),
    paste(dag_results$E_descendants, collapse=", "),
    paste(dag_results$Q_descendants, collapse=", ")
  )
)

```

```{r}
#| label: create-d-sep-df
#| echo: false
#| include: true
#| results: 'hide'

# this chunk only creates a data frame but doesn't display it

# Table 5: D-separation Results
d_sep_df <- data.frame(
  Variables = c("X and Y", "X and Y", "X and Y", "X and Y"),
  Conditioning_On = c("{ }", "E", "Q", "E and Q"),
  Is_D_Separated = c(
    ifelse(dag_results$d_sep_results$XY_given_nothing, "Yes", "No"),
    ifelse(dag_results$d_sep_results$XY_given_E, "Yes", "No"),
    ifelse(dag_results$d_sep_results$XY_given_Q, "Yes", "No"),
    ifelse(dag_results$d_sep_results$XY_given_EQ, "Yes", "No")
  )
)

```

```{r}
#| label: create-adjustment-effect-df
#| echo: false
#| include: true
#| results: 'hide'

# this chunk only creates a data frame but doesn't display it

# Table 6: Impact of Adjustments
adjustment_effect_df <- data.frame(
  Adjustment_Set = names(dag_results$adjustment_effects),
  Total_Paths = sapply(dag_results$adjustment_effects, function(x) x$total_paths),
  Open_Paths = sapply(dag_results$adjustment_effects, function(x) x$open_paths)
)

```

```{r}
#| label: create-instruments-df
#| echo: false
#| include: true
#| results: 'hide'

# this chunk only creates a data frame but doesn't display it

# Instrumental variables table
if(!is.null(dag_results$instruments) && length(dag_results$instruments) > 0) {
  # Convert the instruments to a character vector before creating data frame
  if(class(dag_results$instruments) == "dagitty.ivs") {
    instruments_list <- as.character(dag_results$instruments)
    instruments_df <- data.frame(
      Instruments = instruments_list
    )
  } else {
    instruments_df <- data.frame(
      Instruments = as.character("No valid instrumental variables found")
    )
  }
} else {
  instruments_df <- data.frame(
    Instruments = as.character("No valid instrumental variables found")
  )
}

```

```{r}
#| label: create-dag-plot
#| echo: false
#| include: true
#| results: 'hide'

# this chunk only creates a plot object but doesn't display it

# Create a nice visualization of the DAG using ggdag
dag_plot <- ggdag(peer_bias_dag8) + 
  theme_dag() +
  label("DAG: Peer Bias")

```
<br>

## 2. Results

### 2.1 Table of Key DAG Properties

```{r}
#| label: tbl-key-properties
#| tbl-cap: "Key Properties of the DAG"
#| code-fold: true

DT::datatable(
  properties_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.2 Table of Conditional Independencies

```{r}
#| label: independencies-analysis
#| tbl-cap: "Implied Conditional Independencies"

DT::datatable(
  independencies_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.3 Table of Paths Between X and Y

```{r}
#| label: tbl-paths
#| tbl-cap: "All Paths Between X and Y"

DT::datatable(
  paths_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.4 Table of Ancestors and Descendants

```{r}
#| label: tbl-ancestors-descendants
#| tbl-cap: "Ancestors and Descendants"

DT::datatable(
  ancestors_descendants_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.5 Table of D-Separation Results

```{r}
#| label: tbl-d-separation
#| tbl-cap: "D-Separation Test Results"

DT::datatable(
  d_sep_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.6 Table of Impact of Adjustments

```{r}
#| label: tbl-adjustments
#| tbl-cap: "Effect of Different Adjustment Sets"

DT::datatable(
  adjustment_effect_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.7 Table of Instrumental Variables

```{r}
#| label: tbl-instruments
#| tbl-cap: "Potential Instrumental Variables"

DT::datatable(
  instruments_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>


### 3. Visualizing Status, Adjustment Sets and Paths with ggdag

```{r}
#| fig-cap: "Different visualizations of the DAG"
#| fig-subcap: 
#|   - "Status Plot: Exposure and Outcome"
#|   - "Adjustment Sets for X → Y"
#|   - "All Paths between X and Y"
#| layout-ncol: 1
#| fig-width: 12
#| fig-height: 8

# Create dagitty object with ggdag positioning
dag <- dagitty("dag {
  Y <- X
  E <- X
  Y <- E
  E <- Q
  Y <- Q
  
  X [exposure]
  Y [outcome]
}")

# Set coordinates for visualization in digatty format
dagitty::coordinates(dag) <- list(x = c(X = 1.0, Y = 4.0, E = 2.5, Q = 4.0),  
                         y = c(X = 1.0, Y = 1.0, E = 3.0, Q = 3.0)
)

# Convert to ggdag format
dag_tidy <- ggdag::tidy_dagitty(dag)

# Status plot showing exposure/outcome
ggdag_status(dag_tidy) +
  ggdag::theme_dag(base_size = 16) +
  ggplot2::labs(title = "Status Plot: Exposure and Outcome")

# Adjustment set visualization
ggdag::ggdag_adjustment_set(dag_tidy) +
  ggdag::theme_dag(base_size = 16) +
  ggplot2::labs(title = "Adjustment Sets for X → Y")

# Paths visualization
ggdag::ggdag_paths(dag_tidy) +
  ggdag::theme_dag(base_size = 16) +
  ggplot2::labs(title = "All Paths between X and Y")

```
<br>

## 4. Interpretation and Discussion

### 4.1 Key Insights about this Peer Bias DAG Structure

This DAG represents a causal network with a peer bias structure, examining the relationship between X and Y with E as a mediator and Q as an unmeasured confounder:

1. **Direct Causal Effect (X → Y)**
   - X directly affects Y
   - This represents one component of the causal effect we're interested in measuring

2. **Mediation Path (X → E → Y)**
   - X affects E, which in turn affects Y
   - E is a mediator on the causal pathway from X to Y
   - This represents the indirect effect of X on Y through E

3. **Unmeasured Confounder (Q)**
   - Q affects both E and Y
   - Q creates a backdoor path between E and Y
   - This introduces confounding in the mediator-outcome relationship

4. **Peer Bias**
   - Adjusting for E while failing to adjust for Q can create bias
   - This happens because:
     - Conditioning on E blocks the indirect effect of X on Y through E
     - Conditioning on E opens a non-causal path between X and Y through Q (X → E ← Q → Y)
   - The resulting estimate may not reflect the total causal effect of X on Y

### 4.2 Proper Identification Strategy

To identify the causal effect of X on Y:
  - For the total effect of X on Y, do not adjust for E (the mediator)
  - If adjusting for E is necessary (e.g., to estimate the direct effect), also adjust for Q to block the opened collider path
  - If Q is unmeasured (as is often the case in real-world scenarios), consider:
    - Using sensitivity analysis to assess the potential impact of unmeasured confounding
    - Looking for proxy variables for Q
    - Using mediation analysis methods that can account for unmeasured confounding
  - The key insight is that adjusting for E without adjusting for Q leads to a biased estimate of the total causal effect

<br>


### Glossary

::: {.callout-note collapse="true"}
# DAG Analysis Glossary - Click to Open and Close

### Key DAG Terms and Concepts

**DAG (Directed Acyclic Graph)**: A graphical representation of causal relationships where arrows indicate the direction of causality, and no variable can cause itself through any path (hence "acyclic").

**Exposure**: The variable whose causal effect we want to estimate (often called the treatment or independent variable).

**Outcome**: The variable we are interested in measuring the effect on (often called the dependent variable).

**Confounder**: A variable that influences both the exposure and the outcome, potentially creating a spurious association between them.

**Mediator**: A variable that lies on the causal pathway between the exposure and outcome (exposure → mediator → outcome).

**Collider**: A variable that is influenced by both the exposure and the outcome, or by two variables on a path (e.g., A → C ← B).

**Backdoor path**: Any non-causal path connecting the exposure to the outcome that creates a spurious association.

**Instrumental Variable**: A variable that affects the exposure but has no direct effect on the outcome except through the exposure.

**Peer Bias**: A type of bias that occurs when we adjust for a variable E that is influenced by both the exposure X and an unmeasured confounder Q, which also affects the outcome Y. This can block mediation paths while opening collider paths.

### Understanding the Analysis Tables

#### 2. Conditional Independencies Table

Shows the implied conditional independencies in the DAG - pairs of variables that should be statistically independent when conditioning on specific other variables. These can be used to test the validity of your DAG against observed data.

#### 3. Paths Analysis Table

Enumerates all paths connecting the exposure to the outcome:

- **Path**: The specific variables and connections in each path
- **Length**: Number of edges in the path
- **IsBackdoor**: Whether this is a backdoor path (potential source of confounding)
- **IsDirected**: Whether this is a directed path from exposure to outcome

Testing whether these paths are open or closed under different conditioning strategies is crucial for causal inference.

#### 4. Ancestors and Descendants Table

Shows which variables can causally affect (ancestors) or be affected by (descendants) each variable in the DAG:

- Understanding ancestry relationships helps identify potential confounders
- X is an ancestor of E and Y, while Q is an ancestor of E and Y in this DAG

#### 5. D-Separation Results Table

Shows whether exposure and outcome are conditionally independent (d-separated) when conditioning on different variable sets:

- **Is_D_Separated = Yes**: This set of conditioning variables blocks all non-causal paths
- **Is_D_Separated = No**: Some non-causal association remains

This helps identify sufficient adjustment sets for estimating causal effects.

#### 6. Impact of Adjustments Table

Shows how different adjustment strategies affect the identification of causal effects:

- **Total_Paths**: Total number of paths between exposure and outcome
- **Open_Paths**: Number of paths that remain open after adjustment
  
Ideally, adjusting for the right variables leaves only the causal paths open.

#### 7. Instrumental Variables Table

Lists potential instrumental variables - variables that affect the exposure but have no direct effect on the outcome except through the exposure.

### How to Use This Analysis for Causal Inference

1. **Identify mediation effects**: In this DAG, E is a mediator between X and Y. If you're interested in the total effect of X on Y, don't control for E.

2. **Be cautious with mediator adjustment**: When adjusting for mediators like E, be aware that this can induce collider bias when unmeasured confounders like Q exist.

3. **Validate your DAG**: Use the implied conditional independencies to test your causal assumptions against observed data.

4. **Consider unmeasured confounders**: Always be aware of potential unmeasured confounders like Q and how they might affect your analysis, especially when adjusting for mediators.

5. **Choose appropriate analysis techniques**: When dealing with mediation, consider using formal mediation analysis techniques rather than simple regression adjustment.

Remember that the validity of any causal inference depends on the correctness of your DAG - it represents your causal assumptions about the data-generating process, which should be based on substantive domain knowledge.

#### 1. Key Properties Table

This table provides a high-level overview of the DAG structure and key causal features:

- **Acyclic DAG**: Confirms the graph has no cycles (a prerequisite for valid causal analysis)
- **Causal effect identifiable**: Indicates whether the causal effect can be estimated from observational data
- **Number of paths**: Total number of paths connecting exposure and outcome
- **Number of backdoor paths**: Paths creating potential confounding that need to be blocked
- **Direct effect exists**: Whether there is a direct causal link from exposure to outcome
- **Potential mediators**: Variables that may mediate the causal effect
- **Number of adjustment sets**: How many different sets of variables could be adjusted for
- **Minimal adjustment sets**: The smallest sets of variables that block all backdoor paths

:::
