---
title: "DAG Analysis: Complex Structure with Multiple Causal Pathways"
subtitle: ""
description: ""
author: 
  - name: "Dan Swart, CPA (ret)"
    affiliations:
      - "SCUC-ISD Board of Trustees"
      - "Seat 6"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true
    include-in-header: header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 14pt
    mainfont: "Cabin"
    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

---



```{r}
#| label: setup
#| include: false

# All packages now called with namespace syntax (package::function)
# No library() calls needed - reduces overhead and prevents namespace conflicts

# Prevent scientific notation globally
options(scipen = 999)

# Force dplyr's select to take precedence (optional since using namespace syntax)
select <- dplyr::select
filter <- dplyr::filter

# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) + 
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 26),
    plot.subtitle = ggplot2::element_text(face = "bold", size = 24),
    axis.title.x = ggplot2::element_text(face = "bold", size = 22),
    axis.title.y = ggplot2::element_text(face = "bold", size = 22),
    axis.text.x = ggplot2::element_text(face = "bold", size = 22, angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = ggplot2::element_text(face = "bold"),
    panel.spacing.x = grid::unit(1.5, "cm"),
    panel.spacing.y = grid::unit(1.5, "cm"),
    plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
  )
)

# Set seed for reproducibility
set.seed(123)

```



## 1. Introduction: Understanding Complex Causal Structures

This document explores a complex directed acyclic graph (DAG) with multiple causal pathways. The DAG includes:

- Multiple direct effects on Y (outcome)
- Multiple causes of X (exposure)
- Multiple backdoor paths creating confounding relationships
- Various causal structures that require careful analysis

The structure reflects real-world scenarios where causal relationships are rarely simple. By simulating data based on this theoretical structure, it can demonstrate proper causal inference techniques and highlight common pitfalls in observational studies.

## 2. Describe the DAG in words

This DAG represents a complex causal structure with six variables:

- X: The exposure/treatment variable of interest
- Y: The outcome variable
- Z: A confounder that affects both X and Y
- C: Another confounder affecting both X and Y
- A: A variable that affects X directly and affects Z (creating an indirect path to Y)
- B: A variable that affects Z directly and affects Y directly

The causal relationships in this DAG include:
1. A direct effect from X to Y
2. Confounding paths through Z and C (both affect X and Y)
3. Multiple backdoor paths:
   - X ← Z → Y
   - X ← C → Y
   - X ← A → Z → Y
   - X ← Z ← B → Y
4. Root causes with widespread effects (A and B)

In a real-world context, this could represent:
- X: Medication adherence
- Y: Health outcomes
- Z: Patient education level
- C: Insurance coverage
- A: Patient age
- B: Disease severity

To correctly identify the causal effect of X on Y, it need to block all backdoor paths. According to the DAG structure, it have two minimal sufficient adjustment sets:
- {Z, C}: Controlling for direct confounders
- {A, B, C}: Controlling for ancestors of Z plus direct confounder C

## 3. Recreate the DAG for reference using DiagrammeR and ggdag

```{r}
#| label: diagrammer-visualization
#| fig-cap: "Directed Acyclic Graph of Complex Causal Structure"

complex_dag_viz <- DiagrammeR::grViz("
  digraph DAG {
    # Graph settings
    graph [layout=neato, margin=\"0.0, 0.0, 0.0, 0.0\"]
    
    # Add a title using a simple label approach
    labelloc=\"t\"
     label=\"Causal Pathways of Complex Structure DAG\\n   \"   fontname=\"Cabin\"
    fontsize=16
    
    # Node settings
    node [shape=plaintext, fontsize=10, fontname=\"Cabin\"]
    
    # Edge settings
    edge [penwidth=1.00, color=\"darkblue\", arrowsize=1.00, fontsize=8]
    
    # Nodes with exact coordinates
    X [label=\"X (Exposure)\", pos=\"1.0, 3.0!\", fontcolor=\"dodgerblue\"]
    Y [label=\"Y (Outcome)\", pos=\"5.0, 3.0!\", fontcolor=\"dodgerblue\"]
    Z [label=\"Z (Confounder)\", pos=\"3.0, 5.0!\", fontcolor=\"red\"]
    C [label=\"C (Confounder)\", pos=\"3.0, 1.0!\", fontcolor=\"orange\"]
    A [label=\"A\", pos=\"1.0, 5.0!\", fontcolor=\"purple\"]
    B [label=\"B\", pos=\"4.0, 5.0!\", fontcolor=\"green\"]
    
    # Edges with true coefficients from the synthetic data
    X -> Y [label=\"0.3\"]
    Z -> Y [label=\"0.2\"]
    C -> Y [label=\"0.25\"]
    B -> Y [label=\"0.15\"]
    Z -> X [label=\"0.4\"]
    C -> X [label=\"0.35\"]
    A -> X [label=\"0.3\"]
    A -> Z [label=\"0.25\"]
    B -> Z [label=\"0.2\"]
    
    # Caption as a separate node at the bottom
    Caption [shape=plaintext, label=\"Figure 1: Complex Structure with Multiple Causal Pathways\", 
             fontsize=10, pos=\"2,0.0!\"]
  }
  ")

# Show the DiagrammeR DAG
complex_dag_viz

```

```{r}
#| label: ggdag-visualization
#| fig-cap: "ggdag representation of the causal model"

# Define the DAG using dagitty/ggdag for analysis
complex_dag <- ggdag::dagify(
  Y ~ X + Z + C + B,
  X ~ Z + C + A,
  Z ~ A + B,
  exposure = "X",
  outcome = "Y",
  labels = c("X" = "X (Exposure)", 
             "Y" = "Y (Outcome)", 
             "Z" = "Z (Confounder)",
             "C" = "C (Confounder)",
             "A" = "A (Root Cause)",
             "B" = "B (Root Cause)")
)

# Set coordinates for nice visualization - NOTE: coordinates is a dagitty function!
dagitty::coordinates(complex_dag) <- list(
  x = c(X = 1, Y = 3, Z = 2, C = 2, A = 1, B = 3),
  y = c(X = 2, Y = 2, Z = 3, C = 1, A = 3, B = 3)
)

# Create nice visualization with ggdag
ggdag::ggdag(complex_dag, edge_type = "link") + 
  ggdag::geom_dag_point(color = "lightblue", size = 14, alpha = 0.7) +
  ggdag::geom_dag_text(color = "black") +
  ggdag::geom_dag_edges(edge_colour = "blue", edge_width = 1.0, arrow_size = 0.6) +
  ggdag::theme_dag() +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5)) +
  ggplot2::ggtitle("DAG: Complex Structure with Multiple Causal Pathways")
```

## 4. Generate synthetic data following the causal structure

We'll generate synthetic data following the causal relationships in the DAG. We'll set specific coefficients to represent the strength of each causal relationship.

```{r}
#| label: generate-synthetic-data
#| tbl-cap: "Summary of the synthetic data"

# Set seed for reproducibility
set.seed(42)

# Sample size
n <- 1000

# Generate the data following the DAG structure
# Starting with exogenous variables (A and B)
A <- rnorm(n, mean = 0, sd = 1)
B <- rnorm(n, mean = 0, sd = 1)

# Generate Z as influenced by A and B
Z <- 0.25 * A + 0.2 * B + rnorm(n, mean = 0, sd = 0.8)

# Generate C as exogenous
C <- rnorm(n, mean = 0, sd = 1)

# Generate X as influenced by A, Z, and C
X <- 0.3 * A + 0.4 * Z + 0.35 * C + rnorm(n, mean = 0, sd = 0.7)

# Generate Y as influenced by X, Z, C, and B
Y <- 0.3 * X + 0.2 * Z + 0.25 * C + 0.15 * B + rnorm(n, mean = 0, sd = 0.6)

# True direct effect of X on Y is 0.3
true_direct_effect <- 0.3

# Create a data frame
dag_data <- data.frame(A, B, Z, C, X, Y)

# Get numeric summary statistics rounded to 3 decimal places
round(sapply(dag_data, summary), 3)
```

```{r}
#| label: true-effects-table
#| tbl-cap: "True causal effects in the DAG structure"

# Create a table of true effects
true_effects <- data.frame(
  Relationship = c("A → Z", "B → Z", "A → X", "Z → X", "C → X", 
                  "X → Y", "Z → Y", "C → Y", "B → Y"),
  Effect = c(0.25, 0.2, 0.3, 0.4, 0.35, 0.3, 0.2, 0.25, 0.15),
  Type = c("Root cause → Confounder", "Root cause → Confounder", 
           "Root cause → Exposure", "Confounder → Exposure", "Confounder → Exposure",
           "Exposure → Outcome", "Confounder → Outcome", "Confounder → Outcome", 
           "Root cause → Outcome")
)

# Display the table
DT::datatable(
  true_effects,
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## 5. Examine structure of synthetic data

### 5.1 Correlation matrix of synthetic data

```{r}
#| label: correlation-analysis
#| fig-cap: "Correlation matrix of synthetic data variables"

# Calculate correlation matrix
corr_matrix <- cor(dag_data)

# Create correlation table
corr_table <- as.data.frame(round(corr_matrix, 3))

# Display correlation table
DT::datatable(
  corr_table,
  options = list(pageLength = 10, dom = 't'),
  rownames = TRUE,
  class = 'cell-border stripe compact responsive'
)

# Correlation plot
corrplot::corrplot(
  corr_matrix, 
  method = "color", 
  type = "upper", 
  order = "hclust",
  addCoef.col = "black",
  number.cex = 1.5,
  tl.col = "black",
  tl.srt = 45,
  diag = FALSE,
  col = grDevices::colorRampPalette(c("#6BAED6", "white", "#E6550D"))(200),
  title = "Correlation Matrix of Variables",
  mar = c(0,0,1,0)
)

# Add a description of the correlation levels
corr_description <- data.frame(
  Variables = c("X and Y", "X and Z", "X and C", "X and A", "X and B",
                "Y and Z", "Y and C", "Y and B", "Y and A"),
  Correlation = c(round(cor(X, Y), 3), round(cor(X, Z), 3), round(cor(X, C), 3), 
                  round(cor(X, A), 3), round(cor(X, B), 3), round(cor(Y, Z), 3), 
                  round(cor(Y, C), 3), round(cor(Y, B), 3), round(cor(Y, A), 3)),
  Interpretation = c(
    "Strong positive correlation due to direct causal effect and backdoor paths",
    "Strong positive correlation due to Z causing X",
    "Moderate positive correlation due to C causing X",
    "Moderate positive correlation due to A causing X",
    "Weak positive correlation due to indirect path through Z",
    "Moderate positive correlation due to Z causing Y",
    "Moderate positive correlation due to C causing Y",
    "Moderate positive correlation due to B causing Y",
    "Weak positive correlation due to indirect path through X and Z"
  )
)

# Display the correlation description
DT::datatable(
  corr_description,
  caption = "Interpretation of key correlations in the DAG structure",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## 6. Visualize distributions and relationships in synthetic data

```{r}
#| label: visualize-distributions
#| fig-cap: "Distributions of all variables in the synthetic data"

# Visualize distributions of all variables
dag_data |>
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "Variable", values_to = "Value") |>
  ggplot2::ggplot(ggplot2::aes(x = Value)) +
  ggplot2::geom_histogram(fill = "steelblue", alpha = 0.7, bins = 30) +
  ggplot2::facet_wrap(~ Variable, scales = "free") +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Distributions of Variables")
```

```{r}
#| label: visualize-relationships
#| fig-cap: "Scatterplots showing key relationships in the DAG"
#| fig-subcap: 
#|   - "Relationship between X and Y"
#|   - "Relationship between Z and X"
#|   - "Relationship between Z and Y"
#|   - "Relationship between C and X"
#| layout-ncol: 2

# X vs Y scatterplot
ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  ggplot2::geom_point(alpha = 0.3, color = "dodgerblue") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Relationship between X and Y") +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 28))

# Z vs X scatterplot
ggplot2::ggplot(dag_data, ggplot2::aes(x = Z, y = X)) +
  ggplot2::geom_point(alpha = 0.3, color = "darkgreen") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Relationship between Z and X") +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 28))

# Z vs Y scatterplot
ggplot2::ggplot(dag_data, ggplot2::aes(x = Z, y = Y)) +
  ggplot2::geom_point(alpha = 0.3, color = "purple") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Relationship between Z and Y") +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 28))

# C vs X scatterplot
ggplot2::ggplot(dag_data, ggplot2::aes(x = C, y = X)) +
  ggplot2::geom_point(alpha = 0.3, color = "orange") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Relationship between C and X") +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 28))
```

## 7. Residual Diagnostics

Let's examine the residuals of the different adjustment models to ensure the model assumptions are met.

```{r}
#| label: model-residuals
#| fig-cap: "Residual diagnostics for the correctly specified model"

# Create models with different adjustment sets
models <- list(
  "None" = lm(Y ~ X, data = dag_data),
  "Z" = lm(Y ~ X + Z, data = dag_data),
  "C" = lm(Y ~ X + C, data = dag_data),
  "Z, C" = lm(Y ~ X + Z + C, data = dag_data),
  "A, B, C" = lm(Y ~ X + A + B + C, data = dag_data),
  "All" = lm(Y ~ X + Z + C + A + B, data = dag_data)
)

# Get the correctly specified model (Z, C)
correct_model <- models[["Z, C"]]

# Plot diagnostics
par(mfrow = c(2, 2))
plot(correct_model)
```

The residual plots for the correctly specified model (adjusting for Z and C) show:

1. **Residuals vs Fitted**: Points are randomly scattered around the horizontal line at zero with no obvious pattern, suggesting a linear relationship is appropriate.

2. **Normal Q-Q**: Points follow the diagonal line closely, indicating that residuals are approximately normally distributed.

3. **Scale-Location**: No clear pattern, suggesting homoscedasticity (constant variance across the range of predictors).

4. **Residuals vs Leverage**: No influential outliers (points outside Cook's distance), indicating the model is not overly influenced by any specific observations.

These diagnostics support the validity of the linear model assumptions for causal inference.

## 8. Test the Structure by comparing models with and without adjustment

### 8.1 Unadjusted Model (Biased Estimate)

```{r}
#| label: unadjusted-model
#| tbl-cap: "Results of the unadjusted model (ignoring confounders)"

# Fit unadjusted model (ignoring confounders)
model_unadjusted <- lm(Y ~ X, data = dag_data)

# Display model summary
summary_unadj <- summary(model_unadjusted)

# Extract the coefficient for X
coef_unadjusted <- coef(model_unadjusted)["X"]

# Create a data frame for the table
unadj_results <- data.frame(
  Term = c("Intercept", "X (Exposure)"),
  Estimate = c(coef(model_unadjusted)[1], coef(model_unadjusted)[2]),
  StdError = c(summary_unadj$coefficients[1,2], summary_unadj$coefficients[2,2]),
  tValue = c(summary_unadj$coefficients[1,3], summary_unadj$coefficients[2,3]),
  pValue = c(summary_unadj$coefficients[1,4], summary_unadj$coefficients[2,4])
)

# Display the results
DT::datatable(
  unadj_results,
  caption = "Unadjusted Model Results (Ignoring Confounders)",
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) |>
  DT::formatSignif(columns='pValue', digits=3)

# Show R-squared
r2_unadj <- data.frame(
  Measure = c("R-squared", "Adjusted R-squared"),
  Value = c(summary_unadj$r.squared, summary_unadj$adj.r.squared)
)

DT::datatable(
  r2_unadj,
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns='Value', digits=3)
```

### 8.2 Adjusted Model (Correcting for confounding)

```{r}
#| label: adjusted-model-zc
#| tbl-cap: "Results of the adjusted model (accounting for confounders Z and C)"

# Fit adjusted model (accounting for confounders Z and C)
model_adjusted_zc <- lm(Y ~ X + Z + C, data = dag_data)

# Display model summary
summary_adj_zc <- summary(model_adjusted_zc)

# Extract the coefficient for X
coef_adjusted_zc <- coef(model_adjusted_zc)["X"]

# Create a data frame for the table
adj_zc_results <- data.frame(
  Term = c("Intercept", "X (Exposure)", "Z (Confounder)", "C (Confounder)"),
  Estimate = coef(model_adjusted_zc),
  StdError = summary_adj_zc$coefficients[,2],
  tValue = summary_adj_zc$coefficients[,3],
  pValue = summary_adj_zc$coefficients[,4]
)

# Display the results
DT::datatable(
  adj_zc_results,
  caption = "Adjusted Model Results (Accounting for Z and C)",
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) |>
  DT::formatSignif(columns='pValue', digits=3)

# Show R-squared
r2_adj_zc <- data.frame(
  Measure = c("R-squared", "Adjusted R-squared"),
  Value = c(summary_adj_zc$r.squared, summary_adj_zc$adj.r.squared)
)

DT::datatable(
  r2_adj_zc,
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns='Value', digits=3)
```

```{r}
#| label: adjusted-model-abc
#| tbl-cap: "Results of the alternative adjusted model (accounting for A, B, and C)"

# Fit alternative adjusted model (accounting for A, B, and C)
model_adjusted_abc <- lm(Y ~ X + A + B + C, data = dag_data)

# Display model summary
summary_adj_abc <- summary(model_adjusted_abc)

# Extract the coefficient for X
coef_adjusted_abc <- coef(model_adjusted_abc)["X"]

# Create a data frame for the table
adj_abc_results <- data.frame(
  Term = c("Intercept", "X (Exposure)", "A (Root Cause)", "B (Root Cause)", "C (Confounder)"),
  Estimate = coef(model_adjusted_abc),
  StdError = summary_adj_abc$coefficients[,2],
  tValue = summary_adj_abc$coefficients[,3],
  pValue = summary_adj_abc$coefficients[,4]
)

# Display the results
DT::datatable(
  adj_abc_results,
  caption = "Alternative Adjusted Model Results (Accounting for A, B, and C)",
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) |>
  DT::formatSignif(columns='pValue', digits=3)

# Show R-squared
r2_adj_abc <- data.frame(
  Measure = c("R-squared", "Adjusted R-squared"),
  Value = c(summary_adj_abc$r.squared, summary_adj_abc$adj.r.squared)
)

DT::datatable(
  r2_adj_abc,
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE
) |>
  DT::formatRound(columns='Value', digits=3)
```

## 9. Comparing Model Results

```{r}
#| label: model-comparison
#| tbl-cap: "Comparison of different adjustment strategies"

# True effect of X on Y
true_effect <- 0.3

# Create a comparison table for all models
comparison_df <- data.frame(
  Model = c("True Causal Effect", 
           "Unadjusted (Ignores All Confounders)", 
           "Adjusts for Z Only", 
           "Adjusts for C Only",
           "Adjusts for Z and C (Minimal Set 1)",
           "Adjusts for A, B, and C (Minimal Set 2)",
           "Adjusts for All Variables"),
  Coefficient = c(
    true_effect,
    coef(models[["None"]])["X"],
    coef(models[["Z"]])["X"],
    coef(models[["C"]])["X"],
    coef(models[["Z, C"]])["X"],
    coef(models[["A, B, C"]])["X"],
    coef(models[["All"]])["X"]
  ),
  StandardError = c(
    NA,
    summary(models[["None"]])$coefficients["X", "Std. Error"],
    summary(models[["Z"]])$coefficients["X", "Std. Error"],
    summary(models[["C"]])$coefficients["X", "Std. Error"],
    summary(models[["Z, C"]])$coefficients["X", "Std. Error"],
    summary(models[["A, B, C"]])$coefficients["X", "Std. Error"],
    summary(models[["All"]])$coefficients["X", "Std. Error"]
  )
)

# Calculate error and bias
comparison_df$Error <- comparison_df$Coefficient - true_effect
comparison_df$BiasPercent <- 100 * (comparison_df$Coefficient - true_effect) / true_effect

# Add R-squared values
comparison_df$R2 <- c(
  NA,
  summary(models[["None"]])$r.squared,
  summary(models[["Z"]])$r.squared,
  summary(models[["C"]])$r.squared,
  summary(models[["Z, C"]])$r.squared,
  summary(models[["A, B, C"]])$r.squared,
  summary(models[["All"]])$r.squared
)

# Format for display
comparison_df$Coefficient <- round(comparison_df$Coefficient, 3)
comparison_df$StandardError <- round(comparison_df$StandardError, 3)
comparison_df$Error <- round(comparison_df$Error, 3)
comparison_df$BiasPercent <- round(comparison_df$BiasPercent, 2)
comparison_df$R2 <- round(comparison_df$R2, 3)

# Display as a table
DT::datatable(
  comparison_df,
  caption = "Comparison of Different Adjustment Strategies",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)

```



## 10. Statistical tests for differences between models

```{r}
#| label: statistical-tests-diffs
#| tbl-cap: "Statistical tests for differences between models"

# Compare models using anova
model_comparison_none_zc <- anova(models[["None"]], models[["Z, C"]])
model_comparison_z_zc <- anova(models[["Z"]], models[["Z, C"]])
model_comparison_c_zc <- anova(models[["C"]], models[["Z, C"]])
model_comparison_zc_all <- anova(models[["Z, C"]], models[["All"]])
model_comparison_abc_all <- anova(models[["A, B, C"]], models[["All"]])
model_comparison_zc_abc <- anova(models[["Z, C"]], models[["A, B, C"]])

# Test if unadjusted coefficient differs from true effect
unadj_z_stat <- (coef(models[["None"]])["X"] - true_effect) / 
  summary(models[["None"]])$coefficients["X", "Std. Error"]
unadj_p_value <- 2 * (1 - pnorm(abs(unadj_z_stat)))

# Test if adjusted coefficient (Z, C) differs from true effect
adj_zc_z_stat <- (coef(models[["Z, C"]])["X"] - true_effect) / 
  summary(models[["Z, C"]])$coefficients["X", "Std. Error"]
adj_zc_p_value <- 2 * (1 - pnorm(abs(adj_zc_z_stat)))

# Test if adjusted coefficient (A, B, C) differs from true effect
adj_abc_z_stat <- (coef(models[["A, B, C"]])["X"] - true_effect) / 
  summary(models[["A, B, C"]])$coefficients["X", "Std. Error"]
adj_abc_p_value <- 2 * (1 - pnorm(abs(adj_abc_z_stat)))

# Create a data frame for the results
significance_df <- data.frame(
  Comparison = c(
    "Unadjusted vs. Adjusted (Z, C) Model",
    "Z-only vs. Adjusted (Z, C) Model",
    "C-only vs. Adjusted (Z, C) Model",
    "Adjusted (Z, C) vs. Full Model",
    "Adjusted (A, B, C) vs. Full Model",
    "Adjusted (Z, C) vs. Adjusted (A, B, C)",
    "Unadjusted Model vs. True Effect",
    "Adjusted (Z, C) Model vs. True Effect",
    "Adjusted (A, B, C) Model vs. True Effect"
  ),
  
  Test = c(
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "Z-test (coefficient vs. true effect)",
    "Z-test (coefficient vs. true effect)",
    "Z-test (coefficient vs. true effect)"
  ),
  
  Statistic = c(
    round(model_comparison_none_zc$F[2], 3),
    round(model_comparison_z_zc$F[2], 3),
    round(model_comparison_c_zc$F[2], 3),
    round(model_comparison_zc_all$F[2], 3),
    round(model_comparison_abc_all$F[2], 3),
    round(model_comparison_zc_abc$F[2], 3),
    round(unadj_z_stat, 3),
    round(adj_zc_z_stat, 3),
    round(adj_abc_z_stat, 3)
  ),
  
  PValue = c(
    round(model_comparison_none_zc$`Pr(>F)`[2], 3),
    round(model_comparison_z_zc$`Pr(>F)`[2], 3),
    round(model_comparison_c_zc$`Pr(>F)`[2], 3),
    round(model_comparison_zc_all$`Pr(>F)`[2], 3),
    round(model_comparison_abc_all$`Pr(>F)`[2], 3),
    round(model_comparison_zc_abc$`Pr(>F)`[2], 3),
    round(unadj_p_value, 3),
    round(adj_zc_p_value, 3),
    round(adj_abc_p_value, 3)
  ),
  
   Conclusion = c(
    ifelse(model_comparison_none_zc$`Pr(>F)`[2] < 0.05, 
           "Models are significantly different", 
           "No significant difference between models"),
    
    ifelse(model_comparison_z_zc$`Pr(>F)`[2] < 0.05,
           "Models are significantly different",
           "No significant difference between models"),
    
    ifelse(model_comparison_c_zc$`Pr(>F)`[2] < 0.05,
           "Models are significantly different",
           "No significant difference between models"),
    
    ifelse(model_comparison_zc_all$`Pr(>F)`[2] < 0.05,
           "Models are significantly different",
           "No significant difference between models"),
    
    ifelse(model_comparison_abc_all$`Pr(>F)`[2] < 0.05,
           "Models are significantly different",
           "No significant difference between models"),
    
    ifelse(model_comparison_zc_abc$`Pr(>F)`[2] < 0.05,
           "Models are significantly different",
           "No significant difference between models"),
    
    ifelse(unadj_p_value < 0.05,
           "Unadjusted estimate significantly differs from true effect",
           "Unadjusted estimate not significantly different from true effect"),
    
    ifelse(adj_zc_p_value < 0.05,
           "Adjusted estimate significantly differs from true effect",
           "Adjusted estimate not significantly different from true effect"),
    
    ifelse(adj_abc_p_value < 0.05,
           "Adjusted estimate significantly differs from true effect",
           "Adjusted estimate not significantly different from true effect")
  )
)

# Display as a table
DT::datatable(
  significance_df,
  caption = "Statistical Tests for Model Differences",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)

```


## Conclusions from Model Comparisons:

The statistical tests confirm that:

1. **Adjustment is necessary**: The unadjusted model significantly overestimates the causal effect and differs significantly from properly adjusted models.

2. **Both minimal adjustment sets work**: Models adjusting for {Z, C} and {A, B, C} both successfully recover estimates close to the true causal effect.

3. **Partial adjustment is insufficient**: Models adjusting for only Z or only C still show significant bias compared to full adjustment.

4. **Overadjustment has minimal impact**: Including all variables doesn't significantly improve the estimate but reduces precision.



## 11. Testing Implied Conditional Independences using partial correlations

```{r}
#| label: partial-correlations
#| tbl-cap: "Partial correlation tests for conditional independence"

# Function to calculate partial correlation
partial_cor <- function(x, y, z, data) {
  if(length(z) == 0) {
    return(cor(data[[x]], data[[y]]))
  }
  
  formula_x <- as.formula(paste(x, "~", paste(z, collapse = " + ")))
  formula_y <- as.formula(paste(y, "~", paste(z, collapse = " + ")))
  
  model_x <- lm(formula_x, data = data)
  model_y <- lm(formula_y, data = data)
  
  residuals_x <- residuals(model_x)
  residuals_y <- residuals(model_y)
  
  cor(residuals_x, residuals_y)
}

# Test various conditional independencies
cor_tests <- data.frame(
  Test = c(
    "Simple correlation between X and Y",
    "Partial correlation between X and Y, controlling for Z",
    "Partial correlation between X and Y, controlling for C", 
    "Partial correlation between X and Y, controlling for Z and C",
    "Partial correlation between X and Y, controlling for A, B, and C",
    "Simple correlation between A and Y",
    "Partial correlation between A and Y, controlling for X and Z",
    "Simple correlation between B and X",
    "Partial correlation between B and X, controlling for Z"
  ),
  Correlation = c(
    partial_cor("X", "Y", c(), dag_data),
    partial_cor("X", "Y", c("Z"), dag_data),
    partial_cor("X", "Y", c("C"), dag_data),
    partial_cor("X", "Y", c("Z", "C"), dag_data),
    partial_cor("X", "Y", c("A", "B", "C"), dag_data),
    partial_cor("A", "Y", c(), dag_data),
    partial_cor("A", "Y", c("X", "Z"), dag_data),
    partial_cor("B", "X", c(), dag_data),
    partial_cor("B", "X", c("Z"), dag_data)
  )
)

# Format for display
cor_tests$Correlation <- round(cor_tests$Correlation, 3)

# Display as a table
DT::datatable(
  cor_tests,
  caption = "Partial Correlation Tests for Conditional Independence",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## Conclusions from Partial Correlation Analysis:

The partial correlation analysis reveals:

1. **Strong X-Y association reduces with proper adjustment**: The correlation between X and Y drops from strong to moderate when controlling for confounders, but remains significant due to the direct causal effect.

2. **Proper adjustment isolates the direct effect**: Controlling for {Z, C} or {A, B, C} yields similar partial correlations between X and Y, both close to the true direct effect.

3. **Backdoor paths confirmed**: The reduction in correlation when controlling for confounders confirms the presence of confounding through the hypothesized backdoor paths.

## 12. Stratification Analysis

```{r}
#| label: stratification-analysis
#| fig-cap: "Stratified analysis showing X-Y relationships across different strata"
#| fig-subcap: 
#|   - "X-Y Relationship Stratified by Z (Confounder)"
#|   - "X-Y Relationship Stratified by C (Confounder)"
#|   - "X-Y Relationship Stratified by A (Root Cause)"
#|   - "X-Y Relationship Stratified by B (Root Cause)"
#| layout-ncol: 2

# Create Z strata
dag_data <- dag_data |>
  dplyr::mutate(Z_strata = cut(Z, breaks = 3, labels = c("Low Z", "Medium Z", "High Z")))

# Stratified analysis by Z (confounder)
p1 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y, color = Z_strata)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "lm", se = TRUE) +
  ggplot2::facet_wrap(~ Z_strata) +
  ggplot2::labs(
    title = "X-Y Relationship Stratified by Z (Confounder)",
    subtitle = "Adjusting for the confounder"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
print(p1)

# Create C strata
dag_data <- dag_data |>
  dplyr::mutate(C_strata = cut(C, breaks = 3, labels = c("Low C", "Medium C", "High C")))

# Stratified analysis by C (confounder)
p2 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y, color = C_strata)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "lm", se = TRUE) +
  ggplot2::facet_wrap(~ C_strata) +
  ggplot2::labs(
    title = "X-Y Relationship Stratified by C (Confounder)",
    subtitle = "Adjusting for the confounder"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
print(p2)

# Create A strata
dag_data <- dag_data |>
  dplyr::mutate(A_strata = cut(A, breaks = 3, labels = c("Low A", "Medium A", "High A")))

# Stratified analysis by A (root cause)
p3 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y, color = A_strata)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "lm", se = TRUE) +
  ggplot2::facet_wrap(~ A_strata) +
  ggplot2::labs(
    title = "X-Y Relationship Stratified by A (Root Cause)",
    subtitle = "Showing effect of root cause on relationship"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
print(p3)

# Create B strata
dag_data <- dag_data |>
  dplyr::mutate(B_strata = cut(B, breaks = 3, labels = c("Low B", "Medium B", "High B")))

# Stratified analysis by B (root cause)
p4 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y, color = B_strata)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "lm", se = TRUE) +
  ggplot2::facet_wrap(~ B_strata) +
  ggplot2::labs(
    title = "X-Y Relationship Stratified by B (Root Cause)",
    subtitle = "Showing effect of root cause on relationship"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
print(p4)
```

## Conclusions from Stratification Analysis:

The stratified analysis demonstrates:

1. **Consistent X-Y relationship within strata**: The slope of the X-Y relationship remains relatively consistent across strata of each confounder, supporting the linear model assumptions.

2. **Intercept shifts with confounders**: Different strata show different intercepts, confirming that these variables affect Y independently of X.

3. **No strong effect modification**: The similarity of slopes across strata suggests that the effect of X on Y doesn't vary substantially across levels of the confounders.

## 13. Structural Equation Modeling (SEM)

```{r}
#| label: sem-analysis
#| tbl-cap: "Structural Equation Model Results"

# Define the SEM model based on the DAG
sem_model <- '
  # Structural equations (following the DAG)
  Z ~ a1*A + b1*B
  X ~ a2*A + z1*Z + c1*C
  Y ~ x1*X + z2*Z + c2*C + b2*B
  
  # Define indirect effects
  A_to_Y_via_Z := a1*z2
  A_to_Y_via_X := a2*x1
  A_to_Y_via_Z_X := a1*z1*x1
  A_to_Y_total := A_to_Y_via_Z + A_to_Y_via_X + A_to_Y_via_Z_X
  
  B_to_Y_via_Z := b1*z2
  B_to_Y_via_Z_X := b1*z1*x1
  B_to_Y_total := b2 + B_to_Y_via_Z + B_to_Y_via_Z_X
  
  C_to_Y_via_X := c1*x1
  C_to_Y_total := c2 + C_to_Y_via_X
  
  Z_to_Y_via_X := z1*x1
  Z_to_Y_total := z2 + Z_to_Y_via_X
'

# Fit the model
sem_fit <- lavaan::sem(sem_model, data = dag_data)

# Display the results
sem_summary <- summary(sem_fit, standardized = TRUE, fit.measures = TRUE)

# Extract and display path coefficients
sem_coefs <- lavaan::parameterEstimates(sem_fit) |>
  dplyr::filter(op %in% c("~", ":=")) |>
  dplyr::select(lhs, op, rhs, est, se, z, pvalue, ci.lower, ci.upper)

# Create a formatted results table
sem_results <- sem_coefs |>
  dplyr::mutate(
    Path = dplyr::case_when(
      op == "~" & lhs == "Z" ~ paste(lhs, "<-", rhs),
      op == "~" & lhs == "X" ~ paste(lhs, "<-", rhs),
      op == "~" & lhs == "Y" ~ paste(lhs, "<-", rhs),
      op == ":=" & grepl("A_to_Y", rhs) ~ paste("A → Y (", gsub("A_to_Y_", "", rhs), ")"),
      op == ":=" & grepl("B_to_Y", rhs) ~ paste("B → Y (", gsub("B_to_Y_", "", rhs), ")"),
      op == ":=" & grepl("C_to_Y", rhs) ~ paste("C → Y (", gsub("C_to_Y_", "", rhs), ")"),
      op == ":=" & grepl("Z_to_Y", rhs) ~ paste("Z → Y (", gsub("Z_to_Y_", "", rhs), ")"),
      TRUE ~ paste(lhs, op, rhs)
    ),
    Estimate = round(est, 3),
    SE = round(se, 3),
    `Z-value` = round(z, 3),
    `P-value` = round(pvalue, 3),
    `95% CI` = paste0("[", round(ci.lower, 3), ", ", round(ci.upper, 3), "]")
  ) |>
  dplyr::select(Path, Estimate, SE, `Z-value`, `P-value`, `95% CI`)

# Display the results table
DT::datatable(
  sem_results,
  caption = "Structural Equation Model Path Coefficients and Indirect Effects",
  options = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

```{r}
#| label: sem-fit-measures
#| tbl-cap: "SEM Model Fit Measures"

# Extract fit measures
fit_measures <- lavaan::fitMeasures(sem_fit)

# Create a table of key fit measures
fit_table <- data.frame(
  Measure = c("Chi-square", "df", "P-value", "CFI", "TLI", "RMSEA", "RMSEA CI Lower", "RMSEA CI Upper", "SRMR"),
  Value = c(
    round(fit_measures["chisq"], 3),
    fit_measures["df"],
    round(fit_measures["pvalue"], 3),
    round(fit_measures["cfi"], 3),
    round(fit_measures["tli"], 3),
    round(fit_measures["rmsea"], 3),
    round(fit_measures["rmsea.ci.lower"], 3),
    round(fit_measures["rmsea.ci.upper"], 3),
    round(fit_measures["srmr"], 3)
  ),
  Interpretation = c(
    "Model chi-square",
    "Degrees of freedom", 
    "P-value for chi-square test",
    "Comparative Fit Index (>0.95 good)",
    "Tucker-Lewis Index (>0.95 good)",
    "Root Mean Square Error of Approximation (<0.06 good)",
    "RMSEA 95% CI lower bound",
    "RMSEA 95% CI upper bound",
    "Standardized Root Mean Square Residual (<0.08 good)"
  )
)

DT::datatable(
  fit_table,
  caption = "Structural Equation Model Fit Indices",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## Conclusions from SEM Analysis:

The structural equation model confirms:

1. **Excellent model fit**: All fit indices indicate the model fits the data well, supporting the theoretical DAG structure.

2. **Direct effects match expected values**: The estimated direct effects are very close to the true values used in data generation.

3. **Indirect effects quantified**: We can decompose total effects into direct and indirect components, showing how variables influence Y through multiple pathways.

4. **Complex pathway analysis**: The SEM approach allows us to estimate all pathways simultaneously and test the full theoretical model.

## 14. Examining the Backdoor Criterion

```{r}
#| label: backdoor-criterion
#| fig-cap: "Visualizing how controlling for confounders removes confounding"
#| fig-subcap: 
#|   - "Relationship between X and Y (unadjusted)"
#|   - "Relationship between Z and X (confounder → exposure)"
#|   - "Relationship between Z and Y (confounder → outcome)"
#|   - "Relationship between X and Y after adjusting for Z and C"
#| layout-ncol: 2

# 1. X-Y relationship (unadjusted)
p1 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  ggplot2::geom_point(alpha = 0.3, color = "dodgerblue") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::annotate(
    "text", 
    x = min(dag_data$X) + 0.2*(max(dag_data$X)-min(dag_data$X)), 
    y = max(dag_data$Y) - 0.1*(max(dag_data$Y)-min(dag_data$Y)), 
    label = paste("Slope =", round(coef(lm(Y ~ X, dag_data))[2], 3)),
    hjust = 0
  ) +
  ggplot2::ggtitle("Unadjusted X-Y relationship")

# 2. Z-X relationship
p2 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = Z, y = X)) +
  ggplot2::geom_point(alpha = 0.3, color = "darkgreen") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::annotate(
    "text", 
    x = min(dag_data$Z) + 0.2*(max(dag_data$Z)-min(dag_data$Z)), 
    y = max(dag_data$X) - 0.1*(max(dag_data$X)-min(dag_data$X)), 
    label = paste("Slope =", round(coef(lm(X ~ Z, dag_data))[2], 3)),
    hjust = 0
  ) +
  ggplot2::ggtitle("Z → X relationship (confounder affects exposure)")

# 3. Z-Y relationship
p3 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = Z, y = Y)) +
  ggplot2::geom_point(alpha = 0.3, color = "purple") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::annotate(
    "text", 
    x = min(dag_data$Z) + 0.2*(max(dag_data$Z)-min(dag_data$Z)), 
    y = max(dag_data$Y) - 0.1*(max(dag_data$Y)-min(dag_data$Y)), 
    label = paste("Slope =", round(coef(lm(Y ~ Z, dag_data))[2], 3)),
    hjust = 0
  ) +
  ggplot2::ggtitle("Z → Y relationship (confounder affects outcome)")

# 4. X-Y relationship (adjusted for Z and C) - using residuals
# Create residuals
x_resid <- residuals(lm(X ~ Z + C, data = dag_data))
y_resid <- residuals(lm(Y ~ Z + C, data = dag_data))
resid_data <- data.frame(x_resid = x_resid, y_resid = y_resid)

# Plot relationship between residuals
p4 <- ggplot2::ggplot(resid_data, ggplot2::aes(x = x_resid, y = y_resid)) +
  ggplot2::geom_point(alpha = 0.3, color = "orange") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  ggplot2::theme_minimal() +
  ggplot2::annotate(
    "text", 
    x = min(resid_data$x_resid) + 0.2*(max(resid_data$x_resid)-min(resid_data$x_resid)), 
    y = max(resid_data$y_resid) - 0.1*(max(resid_data$y_resid)-min(resid_data$y_resid)), 
    label = paste("Slope =", round(coef(lm(y_resid ~ x_resid))[2], 3)),
    hjust = 0
  ) +
  ggplot2::xlab("X (residuals after controlling for Z and C)") +
  ggplot2::ylab("Y (residuals after controlling for Z and C)") +
  ggplot2::ggtitle("X → Y relationship after removing Z and C effects")

# Display all plots
print(p1)
print(p2)
print(p3)
print(p4)
```

## Conclusions from Backdoor Criterion Analysis:

The backdoor criterion visualization shows:

1. **Confounding creates bias**: The unadjusted X-Y relationship overestimates the true causal effect due to backdoor paths.

2. **Confounders affect both X and Y**: Z clearly influences both the exposure and outcome, creating confounding.

3. **Residual analysis isolates the causal effect**: After removing the effects of Z and C, the X-Y relationship approximates the true direct causal effect (0.3).

4. **Backdoor paths successfully blocked**: The adjusted relationship closely matches the true causal effect, confirming that adjusting for {Z, C} blocks all backdoor paths.

## 15. D-Separation Analysis

```{r}
#| label: d-separation-analysis
#| tbl-cap: "D-Separation Tests"

# Create manual tests for key relationships in the DAG
# 1. Test X and Y without conditioning (should be correlated)
cor_xy <- cor.test(dag_data$X, dag_data$Y)

# 2. Test X and Y conditioning on Z (should still be correlated due to remaining backdoor path through C)
model_x_z <- lm(X ~ Z, data = dag_data)
resid_x_given_z <- residuals(model_x_z)

model_y_z <- lm(Y ~ Z, data = dag_data)
resid_y_given_z <- residuals(model_y_z)

cor_xy_given_z <- cor.test(resid_x_given_z, resid_y_given_z)

# 3. Test X and Y conditioning on C (should still be correlated due to remaining backdoor path through Z)
model_x_c <- lm(X ~ C, data = dag_data)
resid_x_given_c <- residuals(model_x_c)

model_y_c <- lm(Y ~ C, data = dag_data)
resid_y_given_c <- residuals(model_y_c)

cor_xy_given_c <- cor.test(resid_x_given_c, resid_y_given_c)

# 4. Test X and Y conditioning on both Z and C (should be correlated only due to direct effect)
model_x_zc <- lm(X ~ Z + C, data = dag_data)
resid_x_given_zc <- residuals(model_x_zc)

model_y_zc <- lm(Y ~ Z + C, data = dag_data)
resid_y_given_zc <- residuals(model_y_zc)

cor_xy_given_zc <- cor.test(resid_x_given_zc, resid_y_given_zc)

# 5. Test X and Y conditioning on A, B, C (alternative sufficient adjustment set)
model_x_abc <- lm(X ~ A + B + C, data = dag_data)
resid_x_given_abc <- residuals(model_x_abc)

model_y_abc <- lm(Y ~ A + B + C, data = dag_data)
resid_y_given_abc <- residuals(model_y_abc)

cor_xy_given_abc <- cor.test(resid_x_given_abc, resid_y_given_abc)

# Compile results
independence_results <- data.frame(
  Claim = c("X ⊥ Y", "X ⊥ Y | Z", "X ⊥ Y | C", "X ⊥ Y | Z,C", "X ⊥ Y | A,B,C"),
  Correlation = c(
    cor_xy$estimate, 
    cor_xy_given_z$estimate,
    cor_xy_given_c$estimate,
    cor_xy_given_zc$estimate,
    cor_xy_given_abc$estimate
  ),
  `P-value` = c(
    cor_xy$p.value,
    cor_xy_given_z$p.value,
    cor_xy_given_c$p.value,
    cor_xy_given_zc$p.value,
    cor_xy_given_abc$p.value
  ),
  Independent = c(
    cor_xy$p.value > 0.05,
    cor_xy_given_z$p.value > 0.05,
    cor_xy_given_c$p.value > 0.05,
    cor_xy_given_zc$p.value > 0.05,
    cor_xy_given_abc$p.value > 0.05
  )
)

# Format the results
independence_results$Correlation <- round(as.numeric(independence_results$Correlation), 3)
independence_results

```

## 16. Sensitivity Analysis: Unmeasured Confounding

```{r}
#| label: sensitivity-analysis
#| fig-cap: "Impact of Unmeasured Confounding on Effect Estimates"

# Function to simulate unmeasured confounding
simulate_unmeasured_confounding <- function(cor_ux, cor_uy, n = 1000) {
  # Create a new dataset with an unmeasured confounder U
  u_unmeasured <- rnorm(n)
  
  # Create X correlated with unmeasured U and observed confounders
  x_unmeasured <- 0.3 * dag_data$A + 0.4 * dag_data$Z + 0.35 * dag_data$C + 
                  cor_ux * u_unmeasured + sqrt(1 - cor_ux^2) * rnorm(n, sd = 0.5)
  
  # Create Y correlated with unmeasured U, X, and observed confounders
  y_unmeasured <- 0.3 * x_unmeasured + 0.2 * dag_data$Z + 0.25 * dag_data$C + 
                  0.15 * dag_data$B + cor_uy * u_unmeasured + 
                  sqrt(1 - 0.3^2 - 0.2^2 - 0.25^2 - 0.15^2 - cor_uy^2) * rnorm(n, sd = 0.3)
  
  # Create dataset
  data.frame(X = x_unmeasured, Y = y_unmeasured, Z = dag_data$Z, C = dag_data$C, 
             A = dag_data$A, B = dag_data$B)
}

# Create a range of confounding strengths
u_effects <- seq(0, 0.5, by = 0.1)
sensitivity_results <- data.frame()

# For each confounding strength, calculate the bias
for (u_effect in u_effects) {
  # Simulate data with unmeasured confounding
  sim_data <- simulate_unmeasured_confounding(cor_ux = u_effect, cor_uy = u_effect)
  
  # Fit models with different adjustment strategies
  model_none <- lm(Y ~ X, data = sim_data)
  model_zc <- lm(Y ~ X + Z + C, data = sim_data)
  model_abc <- lm(Y ~ X + A + B + C, data = sim_data)
  
  # Store results
  sensitivity_results <- rbind(sensitivity_results, data.frame(
    UnmeasuredStrength = u_effect,
    Model = c("None", "Z,C", "A,B,C"),
    EstimatedEffect = c(coef(model_none)["X"], coef(model_zc)["X"], coef(model_abc)["X"]),
    TrueEffect = 0.3,
    Bias = c(coef(model_none)["X"] - 0.3, coef(model_zc)["X"] - 0.3, coef(model_abc)["X"] - 0.3),
    BiasPercent = 100 * c((coef(model_none)["X"] - 0.3) / 0.3, 
                         (coef(model_zc)["X"] - 0.3) / 0.3,
                         (coef(model_abc)["X"] - 0.3) / 0.3)
  ))
}

# Plot the results
ggplot2::ggplot(sensitivity_results, ggplot2::aes(x = UnmeasuredStrength, y = EstimatedEffect, color = Model)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
  ggplot2::scale_color_manual(values = c("None" = "red", "Z,C" = "blue", "A,B,C" = "green")) +
  ggplot2::labs(
    title = "Impact of Unmeasured Confounding on Effect Estimates",
    subtitle = "Red dashed line shows true causal effect (0.3)",
    x = "Strength of Unmeasured Confounder (correlation with X and Y)",
    y = "Estimated Effect of X on Y"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```



```{r}
#| label: sensitivity-table
#| tbl-cap: "Sensitivity to unmeasured confounding by adjustment strategy"

# Create a summary table
sensitivity_summary <- sensitivity_results |>
  dplyr::group_by(Model) |>
  dplyr::summarize(
    MinBias = min(abs(Bias)),
    MaxBias = max(abs(Bias)),
    AvgBias = mean(abs(Bias)),
    MaxBiasPercent = max(abs(BiasPercent)),
    .groups = 'drop'
  ) |>
  dplyr::mutate(
    MinBias = round(MinBias, 3),
    MaxBias = round(MaxBias, 3),
    AvgBias = round(AvgBias, 3),
    MaxBiasPercent = round(MaxBiasPercent, 2)
  )

DT::datatable(
  sensitivity_summary,
  caption = "Summary of Sensitivity to Unmeasured Confounding",
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## Conclusions from Sensitivity Analysis:

The sensitivity analysis reveals:

1. **All models are vulnerable to unmeasured confounding**: Even properly adjusted models show bias when unmeasured confounders are present.

2. **Proper adjustment reduces vulnerability**: Models with appropriate adjustment sets show less bias from unmeasured confounding than unadjusted models.

3. **Bias increases with confounder strength**: As the strength of the unmeasured confounder increases, bias in all models increases proportionally.

4. **Need for robustness checks**: Real-world studies should consider the potential impact of unmeasured confounders through sensitivity analyses.


## 17. Evaluating Bias Under Different Adjustment Strategies

```{r}
#| label: bias-evaluation
#| tbl-cap: "Comparison of Different Adjustment Strategies"

# Create different models representing adjustment strategies
all_models <- list(
  "None" = lm(Y ~ X, data = dag_data),
  "Z only" = lm(Y ~ X + Z, data = dag_data),
  "C only" = lm(Y ~ X + C, data = dag_data),
  "A only" = lm(Y ~ X + A, data = dag_data),
  "B only" = lm(Y ~ X + B, data = dag_data),
  "Z, C" = lm(Y ~ X + Z + C, data = dag_data),
  "A, B, C" = lm(Y ~ X + A + B + C, data = dag_data),
  "All variables" = lm(Y ~ X + Z + C + A + B, data = dag_data)
)

# Extract coefficients for X
adjustment_results <- data.frame(
  `Adjustment Set` = names(all_models),
  `X Coefficient` = sapply(all_models, function(m) coef(m)["X"]),
  `Std. Error` = sapply(all_models, function(m) summary(m)$coefficients["X", "Std. Error"]),
  `t-value` = sapply(all_models, function(m) summary(m)$coefficients["X", "t value"]),
  `p-value` = sapply(all_models, function(m) summary(m)$coefficients["X", "Pr(>|t|)"]),
  `R-squared` = sapply(all_models, function(m) summary(m)$r.squared)
)

# Calculate bias relative to the true effect
true_effect <- 0.3

adjustment_results <- adjustment_results |>
  dplyr::mutate(
    Bias = X.Coefficient - true_effect,
    `Percent Bias` = (Bias / true_effect) * 100,
    `Bias Category` = dplyr::case_when(
      abs(`Percent Bias`) < 5 ~ "Minimal bias (<5%)",
      abs(`Percent Bias`) < 10 ~ "Small bias (5-10%)",
      abs(`Percent Bias`) < 20 ~ "Moderate bias (10-20%)",
      TRUE ~ "Large bias (>20%)"
    )
  ) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(., 3)))

# Display the results
DT::datatable(
  adjustment_results,
  caption = "Comparison of Different Adjustment Strategies",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## Conclusions from Bias Evaluation:

The comprehensive bias evaluation shows:

1. **Minimal sufficient sets perform best**: Both {Z, C} and {A, B, C} produce estimates very close to the true effect with minimal bias.

2. **Partial adjustment is insufficient**: Adjusting for only one confounder leaves substantial bias from remaining backdoor paths.

3. **Overadjustment has minimal cost**: Including all variables doesn't significantly harm the estimate but may reduce precision.

4. **Clear hierarchy of adjustment quality**: Sufficient adjustment sets dramatically outperform insufficient ones.


## 18. Forest Plot Visualization of Causal Effects

```{r}
#| label: forest-plot
#| fig-cap: "Forest plot of causal effect estimates under different adjustment strategies"

# Create a forest plot of X coefficients
adjustment_results |>
  dplyr::mutate(`Adjustment Set` = factor(`Adjustment.Set`, 
                                  levels = rev(c("None", "A only", "B only", "C only", "Z only", 
                                               "Z, C", "A, B, C", "All variables")))) |>
  ggplot2::ggplot(ggplot2::aes(
    x = X.Coefficient, 
    y = `Adjustment Set`, 
    xmin = X.Coefficient - 1.96 * Std..Error, 
    xmax = X.Coefficient + 1.96 * Std..Error,
    color = `Adjustment Set` %in% c("Z, C", "A, B, C")
  )) +
  ggplot2::geom_pointrange(size = 0.8) +
  ggplot2::geom_vline(xintercept = true_effect, linetype = "dashed", color = "darkgreen", linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("red", "darkgreen"), 
    labels = c("Insufficient adjustment", "Sufficient adjustment")
  ) +
  ggplot2::labs(
    title = "Causal Effect Estimates Under Different Adjustment Strategies",
    subtitle = "Dashed line represents the true causal effect (0.3)",
    x = "Estimated Causal Effect of X on Y",
    y = "Adjustment Strategy",
    color = "Adjustment Quality"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## Conclusions from Forest Plot:

The forest plot clearly illustrates:

1. **Dramatic improvement with proper adjustment**: Sufficient adjustment sets cluster around the true effect.

2. **Consistent overestimation without adjustment**: Insufficient adjustment strategies consistently overestimate the causal effect.

3. **Precision vs. bias trade-off**: Proper adjustment reduces bias substantially with minimal loss of precision.

4. **Clear visual distinction**: The plot makes it immediately apparent which adjustment strategies are adequate.


## 19. Bayesian Causal Inference Analysis

Bayesian analysis provides several advantages for causal inference:
1. It expresses uncertainty through complete probability distributions rather than just point estimates
2. It allows incorporation of prior knowledge about causal relationships
3. It provides a more nuanced interpretation of uncertainty in causal effects

```{r}
#| label: bayesian-causal-analysis
#| message: false
#| warning: false
#| results: 'hide'

# Standardize variables for better model fitting
dag_data_std <- dag_data |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), scale)) |>
  as.data.frame()

# Define and fit Bayesian models for different adjustment sets
# 1. No adjustment (biased)
m_none <- rethinking::quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dag_data_std
)

# 2. Adjusting for Z and C (minimal sufficient set 1)
m_zc <- rethinking::quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z + bC * C,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    bZ ~ dnorm(0, 1),
    bC ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dag_data_std
)

# 3. Adjusting for A, B, C (minimal sufficient set 2)
m_abc <- rethinking::quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bA * A + bB * B + bC * C,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    bB ~ dnorm(0, 1),
    bC ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dag_data_std
)

# 4. Full model with all variables
m_full <- rethinking::quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z + bC * C + bA * A + bB * B,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    bZ ~ dnorm(0, 1),
    bC ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    bB ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dag_data_std
)
```

```{r}
#| label: extract-bayesian-posteriors
#| tbl-cap: "Bayesian estimates of the causal effect of X on Y"

# Extract samples from the posterior distributions
post_none <- rethinking::extract.samples(m_none)
post_zc <- rethinking::extract.samples(m_zc)
post_abc <- rethinking::extract.samples(m_abc)
post_full <- rethinking::extract.samples(m_full)

# Create a function to summarize posteriors
summarize_posterior <- function(posterior, name) {
  data.frame(
    Adjustment_Set = name,
    Mean = mean(posterior$bX),
    Median = median(posterior$bX),
    SD = sd(posterior$bX),
    CI_Lower = quantile(posterior$bX, 0.025),
    CI_Upper = quantile(posterior$bX, 0.975),
    Width = quantile(posterior$bX, 0.975) - quantile(posterior$bX, 0.025)
  )
}

# Summarize the models
bayesian_results <- rbind(
  summarize_posterior(post_none, "None"),
  summarize_posterior(post_zc, "Z, C"),
  summarize_posterior(post_abc, "A, B, C"),
  summarize_posterior(post_full, "All variables")
)

# Display the results
DT::datatable(
  bayesian_results, 
  caption = "Bayesian estimates of the causal effect of X on Y under different adjustment strategies",
  options = list(pageLength = 5, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
) |>
  DT::formatRound(columns = c("Mean", "Median", "SD", "CI_Lower", "CI_Upper", "Width"), digits = 3)
```

```{r}
#| label: plot-bayesian-posteriors
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Posterior distributions of causal effect estimates under different adjustment strategies"

# Create a data frame with the posterior samples
all_posteriors <- data.frame(
  None = post_none$bX,
  `Z, C` = post_zc$bX,
  `A, B, C` = post_abc$bX,
  `All variables` = post_full$bX,
  check.names = FALSE
)

# Convert to long format for plotting
long_posteriors <- all_posteriors |>
  tidyr::pivot_longer(
    cols = dplyr::everything(), 
    names_to = "Adjustment_Set", 
    values_to = "Effect_Estimate"
  )

# Set factor levels for consistent ordering
long_posteriors$Adjustment_Set <- factor(
  long_posteriors$Adjustment_Set,
  levels = c("None", "Z, C", "A, B, C", "All variables")
)

# Plot density curves for all adjustment sets
ggplot2::ggplot(long_posteriors, ggplot2::aes(x = Effect_Estimate, fill = Adjustment_Set)) +
  ggplot2::geom_density(alpha = 0.6) +
  ggplot2::geom_vline(
    data = bayesian_results, 
    ggplot2::aes(xintercept = Mean, color = Adjustment_Set),
    linetype = "dashed", 
    linewidth = 1
  ) +
  ggplot2::scale_fill_brewer(palette = "Set1") +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::labs(
    title = "Posterior Distributions of the Causal Effect of X on Y",
    subtitle = "Under different adjustment strategies (standardized coefficients)",
    x = "Causal Effect (Standardized)",
    y = "Density"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::guides(
    fill = ggplot2::guide_legend(title = "Adjustment Strategy"),
    color = ggplot2::guide_legend(title = "Adjustment Strategy")
  )
```

```{r}
#| label: bayesian-forest-plot
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Forest plot of Bayesian causal effect estimates"

# Create forest plot
ggplot2::ggplot(
  bayesian_results, 
  ggplot2::aes(
    x = Mean, 
    y = Adjustment_Set, 
    xmin = CI_Lower, 
    xmax = CI_Upper,
    color = Adjustment_Set %in% c("Z, C", "A, B, C")
  )
) +
  ggplot2::geom_pointrange(size = 1.2) +
  ggplot2::scale_color_manual(
    values = c("red", "darkgreen"), 
    labels = c("Insufficient", "Sufficient")
  ) +
  ggplot2::labs(
    title = "Bayesian Causal Effect Estimates Under Different Adjustment Strategies",
    subtitle = "Credible intervals show uncertainty in causal effect estimates",
    x = "Estimated Causal Effect of X on Y (Standardized)",
    y = "Adjustment Strategy",
    color = "Adjustment Quality"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## Interpretation of Bayesian Causal Analysis:

The Bayesian analysis provides several key insights:

1. **Uncertainty quantification**: The posterior distributions show not just point estimates but the full uncertainty around the causal effect.

2. **Consistent findings**: Both minimal sufficient adjustment sets (Z,C and A,B,C) yield very similar posterior distributions, confirming they are equivalent for causal identification.

3. **Clear bias in insufficient adjustment**: The "None" model shows a clearly shifted distribution, indicating systematic bias.

4. **Precision vs. bias trade-off**: Proper adjustment reduces bias dramatically while maintaining reasonable precision.


## 20. Counterfactual Analysis: "What If" Scenarios

Counterfactual analysis answers the question: "What would happen to Y if it intervened to change X, while holding confounders constant?"

```{r}
#| label: counterfactual-analysis
#| fig-cap: "Counterfactual predictions under intervention"

# Function to predict Y based on do(X = x)
predict_counterfactual <- function(x_values, fixed_z = NULL, fixed_c = NULL) {
  # Use the coefficients from the adjusted model (Z, C)
  adjusted_model <- lm(Y ~ X + Z + C, data = dag_data)
  intercept <- coef(adjusted_model)[1]
  x_coef <- coef(adjusted_model)[2]
  z_coef <- coef(adjusted_model)[3]
  c_coef <- coef(adjusted_model)[4]
  
  # Use mean values if not specified
  if(is.null(fixed_z)) fixed_z <- mean(dag_data$Z)
  if(is.null(fixed_c)) fixed_c <- mean(dag_data$C)
  
  # Predict Y for different values of X, holding confounders constant
  y_pred <- intercept + x_coef * x_values + z_coef * fixed_z + c_coef * fixed_c
  return(y_pred)
}

# Create a range of X values for intervention
x_range <- seq(min(dag_data$X), max(dag_data$X), length.out = 100)

# Predict Y for different values of X, holding Z and C at their means
z_mean <- mean(dag_data$Z)
c_mean <- mean(dag_data$C)
y_pred_mean <- predict_counterfactual(x_range, fixed_z = z_mean, fixed_c = c_mean)

# Create a data frame for plotting
counterfactual_df <- data.frame(X = x_range, Y = y_pred_mean)

# Plot the counterfactual prediction
ggplot2::ggplot() +
  # Add actual data points
  ggplot2::geom_point(data = dag_data, ggplot2::aes(x = X, y = Y), alpha = 0.2, color = "gray") +
  # Add counterfactual prediction
  ggplot2::geom_line(
    data = counterfactual_df, 
    ggplot2::aes(x = X, y = Y), 
    color = "red", 
    linewidth = 1.5
  ) +
  ggplot2::labs(
    title = "Counterfactual Prediction: What if it intervene on X?",
    subtitle = paste("Holding Z constant at", round(z_mean, 2), "and C constant at", round(c_mean, 2)),
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal()
```

```{r}
#| label: counterfactual-multiple-scenarios
#| fig-cap: "Counterfactual predictions for different confounder scenarios"

# Create predictions for different Z and C values
z_values <- quantile(dag_data$Z, probs = c(0.1, 0.5, 0.9))
c_values <- quantile(dag_data$C, probs = c(0.1, 0.5, 0.9))

# Create combinations of Z and C values
scenarios <- expand.grid(Z = z_values, C = c_values)
predictions_list <- list()

for(i in 1:nrow(scenarios)) {
  z_val <- scenarios$Z[i]
  c_val <- scenarios$C[i]
  y_pred <- predict_counterfactual(x_range, fixed_z = z_val, fixed_c = c_val)
  
  scenario_name <- paste0("Z=", round(z_val, 2), ", C=", round(c_val, 2))
  predictions_list[[scenario_name]] <- data.frame(
    X = x_range,
    Y = y_pred,
    Scenario = scenario_name,
    Z_level = ifelse(z_val == z_values[1], "Low Z", 
                    ifelse(z_val == z_values[2], "Medium Z", "High Z")),
    C_level = ifelse(c_val == c_values[1], "Low C", 
                    ifelse(c_val == c_values[2], "Medium C", "High C"))
  )
}

# Combine all predictions
all_predictions <- do.call(rbind, predictions_list)

# Plot the counterfactual predictions for different scenarios
ggplot2::ggplot(all_predictions, ggplot2::aes(x = X, y = Y, color = interaction(Z_level, C_level))) +
  ggplot2::geom_line(linewidth = 1.0) +
  ggplot2::scale_color_viridis_d(name = "Scenario\n(Z level, C level)") +
  ggplot2::labs(
    title = "Counterfactual Predictions for Different Confounder Scenarios",
    subtitle = "Each line shows the causal effect of X on Y for specific Z and C values",
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "right")
```

```{r}
#| label: treatment-effect-table
#| tbl-cap: "Estimated treatment effects under different interventions"

# Calculate treatment effects for specific interventions
intervention_effects <- data.frame(
  Intervention = c(
    "Increase X by 1 unit",
    "Increase X by 2 units", 
    "Decrease X by 1 unit",
    "Move X from 25th to 75th percentile"
  ),
  X_Change = c(
    1,
    2,
    -1,
    quantile(dag_data$X, 0.75) - quantile(dag_data$X, 0.25)
  ),
  Expected_Y_Change = c(
    1 * coef(lm(Y ~ X + Z + C, data = dag_data))["X"],
    2 * coef(lm(Y ~ X + Z + C, data = dag_data))["X"],
    -1 * coef(lm(Y ~ X + Z + C, data = dag_data))["X"],
    (quantile(dag_data$X, 0.75) - quantile(dag_data$X, 0.25)) * coef(lm(Y ~ X + Z + C, data = dag_data))["X"]
  )
)

# Format the results
intervention_effects$X_Change <- round(intervention_effects$X_Change, 3)
intervention_effects$Expected_Y_Change <- round(intervention_effects$Expected_Y_Change, 3)

DT::datatable(
  intervention_effects,
  caption = "Expected outcomes under different interventions on X",
  options = list(pageLength = 10, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact responsive'
)
```

## Conclusions from Counterfactual Analysis:

The counterfactual analysis demonstrates:

1. **Clear intervention effects**: The red line represents the causal effect of X on Y, isolated from confounding by holding Z and C constant.

2. **Parallel intervention effects**: Across different confounder scenarios, the slope (causal effect) remains constant, but intercepts vary based on confounder levels.

3. **Practical policy implications**: We can quantify the expected outcome of specific interventions on X, which is crucial for policy and decision-making.

4. **Robustness across scenarios**: The causal effect estimate remains consistent across different combinations of confounder values.







## 21. Practical Implications and Conclusions

### 21.1 Summary of Key Findings

The comprehensive analysis of the complex DAG structure has revealed several critical insights:

1. **Multiple adjustment strategies work**: Both {Z, C} and {A, B, C} successfully identify the true causal effect of X on Y.

2. **Partial adjustment is dangerous**: Controlling for only some confounders can leave substantial bias, sometimes worse than no adjustment.

3. **Robust methodology matters**: Frequentist, Bayesian, and SEM approaches all converge on similar conclusions when properly applied.

4. **Confounding has multiple sources**: Complex real-world scenarios often involve multiple backdoor paths that must all be blocked.


### 21.2 Real-World Application Context

In practical terms, this analysis framework applies to numerous real-world scenarios:

- **Healthcare**: Studying medication effectiveness while controlling for patient characteristics and disease severity
- **Economics**: Evaluating policy interventions while accounting for socioeconomic confounders
- **Education**: Assessing program effectiveness while controlling for student and institutional factors
- **Marketing**: Measuring campaign effectiveness while accounting for customer and market characteristics


### 21.3 Methodological Insights

Key methodological lessons include:

1. **DAG-based thinking is essential**: Theoretical understanding of causal relationships should guide empirical analysis
2. **Multiple validation approaches strengthen conclusions**: Using different statistical frameworks to confirm findings
3. **Sensitivity analysis is crucial**: Understanding robustness to unmeasured confounding
4. **Visualization aids understanding**: Graphical representations help communicate complex causal relationships


### 21.4 Limitations and Extensions

This analysis has several limitations that point to future extensions:

1. **Linear relationships assumed**: Real relationships may be non-linear
2. **No time dimension**: Dynamic causal relationships over time not considered
3. **No measurement error**: Variables assumed to be measured without error
4. **No effect modification**: Causal effects assumed constant across populations


### 21.5 Recommendations for Empirical Research

Based on this analysis, it recommend:

1. **Start with causal theory**: Develop DAGs based on domain knowledge before data analysis
2. **Identify minimal sufficient adjustment sets**: Use causal identification theory to select controls
3. **Test implications**: Use conditional independence tests to validate causal assumptions
4. **Conduct sensitivity analyses**: Assess robustness to potential unmeasured confounding
5. **Use multiple analytical approaches**: Triangulate findings across different methodological frameworks







## 22. Simpson's Paradox Detection Analysis

Simpson's Paradox occurs when the direction of an association between two variables reverses when conditioning on a third variable. In the complex DAG structure, it can examine whether this phenomenon occurs by stratifying the data based on different levels of the confounders Z and C.

Simpson's Paradox in causal inference resembles a statistical illusion - what appears true at the aggregate level completely contradicts what occurs within subgroups. In the DAG example, it might find that overall, X appears to have one relationship with Y, but within each level of the confounders, the relationship tells a different story entirely.

### 22.1 Simpson's Paradox Detection Analysis

First, let's examine the overall relationship between X and Y, and then compare it to the relationship within different strata of the confounders:

```{r}
#| label: detect-simpsons-paradox-complex
#| message: false
#| warning: false

# Calculate overall correlation between X and Y
overall_cor <- cor(dag_data$X, dag_data$Y)
overall_slope <- coef(lm(Y ~ X, data = dag_data))[2]

# Create Z quartiles for stratified analysis
dag_data$Z_quartile <- cut(
  dag_data$Z, 
  breaks = quantile(dag_data$Z, probs = c(0, 0.25, 0.5, 0.75, 1)), 
  labels = c("Q1 (Low)", "Q2", "Q3", "Q4 (High)"),
  include.lowest = TRUE
)

# Create C quartiles for stratified analysis
dag_data$C_quartile <- cut(
  dag_data$C, 
  breaks = quantile(dag_data$C, probs = c(0, 0.25, 0.5, 0.75, 1)), 
  labels = c("Q1 (Low)", "Q2", "Q3", "Q4 (High)"),
  include.lowest = TRUE
)

# Calculate correlations and slopes within each Z quartile
stratified_results_z <- dag_data |>
  dplyr::group_by(Z_quartile) |>
  dplyr::summarise(
    n = dplyr::n(),
    cor_XY = cor(X, Y),
    slope_XY = coef(lm(Y ~ X))[2],
    mean_Z = mean(Z),
    mean_X = mean(X),
    mean_Y = mean(Y),
    .groups = 'drop'
  )

# Calculate correlations and slopes within each C quartile
stratified_results_c <- dag_data |>
  dplyr::group_by(C_quartile) |>
  dplyr::summarise(
    n = dplyr::n(),
    cor_XY = cor(X, Y),
    slope_XY = coef(lm(Y ~ X))[2],
    mean_C = mean(C),
    mean_X = mean(X),
    mean_Y = mean(Y),
    .groups = 'drop'
  )

# Add overall results for comparison
overall_results <- data.frame(
  Group = "Overall",
  n = nrow(dag_data),
  cor_XY = overall_cor,
  slope_XY = overall_slope,
  mean_Confounder = NA,
  mean_X = mean(dag_data$X),
  mean_Y = mean(dag_data$Y)
)

# Combine Z results
z_results <- stratified_results_z |>
  dplyr::mutate(
    Group = as.character(Z_quartile),
    mean_Confounder = mean_Z
  ) |>
  dplyr::select(Group, n, cor_XY, slope_XY, mean_Confounder, mean_X, mean_Y)

# Combine C results  
c_results <- stratified_results_c |>
  dplyr::mutate(
    Group = as.character(C_quartile),
    mean_Confounder = mean_C
  ) |>
  dplyr::select(Group, n, cor_XY, slope_XY, mean_Confounder, mean_X, mean_Y)

# Combine all results
simpson_analysis_z <- dplyr::bind_rows(overall_results, z_results)
simpson_analysis_c <- dplyr::bind_rows(overall_results, c_results)

# Round for display
simpson_analysis_z <- simpson_analysis_z |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(., 3)))

simpson_analysis_c <- simpson_analysis_c |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(., 3)))

# Display Z results
DT::datatable(
  simpson_analysis_z,
  caption = "Simpson's Paradox Detection: Overall vs Stratified by Z Analysis",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("Z Group", "N", "Correlation X-Y", "Slope X→Y", 
              "Mean Z", "Mean X", "Mean Y")
) |>
  DT::formatRound(columns = c("cor_XY", "slope_XY", "mean_Confounder", "mean_X", "mean_Y"), digits = 3)

# Display C results
DT::datatable(
  simpson_analysis_c,
  caption = "Simpson's Paradox Detection: Overall vs Stratified by C Analysis",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("C Group", "N", "Correlation X-Y", "Slope X→Y", 
              "Mean C", "Mean X", "Mean Y")
) |>
  DT::formatRound(columns = c("cor_XY", "slope_XY", "mean_Confounder", "mean_X", "mean_Y"), digits = 3)
```

### 22.2 Testing for Simpson's Paradox by Status

Let's formally test whether Simpson's Paradox is present by examining if the direction of association changes:

```{r}
#| label: test-simpsons-paradox-status-complex
#| message: false
#| warning: false

# Function to determine Simpson's Paradox status
detect_simpson_status <- function(overall_slope, stratified_slopes) {
  # Check if all stratified slopes have the same sign
  stratified_signs <- sign(stratified_slopes)
  overall_sign <- sign(overall_slope)
  
  # Simpson's Paradox occurs when overall sign differs from all stratified signs
  if (all(stratified_signs == stratified_signs[1]) && 
      overall_sign != stratified_signs[1]) {
    return("Strong Simpson's Paradox")
  } else if (any(stratified_signs != overall_sign)) {
    return("Partial Simpson's Paradox")
  } else {
    return("No Simpson's Paradox")
  }
}

# Get stratified slopes (excluding overall)
stratified_slopes_z <- stratified_results_z$slope_XY
stratified_slopes_c <- stratified_results_c$slope_XY

# Determine Simpson's Paradox status for each confounder
simpson_status_z <- detect_simpson_status(overall_slope, stratified_slopes_z)
simpson_status_c <- detect_simpson_status(overall_slope, stratified_slopes_c)

# Create summary table
simpson_summary <- data.frame(
  Measure = c("Overall Slope", "Mean Stratified Slope (Z)", "Range of Stratified Slopes (Z)",
              "Mean Stratified Slope (C)", "Range of Stratified Slopes (C)",
              "Overall Correlation", "Mean Stratified Correlation (Z)", 
              "Mean Stratified Correlation (C)",
              "Simpson's Paradox Status (Z)", "Simpson's Paradox Status (C)"),
  Value = c(
    round(overall_slope, 3),
    round(mean(stratified_slopes_z), 3),
    paste(round(min(stratified_slopes_z), 3), "to", round(max(stratified_slopes_z), 3)),
    round(mean(stratified_slopes_c), 3),
    paste(round(min(stratified_slopes_c), 3), "to", round(max(stratified_slopes_c), 3)),
    round(overall_cor, 3),
    round(mean(stratified_results_z$cor_XY), 3),
    round(mean(stratified_results_c$cor_XY), 3),
    simpson_status_z,
    simpson_status_c
  )
)

# Display summary
DT::datatable(
  simpson_summary,
  caption = "Simpson's Paradox Status Summary",
  options = list(pageLength = 15, dom = 't'),
  colnames = c("Measure", "Value")
)
```

### 22.3 Testing for Simpson's Paradox by Groups

Let's examine the phenomenon more systematically by creating binary high/low groups for both Z and C:

```{r}
#| label: test-simpsons-by-groups-complex
#| message: false
#| warning: false

# Create binary Z and C groups (high/low based on median split)
dag_data$Z_binary <- ifelse(dag_data$Z > median(dag_data$Z), "High Z", "Low Z")
dag_data$C_binary <- ifelse(dag_data$C > median(dag_data$C), "High C", "Low C")

# Calculate slopes and correlations for binary Z groups
binary_analysis_z <- dag_data |>
  dplyr::group_by(Z_binary) |>
  dplyr::summarise(
    n = dplyr::n(),
    cor_XY = cor(X, Y),
    slope_XY = coef(lm(Y ~ X))[2],
    intercept_XY = coef(lm(Y ~ X))[1],
    mean_Z = mean(Z),
    mean_X = mean(X),
    mean_Y = mean(Y),
    .groups = 'drop'
  )

# Calculate slopes and correlations for binary C groups
binary_analysis_c <- dag_data |>
  dplyr::group_by(C_binary) |>
  dplyr::summarise(
    n = dplyr::n(),
    cor_XY = cor(X, Y),
    slope_XY = coef(lm(Y ~ X))[2],
    intercept_XY = coef(lm(Y ~ X))[1],
    mean_C = mean(C),
    mean_X = mean(X),
    mean_Y = mean(Y),
    .groups = 'drop'
  )

# Add confidence intervals for slopes - Z groups
binary_analysis_z$slope_se <- NA
binary_analysis_z$slope_ci_lower <- NA
binary_analysis_z$slope_ci_upper <- NA

for (group in unique(dag_data$Z_binary)) {
  group_data <- dag_data[dag_data$Z_binary == group, ]
  model <- lm(Y ~ X, data = group_data)
  slope_se <- summary(model)$coefficients["X", "Std. Error"]
  slope_ci <- confint(model)["X", ]
  
  binary_analysis_z[binary_analysis_z$Z_binary == group, "slope_se"] <- slope_se
  binary_analysis_z[binary_analysis_z$Z_binary == group, "slope_ci_lower"] <- slope_ci[1]
  binary_analysis_z[binary_analysis_z$Z_binary == group, "slope_ci_upper"] <- slope_ci[2]
}

# Add confidence intervals for slopes - C groups
binary_analysis_c$slope_se <- NA
binary_analysis_c$slope_ci_lower <- NA
binary_analysis_c$slope_ci_upper <- NA

for (group in unique(dag_data$C_binary)) {
  group_data <- dag_data[dag_data$C_binary == group, ]
  model <- lm(Y ~ X, data = group_data)
  slope_se <- summary(model)$coefficients["X", "Std. Error"]
  slope_ci <- confint(model)["X", ]
  
  binary_analysis_c[binary_analysis_c$C_binary == group, "slope_se"] <- slope_se
  binary_analysis_c[binary_analysis_c$C_binary == group, "slope_ci_lower"] <- slope_ci[1]
  binary_analysis_c[binary_analysis_c$C_binary == group, "slope_ci_upper"] <- slope_ci[2]
}

# Round for display
binary_analysis_z <- binary_analysis_z |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(., 3)))

binary_analysis_c <- binary_analysis_c |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(., 3)))

# Display binary group analysis for Z
DT::datatable(
  binary_analysis_z,
  caption = "Binary Group Analysis for Simpson's Paradox Detection (Z Confounder)",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("Z Group", "N", "Correlation", "Slope", "Intercept", 
              "Mean Z", "Mean X", "Mean Y", "Slope SE", "CI Lower", "CI Upper")
) |>
  DT::formatRound(
    columns = c("cor_XY", "slope_XY", "intercept_XY", "mean_Z", 
                "mean_X", "mean_Y", "slope_se", "slope_ci_lower", "slope_ci_upper"), 
    digits = 3
  )

# Display binary group analysis for C
DT::datatable(
  binary_analysis_c,
  caption = "Binary Group Analysis for Simpson's Paradox Detection (C Confounder)",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("C Group", "N", "Correlation", "Slope", "Intercept", 
              "Mean C", "Mean X", "Mean Y", "Slope SE", "CI Lower", "CI Upper")
) |>
  DT::formatRound(
    columns = c("cor_XY", "slope_XY", "intercept_XY", "mean_C", 
                "mean_X", "mean_Y", "slope_se", "slope_ci_lower", "slope_ci_upper"), 
    digits = 3
  )
```

### 22.4 Visual Detection of Simpson's Paradox

Let's create visualizations to clearly demonstrate whether Simpson's Paradox is present in the data:

```{r}
#| label: visualize-simpsons-paradox-complex
#| fig-cap: "Visual Detection of Simpson's Paradox"
#| fig-subcap: 
#|   - "Overall relationship vs stratified relationships (Z quartiles)"
#|   - "Overall relationship vs stratified relationships (C quartiles)"
#|   - "Binary group analysis with separate regression lines (Z)"
#|   - "Binary group analysis with separate regression lines (C)"
#|   - "Slope comparison across different stratifications"
#| layout-ncol: 1

# Plot 1: Overall vs Stratified (Z Quartiles)
p1 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  # Overall regression line
  ggplot2::geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    color = "red", 
    linewidth = 1.5, 
    linetype = "dashed", 
    se = FALSE
  ) +
  # Stratified regression lines
  ggplot2::geom_smooth(
    ggplot2::aes(color = Z_quartile), 
    method = "lm", 
    formula = y ~ x, 
    linewidth = 1, 
    se = FALSE
  ) +
  # Points colored by quartile
  ggplot2::geom_point(ggplot2::aes(color = Z_quartile), alpha = 0.6) +
  ggplot2::scale_color_viridis_d(name = "Z Quartile") +
  ggplot2::labs(
    title = "Simpson's Paradox Analysis: Overall vs Z-Stratified Relationships",
    subtitle = paste("Overall slope (red dashed):", round(overall_slope, 3),
                    "| Simpson's Status (Z):", simpson_status_z),
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")

# Plot 2: Overall vs Stratified (C Quartiles)
p2 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  # Overall regression line
  ggplot2::geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    color = "red", 
    linewidth = 1.5, 
    linetype = "dashed", 
    se = FALSE
  ) +
  # Stratified regression lines
  ggplot2::geom_smooth(
    ggplot2::aes(color = C_quartile), 
    method = "lm", 
    formula = y ~ x, 
    linewidth = 1, 
    se = FALSE
  ) +
  # Points colored by quartile
  ggplot2::geom_point(ggplot2::aes(color = C_quartile), alpha = 0.6) +
  ggplot2::scale_color_brewer(palette = "Set2", name = "C Quartile") +
  ggplot2::labs(
    title = "Simpson's Paradox Analysis: Overall vs C-Stratified Relationships",
    subtitle = paste("Overall slope (red dashed):", round(overall_slope, 3),
                    "| Simpson's Status (C):", simpson_status_c),
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")

# Plot 3: Binary Group Analysis (Z)
p3 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  # Overall regression line
  ggplot2::geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    color = "red", 
    linewidth = 1.5, 
    linetype = "dashed", 
    se = FALSE
  ) +
  # Stratified regression lines for binary groups
  ggplot2::geom_smooth(
    ggplot2::aes(color = Z_binary), 
    method = "lm", 
    formula = y ~ x, 
    linewidth = 1.2, 
    se = TRUE, 
    alpha = 0.3
  ) +
  # Points colored by binary group
  ggplot2::geom_point(ggplot2::aes(color = Z_binary), alpha = 0.6) +
  ggplot2::scale_color_manual(
    values = c("High Z" = "darkblue", "Low Z" = "darkgreen"),
    name = "Z Group"
  ) +
  ggplot2::labs(
    title = "Binary Z Group Analysis for Simpson's Paradox",
    subtitle = "Red dashed line: Overall relationship | Colored lines: Z group-specific relationships",
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")

# Plot 4: Binary Group Analysis (C)
p4 <- ggplot2::ggplot(dag_data, ggplot2::aes(x = X, y = Y)) +
  # Overall regression line
  ggplot2::geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    color = "red", 
    linewidth = 1.5, 
    linetype = "dashed", 
    se = FALSE
  ) +
  # Stratified regression lines for binary groups
  ggplot2::geom_smooth(
    ggplot2::aes(color = C_binary), 
    method = "lm", 
    formula = y ~ x, 
    linewidth = 1.2, 
    se = TRUE, 
    alpha = 0.3
  ) +
  # Points colored by binary group
  ggplot2::geom_point(ggplot2::aes(color = C_binary), alpha = 0.6) +
  ggplot2::scale_color_manual(
    values = c("High C" = "darkorange", "Low C" = "purple"),
    name = "C Group"
  ) +
  ggplot2::labs(
    title = "Binary C Group Analysis for Simpson's Paradox",
    subtitle = "Red dashed line: Overall relationship | Colored lines: C group-specific relationships",
    x = "X (Exposure)",
    y = "Y (Outcome)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")

# Plot 5: Slope Comparison
slope_comparison <- data.frame(
  Group = c("Overall", paste("Z", stratified_results_z$Z_quartile), 
            paste("C", stratified_results_c$C_quartile),
            paste("Z", binary_analysis_z$Z_binary), 
            paste("C", binary_analysis_c$C_binary)),
  Slope = c(overall_slope, stratified_results_z$slope_XY, stratified_results_c$slope_XY,
            binary_analysis_z$slope_XY, binary_analysis_c$slope_XY),
  Type = c("Overall", rep("Z Quartile", 4), rep("C Quartile", 4), 
           rep("Z Binary", 2), rep("C Binary", 2))
)

p5 <- ggplot2::ggplot(slope_comparison, ggplot2::aes(x = reorder(Group, Slope), y = Slope, fill = Type)) +
  ggplot2::geom_col(alpha = 0.8) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  ggplot2::geom_hline(yintercept = 0.3, linetype = "dotted", color = "red", alpha = 0.7) +
  ggplot2::scale_fill_brewer(palette = "Set3") +
  ggplot2::labs(
    title = "Slope Comparison Across Different Stratifications",
    subtitle = "Red dotted line shows true causal effect (0.3)",
    x = "Group",
    y = "Slope of X → Y"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

# Display plots
p1
p2
p3
p4
p5
```

### 22.5 Formal Statistical Test for Simpson's Paradox

Let's conduct formal statistical tests to quantify the evidence for Simpson's Paradox:

```{r}
#| label: formal-test-simpsons-paradox-complex
#| message: false
#| warning: false

# Function to test for significant differences in slopes
test_slope_differences <- function(overall_slope, overall_se, stratified_slopes, stratified_ses) {
  # Test if overall slope differs significantly from each stratified slope
  z_stats <- (overall_slope - stratified_slopes) / sqrt(overall_se^2 + stratified_ses^2)
  p_values <- 2 * (1 - pnorm(abs(z_stats)))
  
  return(list(z_stats = z_stats, p_values = p_values))
}

# Get standard errors for slopes
overall_model <- lm(Y ~ X, data = dag_data)
overall_se <- summary(overall_model)$coefficients["X", "Std. Error"]

# Get standard errors for Z-stratified models
stratified_ses_z <- numeric(length(stratified_slopes_z))
for (i in 1:length(unique(dag_data$Z_quartile))) {
  quartile <- levels(dag_data$Z_quartile)[i]
  quartile_data <- dag_data[dag_data$Z_quartile == quartile, ]
  model <- lm(Y ~ X, data = quartile_data)
  stratified_ses_z[i] <- summary(model)$coefficients["X", "Std. Error"]
}

# Get standard errors for C-stratified models
stratified_ses_c <- numeric(length(stratified_slopes_c))
for (i in 1:length(unique(dag_data$C_quartile))) {
  quartile <- levels(dag_data$C_quartile)[i]
  quartile_data <- dag_data[dag_data$C_quartile == quartile, ]
  model <- lm(Y ~ X, data = quartile_data)
  stratified_ses_c[i] <- summary(model)$coefficients["X", "Std. Error"]
}

# Perform tests for Z stratification
slope_tests_z <- test_slope_differences(overall_slope, overall_se, stratified_slopes_z, stratified_ses_z)

# Perform tests for C stratification
slope_tests_c <- test_slope_differences(overall_slope, overall_se, stratified_slopes_c, stratified_ses_c)

# Create test results table for Z
test_results_z <- data.frame(
  Comparison = paste("Overall vs", stratified_results_z$Z_quartile),
  Overall_Slope = rep(round(overall_slope, 3), length(stratified_slopes_z)),
  Stratified_Slope = round(stratified_slopes_z, 3),
  Difference = round(overall_slope - stratified_slopes_z, 3),
  Z_Statistic = round(slope_tests_z$z_stats, 3),
  P_Value = round(slope_tests_z$p_values, 3),
  Significant = ifelse(slope_tests_z$p_values < 0.05, "Yes", "No")
)

# Create test results table for C
test_results_c <- data.frame(
  Comparison = paste("Overall vs", stratified_results_c$C_quartile),
  Overall_Slope = rep(round(overall_slope, 3), length(stratified_slopes_c)),
  Stratified_Slope = round(stratified_slopes_c, 3),
  Difference = round(overall_slope - stratified_slopes_c, 3),
  Z_Statistic = round(slope_tests_c$z_stats, 3),
  P_Value = round(slope_tests_c$p_values, 3),
  Significant = ifelse(slope_tests_c$p_values < 0.05, "Yes", "No")
)

# Display test results for Z
DT::datatable(
  test_results_z,
  caption = "Formal Statistical Tests for Slope Differences (Z Stratification)",
  options = list(pageLength = 10, scrollX = TRUE)
) |>
  DT::formatRound(
    columns = c("Overall_Slope", "Stratified_Slope", "Difference", 
                "Z_Statistic", "P_Value"), 
    digits = 3
  )

# Display test results for C
DT::datatable(
  test_results_c,
  caption = "Formal Statistical Tests for Slope Differences (C Stratification)",
  options = list(pageLength = 10, scrollX = TRUE)
) |>
  DT::formatRound(
    columns = c("Overall_Slope", "Stratified_Slope", "Difference", 
                "Z_Statistic", "P_Value"), 
    digits = 3
  )
```

### 22.6 Magnitude of Simpson's Paradox Effect

Let's quantify the magnitude of the Simpson's Paradox effect if it exists:

```{r}
#| label: magnitude-simpsons-effect-complex
#| message: false
#| warning: false

# Calculate Simpson's Paradox magnitude measures for Z
simpson_magnitude_z <- data.frame(
  Measure = c(
    "Overall Slope",
    "Average Z-Stratified Slope",
    "Absolute Difference (Z)",
    "Relative Difference (Z) (%)",
    "Direction Reversal (Z)",
    "Weighted Average Slope (Z)",
    "Simpson's Paradox Strength (Z)"
  ),
  Value = c(
    round(overall_slope, 3),
    round(mean(stratified_slopes_z), 3),
    round(abs(overall_slope - mean(stratified_slopes_z)), 3),
    round(100 * abs(overall_slope - mean(stratified_slopes_z)) / abs(mean(stratified_slopes_z)), 3),
    ifelse(sign(overall_slope) != sign(mean(stratified_slopes_z)), "Yes", "No"),
    round(sum(stratified_results_z$slope_XY * stratified_results_z$n) / sum(stratified_results_z$n), 3),
    round(abs(overall_slope - sum(stratified_results_z$slope_XY * stratified_results_z$n) / sum(stratified_results_z$n)), 3)
  )
)

# Calculate Simpson's Paradox magnitude measures for C
simpson_magnitude_c <- data.frame(
  Measure = c(
    "Overall Slope",
    "Average C-Stratified Slope",
    "Absolute Difference (C)",
    "Relative Difference (C) (%)",
    "Direction Reversal (C)",
    "Weighted Average Slope (C)",
    "Simpson's Paradox Strength (C)"
  ),
  Value = c(
    round(overall_slope, 3),
    round(mean(stratified_slopes_c), 3),
    round(abs(overall_slope - mean(stratified_slopes_c)), 3),
    round(100 * abs(overall_slope - mean(stratified_slopes_c)) / abs(mean(stratified_slopes_c)), 3),
    ifelse(sign(overall_slope) != sign(mean(stratified_slopes_c)), "Yes", "No"),
    round(sum(stratified_results_c$slope_XY * stratified_results_c$n) / sum(stratified_results_c$n), 3),
    round(abs(overall_slope - sum(stratified_results_c$slope_XY * stratified_results_c$n) / sum(stratified_results_c$n)), 3)
  )
)

# Display magnitude analysis for Z
DT::datatable(
  simpson_magnitude_z,
  caption = "Magnitude of Simpson's Paradox Effect (Z Confounder)",
  options = list(pageLength = 10, dom = 't'),
  colnames = c("Measure", "Value")
)

# Display magnitude analysis for C
DT::datatable(
  simpson_magnitude_c,
  caption = "Magnitude of Simpson's Paradox Effect (C Confounder)",
  options = list(pageLength = 10, dom = 't'),
  colnames = c("Measure", "Value")
)
```

### 22.7 Weighted vs Unweighted Analysis

Let's compare weighted and unweighted analyses to understand the role of group sizes:

```{r}
#| label: weighted-unweighted-analysis-complex
#| message: false
#| warning: false

# Calculate weighted statistics for Z
weighted_slope_z <- sum(stratified_results_z$slope_XY * stratified_results_z$n) / sum(stratified_results_z$n)
weighted_cor_z <- sum(stratified_results_z$cor_XY * stratified_results_z$n) / sum(stratified_results_z$n)

# Calculate unweighted statistics for Z
unweighted_slope_z <- mean(stratified_results_z$slope_XY)
unweighted_cor_z <- mean(stratified_results_z$cor_XY)

# Calculate weighted statistics for C
weighted_slope_c <- sum(stratified_results_c$slope_XY * stratified_results_c$n) / sum(stratified_results_c$n)
weighted_cor_c <- sum(stratified_results_c$cor_XY * stratified_results_c$n) / sum(stratified_results_c$n)

# Calculate unweighted statistics for C
unweighted_slope_c <- mean(stratified_results_c$slope_XY)
unweighted_cor_c <- mean(stratified_results_c$cor_XY)

# Group size analysis for Z
group_size_analysis_z <- stratified_results_z |>
  dplyr::mutate(
    prop_of_total = round(n / sum(n), 3),
    slope_weight = round(slope_XY * prop_of_total, 3),
    cor_weight = round(cor_XY * prop_of_total, 3)
  ) |>
  dplyr::select(Z_quartile, n, prop_of_total, slope_XY, slope_weight, cor_XY, cor_weight)

# Add summary row for Z
summary_row_z <- data.frame(
  Z_quartile = "Weighted Total",
  n = sum(group_size_analysis_z$n),
  prop_of_total = 1.000,
  slope_XY = round(weighted_slope_z, 3),
  slope_weight = round(sum(group_size_analysis_z$slope_weight), 3),
  cor_XY = round(weighted_cor_z, 3),
  cor_weight = round(sum(group_size_analysis_z$cor_weight), 3)
)

group_analysis_complete_z <- dplyr::bind_rows(group_size_analysis_z, summary_row_z)

# Group size analysis for C
group_size_analysis_c <- stratified_results_c |>
  dplyr::mutate(
    prop_of_total = round(n / sum(n), 3),
    slope_weight = round(slope_XY * prop_of_total, 3),
    cor_weight = round(cor_XY * prop_of_total, 3)
  ) |>
  dplyr::select(C_quartile, n, prop_of_total, slope_XY, slope_weight, cor_XY, cor_weight)

# Add summary row for C
summary_row_c <- data.frame(
  C_quartile = "Weighted Total",
  n = sum(group_size_analysis_c$n),
  prop_of_total = 1.000,
  slope_XY = round(weighted_slope_c, 3),
  slope_weight = round(sum(group_size_analysis_c$slope_weight), 3),
  cor_XY = round(weighted_cor_c, 3),
  cor_weight = round(sum(group_size_analysis_c$cor_weight), 3)
)

group_analysis_complete_c <- dplyr::bind_rows(group_size_analysis_c, summary_row_c)

# Display weighted analysis for Z
DT::datatable(
  group_analysis_complete_z,
  caption = "Weighted vs Unweighted Analysis by Group Size (Z Confounder)",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("Z Group", "N", "Proportion", "Slope", "Weighted Slope", 
              "Correlation", "Weighted Correlation")
) |>
  DT::formatRound(
    columns = c("prop_of_total", "slope_XY", "slope_weight", 
                "cor_XY", "cor_weight"), 
    digits = 3
  )

# Display weighted analysis for C
DT::datatable(
  group_analysis_complete_c,
  caption = "Weighted vs Unweighted Analysis by Group Size (C Confounder)",
  options = list(pageLength = 10, scrollX = TRUE),
  colnames = c("C Group", "N", "Proportion", "Slope", "Weighted Slope", 
              "Correlation", "Weighted Correlation")
) |>
  DT::formatRound(
    columns = c("prop_of_total", "slope_XY", "slope_weight", 
                "cor_XY", "cor_weight"), 
    digits = 3
  )

# Summary comparison table
weighting_comparison <- data.frame(
  Analysis_Type = c("Overall (Marginal)", "Unweighted Average (Z)", "Weighted Average (Z)", 
                   "Unweighted Average (C)", "Weighted Average (C)", "True Causal Effect"),
  Slope = c(round(overall_slope, 3), round(unweighted_slope_z, 3), 
           round(weighted_slope_z, 3), round(unweighted_slope_c, 3),
           round(weighted_slope_c, 3), 0.300),
  Correlation = c(round(overall_cor, 3), round(unweighted_cor_z, 3), 
                 round(weighted_cor_z, 3), round(unweighted_cor_c, 3),
                 round(weighted_cor_c, 3), NA)
)

DT::datatable(
  weighting_comparison,
  caption = "Comparison of Different Analysis Approaches",
  options = list(pageLength = 10, dom = 't'),
  colnames = c("Analysis Type", "Slope Estimate", "Correlation")
)
```

### 22.8 Conclusions from Simpson's Paradox Analysis

Based on the comprehensive analysis, here are the summarized findings about Simpson's Paradox in this synthetic dataset:

```{r}
#| label: simpsons-conclusions-complex
#| message: false
#| warning: false

# Calculate key metrics for conclusions
direction_reversal_z <- sign(overall_slope) != sign(mean(stratified_slopes_z))
direction_reversal_c <- sign(overall_slope) != sign(mean(stratified_slopes_c))
magnitude_difference_z <- abs(overall_slope - mean(stratified_slopes_z))
magnitude_difference_c <- abs(overall_slope - mean(stratified_slopes_c))
relative_magnitude_z <- 100 * magnitude_difference_z / abs(mean(stratified_slopes_z))
relative_magnitude_c <- 100 * magnitude_difference_c / abs(mean(stratified_slopes_c))

# Create conclusions summary
conclusions_data <- data.frame(
  Finding = c(
    "Simpson's Paradox Present (Z)?",
    "Simpson's Paradox Present (C)?",
    "Direction of Overall Effect",
    "Direction of Z-Stratified Effects",
    "Direction of C-Stratified Effects", 
    "Magnitude of Difference (Z)",
    "Magnitude of Difference (C)",
    "Relative Magnitude Z (%)",
    "Relative Magnitude C (%)",
    "Primary Explanation",
    "Causal Interpretation",
    "Practical Implication"
  ),
  Result = c(
    simpson_status_z,
    simpson_status_c,
    ifelse(overall_slope > 0, "Positive", "Negative"),
    ifelse(mean(stratified_slopes_z) > 0, "Positive", "Negative"),
    ifelse(mean(stratified_slopes_c) > 0, "Positive", "Negative"),
    paste(round(magnitude_difference_z, 3), "units"),
    paste(round(magnitude_difference_c, 3), "units"),
    paste(round(relative_magnitude_z, 1), "%"),
    paste(round(relative_magnitude_c, 1), "%"),
    "Multiple confounders create different marginal vs conditional relationships",
    "Proper adjustment for all confounders reveals true causal effect",
    "Controlling for all relevant confounders is essential for valid causal inference"
  )
)

# Display conclusions
DT::datatable(
  conclusions_data,
  caption = "Summary of Simpson's Paradox Analysis Conclusions",
  options = list(pageLength = 15, dom = 't', scrollX = TRUE),
  colnames = c("Key Finding", "Result/Interpretation")
)

# Create comprehensive summary table for key conclusions
paradox_present_z <- simpson_status_z != "No Simpson's Paradox"
paradox_present_c <- simpson_status_c != "No Simpson's Paradox"
any_paradox_present <- paradox_present_z || paradox_present_c

key_conclusions <- data.frame(
  Category = c(
    "Simpson's Paradox Detection",
    "Simpson's Paradox Detection", 
    "Simpson's Paradox Detection",
    "Simpson's Paradox Detection",
    "Magnitude Analysis",
    "Magnitude Analysis",
    "Magnitude Analysis",
    "Magnitude Analysis",
    "Causal Inference Implications",
    "Causal Inference Implications",
    "Causal Inference Implications"
  ),
  Finding = c(
    "Z-Stratified Paradox Present?",
    "C-Stratified Paradox Present?",
    "Overall Paradox Status",
    "Mechanistic Explanation",
    "Overall Slope",
    "Average Z-Stratified Slope", 
    "Average C-Stratified Slope",
    "Maximum Relative Difference",
    "Adjustment Set Validation",
    "Multiple Confounder Understanding",
    "Practical Importance"
  ),
  Result = c(
    ifelse(paradox_present_z, "✓ YES - Detected in Z strata", "✗ NO - Not detected in Z strata"),
    ifelse(paradox_present_c, "✓ YES - Detected in C strata", "✗ NO - Not detected in C strata"),
    ifelse(any_paradox_present, "Present when stratifying by individual confounders", 
           "Not present in any stratification"),
    ifelse(
      any_paradox_present, 
      "Individual confounders create paradoxical marginal vs conditional relationships",
      "Relationships remain consistent across confounder strata"
    ),
    round(overall_slope, 3),
    round(mean(stratified_slopes_z), 3),
    round(mean(stratified_slopes_c), 3),
    paste(round(max(relative_magnitude_z, relative_magnitude_c), 1), "%"),
    "Reinforces DAG-based conclusions about need for complete adjustment sets",
    "Validates understanding that partial adjustment can be misleading", 
    "Demonstrates why comprehensive confounder identification is essential"
  )
)

# Display the comprehensive conclusions table
DT::datatable(
  key_conclusions,
  caption = "Key Conclusions from Simpson's Paradox Analysis",
  options = list(
    pageLength = 15,
    ordering = FALSE,
    searching = FALSE,
    scrollX = TRUE
  ),
  class = 'cell-border stripe compact responsive',
  rownames = FALSE,
  width = "90%"
)
```

<br>

The Simpson's Paradox analysis reveals important insights about the nature of confounding in complex causal structures. In this synthetic dataset, we observe **`r simpson_status_z`** when stratifying by Z and **`r simpson_status_c`** when stratifying by C, which demonstrates how individual confounders can create different patterns when we examine marginal versus conditional relationships between X and Y.

The key insight is that Simpson's Paradox in complex DAGs illustrates why piecemeal adjustment strategies can be misleading. While stratifying by individual confounders may reveal paradoxical relationships, the complete picture emerges only when we properly adjust for all relevant confounders simultaneously. In the complex structure, both Z and C act as confounders, and the interaction between these multiple confounding pathways creates the nuanced patterns we observe.

The magnitude of the Simpson's Paradox effects (Z: **`r paste(round(relative_magnitude_z, 1), "%")`** relative difference; C: **`r paste(round(relative_magnitude_c, 1), "%")`** relative difference) underscores the practical importance of comprehensive confounder identification in observational studies. This analysis reinforces the earlier findings about the critical role of using DAG-based methods to identify complete adjustment sets rather than adjusting for confounders one at a time.

This phenomenon demonstrates why randomized controlled trials - which simultaneously balance all confounders across treatment groups - provide such strong evidence for causal inference. In observational studies, we must be equally comprehensive in our approach to confounder control, ensuring we identify and adjust for all relevant variables that create backdoor paths from exposure to outcome.






## Session Information for Reproducibility

```{r}
#| label: session-info
#| echo: false

# Session information for reproducibility
sessionInfo()
```

