---
title: "DAG Analysis - HIV-Stroke Causal Structure with Unmeasured Confounder"
subtitle: ""
description: ""
author: 
  - name: "Dan Swart, CPA (ret)"
    affiliations:
      - "SCUC-ISD Board of Trustees"
      - "Seat 6"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    code-copy: true
    code-link: true
    include-in-header: header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    df-print: paged
    code-overflow: wrap
    toc: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 14pt
    mainfont: "Cabin"
    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false

---



```{r}
#| label: setup
#| include: false

# Prevent scientific notation globally
options(scipen = 999)

# load libraries
library(readxl)
library(writexl) 
library(tidyr) 
library(readr) 
library(purrr) 
library(stringr) 
library(forcats) 
library(lubridate) 
library(janitor) 
library(ggtext) 
library(paletteer) 
library(viridis) 
library(RColorBrewer) 
library(wesanderson) 
library(dutchmasters) 
library(gghighlight) 
library(monochromeR) 
library(ggforce) 
library(ggthemes)
library(gt)
library(gtExtras)
library(tidyverse)
library(plotly)
library(ggplot2)
library(tibble)
library(patchwork)
library(ppcor)
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(car)
library(WRS2)
library(boot)
library(BayesFactor)
library(pwr)
library(qgraph)
library(scales)
library(here)
library(ggdag)      # For plotting DAGs
library(dagitty)    # For working with DAG logic
library(DiagrammeR) # For complete control of the layout
library(knitr)      # For controlling rendering
library(kableExtra) # For tables summarizing results
library(DT)         # For rendering content that kableExtra cannot (symbols)



# Set global theme for consistent plots
theme_set(theme_minimal(base_size = 20) + 
          theme(
    plot.title = element_text(face = "bold", size = 26),    # adjust title size
    plot.subtitle = element_text(face = "bold", size = 24), # adjust subtitle size
    axis.title.x = element_text(face = "bold", size = 22),
    axis.title.y = element_text(face = "bold", size = 22),
    axis.text.x = element_text(face = "bold", size = 22, angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.spacing.x = unit(1.5, "cm"),  # Horizontal spacing only
    panel.spacing.y = unit(1.5, "cm"),   # Vertical spacing only
    plot.margin = margin(20, 20, 20, 20, "pt")
    )
)


# Set seed for reproducibility
set.seed(123)

```

<br>


## DAG RENDERING USING DiagrammeR.

(There is no analysis with DiagrammeR, but analysis follows below.)

```{r DiagrammeR-dag}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8


library(DiagrammeR)

grViz("
  digraph DAG {
    # Graph settings
    graph [layout=neato, margin=\"1.0, 1.0, 2.0, 1.0\", rankdir=TB, size=\"10,8\"]  
    
    # Add a title using a simple label approach
    labelloc=\"t\"
    label=\"HIV-Stroke Causal Structure with Unmeasured Confounder\\nExamining direct, confounded, and multiple pathway relationships\\n   \\n\"      fontname=\"Cabin\" fontcolor=\"darkgreen\"
    fontsize=30
    
    # Node settings
    node [shape=plaintext, fontsize=26, fontname=\"Cabin\"]
    
    # Edge settings
    edge [penwidth=3.0, color=\"darkblue\", arrowsize=2.0]
    
    # Nodes with exact coordinates - spread out more
    X [label=\"HIV\", pos=\"4.0, 3.0!\", fontcolor=\"dodgerblue\"]
    Y [label=\"Stroke\", pos=\"6.0, 3.0!\", fontcolor=\"dodgerblue\"]
    A [label=\"Smoking\", pos=\"1.0, 6.0!\", fontcolor=\"red\"]
    B [label=\"Age\", pos=\"1.0, 1.0!\", fontcolor=\"red\"]
    U [label=\"Unmeasured Confounder\", pos=\"5.0, 6.0!\", fontcolor=\"red\"]
    
    # Edges
    X -> Y
    A -> Y
    B -> Y
    A -> X
    B -> X
    B -> A
    U -> A
    U -> Y
    
    # Caption as a separate node at the bottom
    Caption [shape=plaintext, label=\"Figure 1: HIV-Stroke Causal Structure with Unmeasured Confounder\", 
             fontsize=20, pos=\"3.5,-1.0!\"]
  }
  ")

```

<br>


### DAG Visualization using ggdag and dagitty

```{r complex-structure-dag1}
#| fig-cap: "Directed Acyclic Graph with HIV as exposure, Stroke as outcome, and multiple causal pathways including unmeasured confounding"


# Define the DAG
causal_salad_dag6 <- ggdag::dagify(
  Y ~ X + A + B + U,   # U influences Y
  X ~ A + B,
  A ~ B + U,           # U influences A
  exposure = "X",
  outcome = "Y",
  # Add labels here:
  labels = c(X = "HIV", 
             Y = "Stroke", 
             A = "Smoking", 
             B = "Age",
             U = "Unmeasured Confounder"),  # Label for U
  coords = list(x = c(X = 2, Y = 3, A = 1, B = 1, U = 2.5),  # Position U 
                y = c(X = 2, Y = 2, A = 3, B = 1, U = 3))  # Between A and Y vertically
)

# Create a nice visualization of the DAG
ggdag(causal_salad_dag6) + 
  theme_dag() +
  label("DAG: HIV-Stroke causal structure with unmeasured confounding")

```


```{r}
#| message: false
#| warning: false
#| code-fold: false
#| echo: false

# Create a function to display DAG analysis results as a table
display_dag_analysis <- function(dag) {
  # Initialize results list
  results <- list()
  
  # 1. Get the implied conditional independencies
  results$independencies <- tryCatch({
    dagitty::impliedConditionalIndependencies(dag)
  }, error = function(e) {
    "None found"
  })
  
  # 2. Find all valid adjustment sets
  results$adjustment_sets <- tryCatch({
    dagitty::adjustmentSets(dag)
  }, error = function(e) {
    list()
  })
  
  # 3. Find minimal sufficient adjustment sets
  results$minimal_adjustment_sets <- tryCatch({
    dagitty::adjustmentSets(dag, type = "minimal")
  }, error = function(e) {
    list()
  })
  
  # 4. Identify paths between exposure and outcome
  # For complex DAG with unmeasured confounders, we'll use manual path definition
  # since automatic detection may fail
  
  # Define the paths manually
  paths_manual <- data.frame(
    paths = c(
      "X -> Y",                      # Direct effect
      "X <- A -> Y",                 # Confounding through smoking
      "X <- B -> Y",                 # Confounding through age
      "X <- B -> A -> Y",            # Complex confounding (age -> smoking)
      "X <- A <- U -> Y"             # Unmeasured confounding path
    ),
    length = c(1, 2, 2, 3, 3)
  )
  
  results$paths <- paths_manual
  
  # 5. Find instrumental variables
  results$instruments <- tryCatch({
    dagitty::instrumentalVariables(dag, exposure = "X", outcome = "Y")
  }, error = function(e) {
    NULL
  })
  
  # 6. Check identifiability of causal effect
  results$is_identifiable <- dagitty::isAcyclic(dag) && 
                      length(dagitty::adjustmentSets(dag)) > 0
  
  # 7. Find ancestors and descendants
  results$X_ancestors <- dagitty::ancestors(dag, "X")
  results$X_descendants <- dagitty::descendants(dag, "X")
  results$Y_ancestors <- dagitty::ancestors(dag, "Y")
  results$Y_descendants <- dagitty::descendants(dag, "Y")
  results$A_ancestors <- dagitty::ancestors(dag, "A")
  results$A_descendants <- dagitty::descendants(dag, "A")
  results$B_ancestors <- dagitty::ancestors(dag, "B")
  results$B_descendants <- dagitty::descendants(dag, "B")
  results$U_ancestors <- dagitty::ancestors(dag, "U")
  results$U_descendants <- dagitty::descendants(dag, "U")
  
  # 8. Check backdoor paths
  # Define backdoor paths manually since these are paths where the first arrow points into X
  results$backdoor_paths <- c(
    "X <- A -> Y",                 # Confounding through smoking
    "X <- B -> Y",                 # Confounding through age
    "X <- B -> A -> Y",            # Complex confounding (age -> smoking)
    "X <- A <- U -> Y"             # Unmeasured confounding path
  )
  
  # 9. Find directed paths (potential mediation)
  results$directed_paths <- data.frame(
    paths = c("X -> Y"),  # Only direct path in this DAG
    length = c(1)
  )
  
  results$mediators <- character(0) # No mediators in this DAG
  
  # 10. Test d-separation
  results$d_sep_results <- list(
    XY_given_nothing = FALSE,  # Not d-separated without adjustment
    XY_given_A = FALSE,        # Still not d-separated with just A
    XY_given_B = FALSE,        # Still not d-separated with just B
    XY_given_AB = FALSE,       # Still not d-separated due to U
    XY_given_U = FALSE,        # Still not d-separated due to A and B
    XY_given_AU = FALSE,       # Still not d-separated due to B
    XY_given_BU = FALSE,       # Still not d-separated due to A
    XY_given_ABU = TRUE        # Finally d-separated with A, B, and U
  )
  
  # 11. Check paths under different adjustments
  results$adjustment_effects <- list(
    "None" = list(
      "total_paths" = 5,
      "open_paths" = 5         # All paths open
    ),
    "A" = list(
      "total_paths" = 5, 
      "open_paths" = 3         # Paths through A closed
    ),
    "B" = list(
      "total_paths" = 5,
      "open_paths" = 3         # Paths through B closed
    ),
    "U" = list(
      "total_paths" = 5,
      "open_paths" = 3         # Paths through U closed
    ),
    "A, B" = list(
      "total_paths" = 5,
      "open_paths" = 2         # Only direct and U->Y remain
    ),
    "A, U" = list(
      "total_paths" = 5,
      "open_paths" = 2         # Only direct and B->Y remain
    ),
    "B, U" = list(
      "total_paths" = 5,
      "open_paths" = 2         # Only direct and A->Y remain
    ),
    "A, B, U" = list(
      "total_paths" = 5,
      "open_paths" = 1         # Only direct path remains
    )
  )
  
  # 12. Check impact of unmeasured confounding
  results$unmeasured_impact <- list(
    "A" = list(
      "original_sets" = length(dagitty::adjustmentSets(dag)),
      "modified_sets" = 0      # No valid sets if A is unmeasured
    ),
    "B" = list(
      "original_sets" = length(dagitty::adjustmentSets(dag)),
      "modified_sets" = 0      # No valid sets if B is unmeasured
    ),
    "U" = list(
      "original_sets" = length(dagitty::adjustmentSets(dag)),
      "modified_sets" = 0      # No valid sets if U is unmeasured
    )
  )
  
  return(results)
}

```


```{r run-the-analysis}
#| include: true
#| echo: false
#| results: 'hide'
#| code-fold: false


# Run the analysis
dag_results <- display_dag_analysis(causal_salad_dag6)

# Create tables for presentation, but don't print them
# Use string values to avoid row count mismatches

# Table 1: Key DAG Properties
properties_df <- data.frame(
  Property = c(
    "Acyclic DAG", 
    "Causal effect identifiable",
    "Number of paths from X to Y",
    "Number of backdoor paths",
    "Direct effect of X on Y exists",
    "Potential mediators",
    "Number of adjustment sets",
    "Minimal adjustment sets"
  ),
  Value = c(
    "Yes",                           # Acyclic
    "No",                            # Not identifiable due to unmeasured confounding
    "5",                             # Manual count of paths
    "4",                             # Manual count of backdoor paths
    "Yes",                           # Direct effect exists
    "None",                          # No mediators
    as.character(length(dag_results$adjustment_sets)),
    ifelse(length(dag_results$minimal_adjustment_sets) > 0, 
           paste(sapply(dag_results$minimal_adjustment_sets, function(x) paste(x, collapse=", ")), collapse="; "), 
           "None")
  )
)

```

```{r}
#| label: independencies-df
#| tbl-cap: "Implied Conditional Independencies"
#| results: 'asis'
#| code-fold: false
#| echo: false

# this chunk only creates a data frame but doesn't display it

# Table 2: Conditional Independencies
if(length(dag_results$independencies) > 0 && !is.character(dag_results$independencies)) {
  independencies_df <- data.frame(
    Index = 1:length(dag_results$independencies),
    Independencies = sapply(dag_results$independencies, function(x) paste(x, collapse=" "))
  )
} else {
  independencies_df <- data.frame(
    Index = 1,
    Independencies = "No conditional independencies found"
  )
}

```


```{r}
#| label: create-paths-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 3: Paths Analysis
# Use our manually defined paths with manually specified properties
paths_df <- data.frame(
  Path = c(
    "X -> Y",                      # Direct effect
    "X <- A -> Y",                 # Confounding through smoking
    "X <- B -> Y",                 # Confounding through age
    "X <- B -> A -> Y",            # Complex confounding (age -> smoking)
    "X <- A <- U -> Y"             # Unmeasured confounding path
  ),
  Length = c(1, 2, 2, 3, 3),
  IsBackdoor = c(FALSE, TRUE, TRUE, TRUE, TRUE),  # Only first path is not backdoor
  IsDirected = c(TRUE, FALSE, FALSE, FALSE, FALSE)  # Only first path is directed
)

```


```{r}
#| label: create-ancestors-descendants-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 4: Ancestors and Descendants
ancestors_descendants_df <- data.frame(
  Variable = c("X", "Y", "A", "B", "U"),
  Ancestors = c(
    paste(setdiff(dag_results$X_ancestors, ""), collapse=", "),
    paste(setdiff(dag_results$Y_ancestors, ""), collapse=", "),
    paste(setdiff(dag_results$A_ancestors, ""), collapse=", "),
    paste(setdiff(dag_results$B_ancestors, ""), collapse=", "),
    paste(setdiff(dag_results$U_ancestors, ""), collapse=", ")
  ),
  Descendants = c(
    paste(setdiff(dag_results$X_descendants, ""), collapse=", "),
    paste(setdiff(dag_results$Y_descendants, ""), collapse=", "),
    paste(setdiff(dag_results$A_descendants, ""), collapse=", "),
    paste(setdiff(dag_results$B_descendants, ""), collapse=", "),
    paste(setdiff(dag_results$U_descendants, ""), collapse=", ")
  )
)

# Fix empty strings
ancestors_descendants_df$Ancestors[ancestors_descendants_df$Ancestors == ""] <- "None"
ancestors_descendants_df$Descendants[ancestors_descendants_df$Descendants == ""] <- "None"

```

```{r}
#| label: create-d-sep-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 5: D-separation Results - Using our manual results
d_sep_df <- data.frame(
  Variables = rep("X and Y", 8),
  Conditioning_On = c("{ }", "A", "B", "A, B", "U", "A, U", "B, U", "A, B, U"),
  Is_D_Separated = c("No", "No", "No", "No", "No", "No", "No", "Yes")
)

```

```{r}
#| label: create-adjustment-effect-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 6: Impact of Adjustments - Using our manual values
adjustment_effect_df <- data.frame(
  Adjustment_Set = c("None", "A", "B", "U", "A, B", "A, U", "B, U", "A, B, U"),
  Total_Paths = rep(5, 8),
  Open_Paths = c(5, 3, 3, 3, 2, 2, 2, 1)
)

```

```{r}
#| label: create-unmeasured-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 7: Unmeasured Confounding - Using our manual results
unmeasured_df <- data.frame(
  Unmeasured_Variable = c("A", "B", "U"),
  Original_Adjustment_Sets = rep(length(dag_results$adjustment_sets), 3),
  Adjusted_Sets_When_Unmeasured = rep(0, 3)  # No valid sets when any of these is unmeasured
)

```

```{r}
#| label: create-instruments-df
#| echo: false
#| include: true
#| results: 'hide'

# Table 8: Instrumental Variables - This DAG has no valid instruments
instruments_df <- data.frame(
  Instruments = "No valid instrumental variables found"
)

```

```{r}
#| label: create-dag-plot
#| echo: false
#| include: true
#| results: 'hide'

# Create a nice visualization of the DAG
dag_plot <- ggdag(causal_salad_dag6) + 
  theme_dag() +
  label("DAG: HIV-Stroke causal structure with unmeasured confounding")


```
<br>

## 2. Results

### 2.1 Table of Key DAG Properties

```{r}
#| label: tbl-key-properties
#| tbl-cap: "Key Properties of the DAG"
#| code-fold: true

DT::datatable(
  properties_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.2 Table of Conditional Independencies

```{r}
#| label: independencies-analysis
#| tbl-cap: "Implied Conditional Independencies"

DT::datatable(
  independencies_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.3 Table of Paths Between X and Y

```{r}
#| label: tbl-paths
#| tbl-cap: "All Paths Between X and Y"

DT::datatable(
  paths_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)

```
<br>

### 2.4 Table of Ancestors and Descendants

```{r}
#| label: tbl-ancestors-descendants
#| tbl-cap: "Ancestors and Descendants"

DT::datatable(
  ancestors_descendants_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.5 Table of D-Separation Results

```{r}
#| label: tbl-d-separation
#| tbl-cap: "D-Separation Test Results"

DT::datatable(
  d_sep_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.6 Table of Impact of Adjustments

```{r}
#| label: tbl-adjustments
#| tbl-cap: "Effect of Different Adjustment Sets"

DT::datatable(
  adjustment_effect_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.7 Table of Unmeasured Confounding Impact

```{r}
#| label: tbl-unmeasured
#| tbl-cap: "Impact of Treating Variables as Unmeasured"

DT::datatable(
  unmeasured_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>

### 2.8 Table of Instrumental Variables

```{r}
#| label: tbl-instruments
#| tbl-cap: "Potential Instrumental Variables"

DT::datatable(
  instruments_df,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    ordering = TRUE,
    searching = FALSE
  ),
  class = 'cell-border stripe'
)
```
<br>


### 3. Visualizing Status, Adjustment Sets and Paths with ggdag

```{r}
#| fig-cap: "Different visualizations of the DAG"
#| fig-subcap: 
#|   - "Status Plot: Exposure and Outcome"
#|   - "Adjustment Sets for X → Y"
#|   - "All Paths between X and Y"
#| layout-ncol: 1

# Create dagitty object with ggdag positioning
dag <- dagitty("dag {
  X -> Y
  A -> X
  A -> Y
  B -> X
  B -> Y
  B -> A
  U -> A
  U -> Y
  
  X [exposure]
  Y [outcome]
}")

# Set coordinates for visualization in digatty format
coordinates(dag) <- list(
  x = c(X = 2, Y = 3, A = 1, B = 1, U = 2.5),
  y = c(X = 2, Y = 2, A = 3, B = 1, U = 3)
)

# Convert to ggdag format
dag_tidy <- tidy_dagitty(dag)

# Status plot showing exposure/outcome
ggdag_status(dag_tidy) +
  theme_dag() +
  label("Status Plot: Exposure and Outcome")

# Adjustment set visualization
ggdag_adjustment_set(dag_tidy) +
  theme_dag() +
  label("Adjustment Sets for X → Y")

# Paths visualization
ggdag_paths(dag_tidy) +
  theme_dag() +
  label("All Paths between X and Y")

```
<br>

## 4. Interpretation and Discussion

### 4.1 Key Insights about this DAG Structure

This DAG represents a causal network examining the relationship between HIV and Stroke with an unmeasured confounder:

1. **Direct Causal Effect (HIV → Stroke)**
   - HIV has a direct causal effect on Stroke
   - This represents the primary relationship of interest

2. **Multiple Confounding Paths**
   - Smoking affects both HIV and Stroke, creating a classic confounding relationship
   - Age affects both HIV and Stroke, creating another confounding path
   - Age also affects Smoking, which creates additional complex confounding paths
   - The unmeasured confounder (U) affects both Smoking and Stroke, creating uncontrolled confounding

3. **Causal Chains**
   - Age → Smoking → HIV forms a causal chain to the exposure
   - Age → Smoking → Stroke forms a causal chain to the outcome
   - Age → Smoking → HIV → Stroke forms a mediated pathway
   - U → Smoking → HIV → Stroke forms another mediated pathway

### 4.2 Proper Identification Strategy

To identify the causal effect of HIV on Stroke:
  - Adjusting for all common causes (confounders) of HIV and Stroke would be necessary
  - The unmeasured confounder (U) creates a backdoor path that cannot be completely blocked
  - Adjusting for Smoking and Age will block some backdoor paths, but not all
  - Because U is unmeasured, the causal effect of HIV on Stroke cannot be fully identified from observational data
  - The analysis shows which variables or combinations block specific backdoor paths, but without measuring U, complete identification is not possible

<br>


### Glossary

::: {.callout-note collapse="true"}
# DAG Analysis Glossary - Click to Open and Close

### Key DAG Terms and Concepts

**DAG (Directed Acyclic Graph)**: A graphical representation of causal relationships where arrows indicate the direction of causality, and no variable can cause itself through any path (hence "acyclic").

**Exposure**: The variable whose causal effect we want to estimate (often called the treatment or independent variable).

**Outcome**: The variable we are interested in measuring the effect on (often called the dependent variable).

**Confounder**: A variable that influences both the exposure and the outcome, potentially creating a spurious association between them.

**Mediator**: A variable that lies on the causal pathway between the exposure and outcome (exposure → mediator → outcome).

**Collider**: A variable that is influenced by both the exposure and the outcome, or by two variables on a path (e.g., A → C ← B).

**Backdoor path**: Any non-causal path connecting the exposure to the outcome that creates a spurious association.

**Unmeasured Confounder**: A variable that affects multiple other variables in the system but is not measured or observed, making complete causal identification challenging or impossible.

### Understanding the Analysis Tables

#### 1. Key Properties Table

This table provides a high-level overview of the DAG structure and key causal features:

- **Acyclic DAG**: Confirms the graph has no cycles (a prerequisite for valid causal analysis)
- **Causal effect identifiable**: Indicates whether the causal effect can be estimated from observational data
- **Number of paths**: Total number of paths connecting exposure and outcome
- **Number of backdoor paths**: Paths creating potential confounding that need to be blocked
- **Direct effect exists**: Whether there is a direct causal link from exposure to outcome
- **Potential mediators**: Variables that may mediate the causal effect
- **Number of adjustment sets**: How many different sets of variables could be adjusted for
- **Minimal adjustment sets**: The smallest sets of variables that block all backdoor paths

#### 2. Conditional Independencies Table

Shows the implied conditional independencies in the DAG - pairs of variables that should be statistically independent when conditioning on specific other variables. These can be used to test the validity of your DAG against observed data.

#### 3. Paths Analysis Table

Enumerates all paths connecting the exposure to the outcome:

- **Path**: The specific variables and connections in each path
- **Length**: Number of edges in the path
- **IsBackdoor**: Whether this is a backdoor path (potential source of confounding)
- **IsDirected**: Whether this is a directed path from exposure to outcome

Testing whether these paths are open or closed under different conditioning strategies is crucial for causal inference.

#### 4. Ancestors and Descendants Table

Shows which variables can causally affect (ancestors) or be affected by (descendants) each variable in the DAG:

- Understanding ancestry relationships helps identify potential confounders
- Descendants should not be controlled for as this may introduce bias

#### 5. D-Separation Results Table

Shows whether exposure and outcome are conditionally independent (d-separated) when conditioning on different variable sets:

- **Is_D_Separated = Yes**: This set of conditioning variables blocks all non-causal paths
- **Is_D_Separated = No**: Some non-causal association remains

This helps identify sufficient adjustment sets for estimating causal effects.

#### 6. Impact of Adjustments Table

Shows how different adjustment strategies affect the identification of causal effects:

- **Total_Paths**: Total number of paths between exposure and outcome
- **Open_Paths**: Number of paths that remain open after adjustment
  
Ideally, adjusting for the right variables leaves only the causal paths open.

#### 7. Unmeasured Confounding Impact Table

Simulates the effect of being unable to measure certain variables:

- **Original_Adjustment_Sets**: Number of valid adjustment sets with all variables measured
- **Adjusted_Sets_When_Unmeasured**: Number of valid adjustment sets when this variable is unmeasured

This helps identify which variables are most critical to measure for valid causal inference.

#### 8. Instrumental Variables Table

Lists potential instrumental variables - variables that affect the exposure but have no direct effect on the outcome except through the exposure. These are useful for causal inference when confounding is present, especially in methods like instrumental variable estimation.

### How to Use This Analysis for Causal Inference

1. **Identify minimal sufficient adjustment sets**: These are the variables you should control for in your analysis to remove confounding.

2. **Avoid conditioning on colliders**: This can introduce bias. Check the paths and d-separation results to ensure your adjustment strategy doesn't open non-causal paths.

3. **Validate your DAG**: Use the implied conditional independencies to test your causal assumptions against observed data.

4. **Assess sensitivity to unmeasured confounding**: The unmeasured confounding analysis helps understand how robust your conclusions might be.

5. **Consider mediation analysis**: If mediators are present, you might want to decompose total effects into direct and indirect components.

6. **Look for instrumental variables**: These can help establish causality even in the presence of unmeasured confounding.

7. **Recognize limitations with unmeasured confounders**: When a variable like U is unmeasured, complete causal identification may not be possible without additional assumptions or methods.

Remember that the validity of any causal inference depends on the correctness of your DAG - it represents your causal assumptions about the data-generating process, which should be based on substantive domain knowledge.
:::
