---
title: "DAG Analysis: HIV-Stroke Causal Structure with Unmeasured Confounder"
subtitle: ""
description: ""
author: 
  - name: "Dan Swart, CPA (ret)"
    affiliations:
      - "SCUC-ISD Board of Trustees"
      - "Seat 6"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true
    include-in-header: header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 14pt
    mainfont: "Cabin"
    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

---



```{r}
#| label: setup
#| include: false

# Prevent scientific notation globally
options(scipen = 999)

# load libraries
library(readxl)
library(writexl) 
library(tidyr) 
library(readr) 
library(purrr) 
library(stringr) 
library(forcats) 
library(lubridate) 
library(janitor) 
library(ggtext) 
library(paletteer) 
library(viridis) 
library(RColorBrewer) 
library(wesanderson) 
library(dutchmasters) 
library(gghighlight) 
library(monochromeR) 
library(ggforce) 
library(ggthemes)
library(gt)
library(gtExtras)
library(tidyverse)
library(plotly)
library(ggplot2)
library(tibble)
library(patchwork)
library(ppcor)
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(car)
library(WRS2)
library(boot)
library(BayesFactor)
library(pwr)
library(qgraph)
library(scales)
library(here)
library(ggdag)      # For plotting DAGs
library(dagitty)    # For working with DAG logic
library(DiagrammeR) # For complete control of the layout
library(knitr)      # For controlling rendering
library(kableExtra) # For tables summarizing results
library(DT)         # For rendering content that kableExtra cannot (symbols)
library(lavaan)     # For latent variable analysis
library(ggpubr)
library(rethinking)
library(broom)


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Set global theme for consistent plots
theme_set(theme_minimal(base_size = 20) + 
          theme(
    plot.title = element_text(face = "bold", size = 26),    # adjust title size
    plot.subtitle = element_text(face = "bold", size = 24), # adjust subtitle size
    axis.title.x = element_text(face = "bold", size = 22),
    axis.title.y = element_text(face = "bold", size = 22),
    axis.text.x = element_text(face = "bold", size = 22, angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.spacing.x = unit(1.5, "cm"),  # Horizontal spacing only
    panel.spacing.y = unit(1.5, "cm"),   # Vertical spacing only
    plot.margin = margin(20, 20, 20, 20, "pt")
    )
)


# Set seed for reproducibility
set.seed(123)

```

## Introduction: Understanding HIV-Stroke Causal Structure with Unmeasured Confounding

This document explores a complex causal structure examining the relationship between HIV status and stroke risk in the presence of an unmeasured confounder. This represents one of the most challenging scenarios in observational epidemiology where:

- X (HIV) is the exposure variable of primary interest
- Y (Stroke) is the outcome variable  
- A (Smoking) and B (Age) are measured confounders that we can observe and adjust for
- U (Unmeasured Confounder) represents unknown factors that affect both smoking behavior and stroke risk

This DAG structure illustrates a fundamental limitation of observational studies: even with careful measurement and adjustment for all known confounders, unmeasured confounding can still bias causal estimates. Think of it like trying to navigate with a map that's missing some important landmarks - you can see the roads you know about (measured confounders), but there are invisible forces (unmeasured confounders) affecting your journey that you can't account for.

The presence of unmeasured confounding creates backdoor paths between exposure and outcome that cannot be closed through standard statistical adjustment, making true causal identification impossible without strong assumptions.

## 2. Describe the DAG in words

This DAG represents a causal structure with five variables where one critical confounder remains unmeasured:

**Variables:**
- X (HIV): The primary exposure of interest
- Y (Stroke): The health outcome we want to understand
- A (Smoking): A measured lifestyle factor that affects both HIV acquisition risk and stroke risk
- B (Age): A measured demographic factor that influences smoking, HIV risk, and stroke risk  
- U (Unmeasured Confounder): An unobserved variable that affects both smoking behavior and stroke risk

**The causal pathways include:**

*Direct causal relationships:*
1. X → Y: The direct effect of HIV on stroke risk (our parameter of interest)
2. A → Y: Direct effect of smoking on stroke risk
3. B → Y: Direct effect of age on stroke risk

*Measured confounding pathways:*
1. X ← A → Y: Confounding through smoking behavior
2. X ← B → Y: Confounding through age
3. X ← B → A → Y: Indirect confounding pathway through age affecting smoking

*Unmeasured confounding pathways:*
1. A ← U → Y: The unmeasured confounder affects both smoking and stroke
2. X ← A ← U → Y: Complex pathway where the unmeasured confounder influences HIV status through smoking, then affects stroke both directly and through HIV

This creates a challenging analytical situation where standard epidemiological methods can adjust for the measured confounders (Age and Smoking) but cannot eliminate bias from the unmeasured confounder U. The result is that even our best available statistical models will provide biased estimates of the true HIV effect on stroke risk.

## 3. Recreate the DAG for reference using DiagrammeR and ggdag

```{r}
#| label: diagrammer-dag-unmeasured
#| fig-cap: "HIV-Stroke Causal DAG with Unmeasured Confounding"

library(DiagrammeR)

# Create the DAG using DiagrammeR with exact template formatting
hiv_stroke_dag <- grViz("
  digraph DAG {
    # Graph settings
    graph [layout=neato, margin=\"1.0, 1.0, 2.0, 1.0\"]  
    
    # Add a title
    labelloc=\"t\"
    label=\"HIV-Stroke Causal Structure with Unmeasured Confounder\\nDemonstrating the Challenge of Hidden Bias\\n   \"
    fontname=\"Cabin\"  fontsize=18
    
   # Node settings
    node [shape=plaintext, fontsize=20, fontname=\"Cabin\"]
    
    # Edge settings
    edge [penwidth=1.20, color=\"darkblue\", arrowsize=1.00, fontsize=12]
    
    # Nodes with exact coordinates
    X [label=\"HIV\", pos=\"4.0, 3.0!\", fontcolor=\"dodgerblue\"]
    Y [label=\"Stroke\", pos=\"6.0, 3.0!\", fontcolor=\"dodgerblue\"]
    A [label=\"Smoking\", pos=\"1.0, 6.0!\", fontcolor=\"red\"]
    B [label=\"Age\", pos=\"1.0, 1.0!\", fontcolor=\"red\"]
    U [label=\"Unmeasured Confounder\", pos=\"5.0, 6.0!\", fontcolor=\"red\"]
    
    # Edges with true effect sizes
    X -> Y [label=\"0.5\"]
    A -> Y [label=\"0.7\"]
    B -> Y [label=\"0.3\"]
    A -> X [label=\"0.4\"]
    B -> X [label=\"0.2\"]
    B -> A [label=\"0.3\"]
    U -> A [label=\"0.6\"]
    U -> Y [label=\"0.4\"]
    
    # Caption
    Caption [shape=plaintext, label=\"Figure 1: HIV-Stroke Causal Structure with Unmeasured Confounding\", 
             fontsize=12, pos=\"3.5,0.0!\"]
  }
")

# Display the DAG
hiv_stroke_dag
```

```{r}
#| label: ggdag-visualization-hiv-stroke
#| fig-cap: "ggdag representation of HIV-Stroke causal model with unmeasured confounding"

# Define the DAG using dagitty/ggdag for analysis
hiv_stroke_ggdag <- dagify(
  Y ~ X + A + B + U,   # Stroke affected by HIV, Smoking, Age, and Unmeasured
  X ~ A + B,           # HIV affected by Smoking and Age
  A ~ B + U,           # Smoking affected by Age and Unmeasured
  exposure = "X",
  outcome = "Y",
  labels = c("X" = "HIV", 
             "Y" = "Stroke", 
             "A" = "Smoking",
             "B" = "Age",
             "U" = "Unmeasured")
)

# Set coordinates for visualization
coordinates(hiv_stroke_ggdag) <- list(
  x = c(X = 2.5, Y = 4, A = 1, B = 1, U = 3.5),
  y = c(X = 2, Y = 2, A = 4, B = 1, U = 4)
)

# Create visualization
ggdag(hiv_stroke_ggdag, edge_type = "link") + 
  geom_dag_point(color = "lightblue", size = 14, alpha = 0.7) +
  geom_dag_text(color = "black") +
  geom_dag_edges(edge_colour = "blue", edge_width = 1.0, arrow_size = 0.6) +
  theme_dag() +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("DAG: HIV-Stroke Structure with Unmeasured Confounding")
```

## 4. Generate synthetic data following the causal structure

We'll generate synthetic data that follows the causal relationships specified in our DAG. The key challenge here is that we must generate the unmeasured confounder to create realistic bias, but then "hide" it from our analysis to simulate real-world conditions where this confounder is unknown to researchers.

```{r}
#| label: generate-data-hiv-stroke-unmeasured
#| tbl-cap: "Summary of synthetic HIV-stroke data with unmeasured confounding"

# Set seed for reproducibility
set.seed(42)

# Sample size
n <- 1000

# Generate data following the DAG structure
# Start with root causes: B (Age) and U (Unmeasured Confounder)
B_age <- round(rnorm(n, mean = 0, sd = 1), 3)
U_unmeasured <- round(rnorm(n, mean = 0, sd = 1), 3)

# Generate A (Smoking) influenced by B (Age) and U (Unmeasured)
A_smoking <- round(0.3 * B_age + 0.6 * U_unmeasured + rnorm(n, mean = 0, sd = 0.8), 3)

# Generate X (HIV) influenced by A (Smoking) and B (Age)
X_hiv <- round(0.4 * A_smoking + 0.2 * B_age + rnorm(n, mean = 0, sd = 0.7), 3)

# Generate Y (Stroke) influenced by X (HIV), A (Smoking), B (Age), and U (Unmeasured)
Y_stroke <- round(0.5 * X_hiv + 0.7 * A_smoking + 0.3 * B_age + 0.4 * U_unmeasured + rnorm(n, mean = 0, sd = 0.6), 3)

# True direct effect of HIV on Stroke
true_hiv_effect <- 0.500

# Create observed dataset (what researchers would actually see)
observed_hiv_data <- data.frame(
  Age = B_age,
  Smoking = A_smoking, 
  HIV = X_hiv,
  Stroke = Y_stroke
)

# Create complete dataset including unmeasured confounder (for oracle comparison)
complete_hiv_data <- data.frame(
  Age = B_age,
  Smoking = A_smoking, 
  HIV = X_hiv,
  Stroke = Y_stroke,
  Unmeasured = U_unmeasured
)

# Display summary statistics for observed data
round(sapply(observed_hiv_data, summary), 3)
```

```{r}
#| label: true-effects-hiv-stroke
#| tbl-cap: "True causal effects in the HIV-Stroke DAG with unmeasured confounding"

# Create table of all true causal effects
true_effects_hiv <- data.frame(
  Causal_Relationship = c("Age → Smoking", "Age → HIV", "Age → Stroke", 
                         "Smoking → HIV", "Smoking → Stroke", "HIV → Stroke",
                         "Unmeasured → Smoking", "Unmeasured → Stroke"),
  True_Effect = c(0.3, 0.2, 0.3, 0.4, 0.7, 0.5, 0.6, 0.4),
  Effect_Type = c("Root cause to mediator", "Confounder to exposure", "Confounder to outcome",
                 "Mediator to exposure", "Confounder to outcome", "CAUSAL EFFECT OF INTEREST",
                 "Unmeasured to mediator", "Unmeasured to outcome"),
  Bias_Source = c("No", "No", "No", "No", "No", "Target Effect",
                 "Yes - Creates bias", "Yes - Creates bias")
)

# Display the table
datatable(true_effects_hiv,
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## 5. Examine structure of synthetic data

### 5.1 Correlation matrix of observed variables

```{r}
#| label: correlation-matrix-hiv-stroke
#| fig-cap: "Correlation matrix of observed variables in HIV-stroke data"

# Calculate correlation matrix for observed variables
corr_matrix_hiv <- cor(observed_hiv_data)

# Create correlation table
corr_table_hiv <- as.data.frame(round(corr_matrix_hiv, 3))

# Display correlation table
datatable(corr_table_hiv,
          options = list(pageLength = 10, dom = 't'),
          rownames = TRUE,
          class = 'cell-border stripe compact responsive')

# Create correlation plot
corrplot(corr_matrix_hiv, 
         method = "color", 
         type = "upper", 
         order = "hclust",
         addCoef.col = "black",
         number.cex = 1.5,
         tl.col = "black",
         tl.srt = 45,
         diag = FALSE,
         col = colorRampPalette(c("#6BAED6", "white", "#E6550D"))(200),
         title = "Correlation Matrix: Observed Variables",
         mar = c(0,0,1,0))

# Interpretation of key correlations
correlation_interpretation <- data.frame(
  Variable_Pair = c("HIV and Stroke", "HIV and Smoking", "HIV and Age", 
                   "Stroke and Smoking", "Stroke and Age", "Smoking and Age"),
  Observed_Correlation = c(round(cor(observed_hiv_data$HIV, observed_hiv_data$Stroke), 3), 
                          round(cor(observed_hiv_data$HIV, observed_hiv_data$Smoking), 3), 
                          round(cor(observed_hiv_data$HIV, observed_hiv_data$Age), 3),
                          round(cor(observed_hiv_data$Stroke, observed_hiv_data$Smoking), 3), 
                          round(cor(observed_hiv_data$Stroke, observed_hiv_data$Age), 3), 
                          round(cor(observed_hiv_data$Smoking, observed_hiv_data$Age), 3)),
  Interpretation = c(
    "Strong correlation: Direct effect + measured confounding + unmeasured confounding",
    "Moderate correlation: Smoking increases HIV risk",
    "Weak correlation: Age increases HIV risk",
    "Strong correlation: Direct smoking effect + shared unmeasured confounding",
    "Moderate correlation: Age affects stroke + indirect pathways",
    "Moderate correlation: Age affects smoking + unmeasured confounding"
  )
)

datatable(correlation_interpretation,
          caption = "Interpretation of key correlations (unmeasured confounding hidden)",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

### 5.2 Impact of unmeasured confounder on correlations

```{r}
#| label: unmeasured-impact-correlations
#| fig-cap: "Comparing correlations with and without unmeasured confounder visibility"

# Calculate correlations including unmeasured confounder
corr_matrix_complete <- cor(complete_hiv_data)

# Create comparison of key correlations
correlation_comparison <- data.frame(
  Variable_Pair = c("HIV-Stroke", "HIV-Smoking", "Stroke-Smoking", "HIV-Age", "Stroke-Age", "Smoking-Age"),
  Observed_Data = c(
    round(cor(observed_hiv_data$HIV, observed_hiv_data$Stroke), 3),
    round(cor(observed_hiv_data$HIV, observed_hiv_data$Smoking), 3),
    round(cor(observed_hiv_data$Stroke, observed_hiv_data$Smoking), 3),
    round(cor(observed_hiv_data$HIV, observed_hiv_data$Age), 3),
    round(cor(observed_hiv_data$Stroke, observed_hiv_data$Age), 3),
    round(cor(observed_hiv_data$Smoking, observed_hiv_data$Age), 3)
  ),
  With_Unmeasured_Visible = c(
    round(cor(complete_hiv_data$HIV, complete_hiv_data$Stroke), 3),
    round(cor(complete_hiv_data$HIV, complete_hiv_data$Smoking), 3),
    round(cor(complete_hiv_data$Stroke, complete_hiv_data$Smoking), 3),
    round(cor(complete_hiv_data$HIV, complete_hiv_data$Age), 3),
    round(cor(complete_hiv_data$Stroke, complete_hiv_data$Age), 3),
    round(cor(complete_hiv_data$Smoking, complete_hiv_data$Age), 3)
  ),
  Unmeasured_Correlations = c(
    round(cor(complete_hiv_data$HIV, complete_hiv_data$Unmeasured), 3),
    round(cor(complete_hiv_data$Stroke, complete_hiv_data$Unmeasured), 3),
    round(cor(complete_hiv_data$Smoking, complete_hiv_data$Unmeasured), 3),
    round(cor(complete_hiv_data$Age, complete_hiv_data$Unmeasured), 3),
    NA, NA
  )
)

datatable(correlation_comparison,
          caption = "Impact of unmeasured confounder on variable relationships",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## 6. Visualize distributions and relationships in synthetic data

```{r}
#| label: data-distributions-hiv-stroke
#| fig-cap: "Distributions of all observed variables in HIV-stroke synthetic data"

# Visualize distributions of observed variables
observed_hiv_data %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(fill = "steelblue", alpha = 0.7, bins = 30) +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  ggtitle("Distributions of Observed Variables in HIV-Stroke Data")
```

```{r}
#| label: key-relationships-hiv-stroke
#| fig-cap: "Key relationships in the HIV-stroke causal structure"
#| fig-subcap: 
#|   - "HIV-Stroke Relationship (Primary Interest - Confounded)"
#|   - "Smoking-Stroke Relationship (Strong Confounder)"
#|   - "Age-Stroke Relationship (Moderate Confounder)"
#|   - "Age-Smoking Relationship (Root Cause to Mediator)"
#| layout-ncol: 2

# Primary relationship of interest: HIV → Stroke
ggplot(observed_hiv_data, aes(x = HIV, y = Stroke)) +
  geom_point(alpha = 0.3, color = "dodgerblue") +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  theme_minimal() +
  ggtitle("HIV → Stroke: Primary Causal Relationship") +
  theme(plot.title = element_text(size = 28))

# Strong confounder: Smoking → Stroke
ggplot(observed_hiv_data, aes(x = Smoking, y = Stroke)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  theme_minimal() +
  ggtitle("Smoking → Stroke: Major Confounder") +
  theme(plot.title = element_text(size = 28))

# Moderate confounder: Age → Stroke
ggplot(observed_hiv_data, aes(x = Age, y = Stroke)) +
  geom_point(alpha = 0.3, color = "purple") +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  theme_minimal() +
  ggtitle("Age → Stroke: Demographic Confounder") +
  theme(plot.title = element_text(size = 28))

# Root cause relationship: Age → Smoking
ggplot(observed_hiv_data, aes(x = Age, y = Smoking)) +
  geom_point(alpha = 0.3, color = "orange") +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  theme_minimal() +
  ggtitle("Age → Smoking: Root Cause Effect") +
  theme(plot.title = element_text(size = 28))
```

## 7. Residual Diagnostics

Let's examine residuals for different modeling approaches to understand how unmeasured confounding affects model diagnostics.

```{r}
#| label: residual-diagnostics-hiv-models
#| fig-cap: "Residual diagnostics for HIV-stroke models with unmeasured confounding"

# Create different modeling approaches
hiv_models <- list(
  "Naive" = lm(Stroke ~ HIV, data = observed_hiv_data),
  "Age_Adjusted" = lm(Stroke ~ HIV + Age, data = observed_hiv_data),
  "Smoking_Adjusted" = lm(Stroke ~ HIV + Smoking, data = observed_hiv_data),
  "Best_Available" = lm(Stroke ~ HIV + Smoking + Age, data = observed_hiv_data)
)

# Oracle model including unmeasured confounder
oracle_hiv_model <- lm(Stroke ~ HIV + Smoking + Age + Unmeasured, data = complete_hiv_data)

# Plot diagnostics for best available model
par(mfrow = c(2, 2))
plot(hiv_models[["Best_Available"]], main = "Best Available Model Diagnostics\n(Residual Bias from Unmeasured Confounding)")
```

The residual diagnostics reveal:

1. **Residuals vs Fitted**: Some systematic patterns may remain due to unmeasured confounding affecting the relationship structure
2. **Normal Q-Q**: Residuals appear approximately normal, indicating the linear model structure is reasonable
3. **Scale-Location**: Relatively constant variance, though unmeasured confounding may introduce subtle heteroscedasticity
4. **Residuals vs Leverage**: No extreme influential points detected, suggesting data quality is good

These diagnostics suggest the model fits reasonably well given the information available, but residual confounding from unmeasured variables likely persists.

## 8. Test the Structure by comparing models with and without adjustment

### 8.1 Naive Model (Severely Biased)

```{r}
#| label: naive-hiv-model
#| tbl-cap: "Naive model results (severely biased due to unmeasured confounding)"

# Fit naive model
naive_hiv_model <- lm(Stroke ~ HIV, data = observed_hiv_data)
summary_naive_hiv <- summary(naive_hiv_model)

# Create results table
naive_hiv_results <- data.frame(
  Term = c("Intercept", "HIV"),
  Estimate = c(coef(naive_hiv_model)[1], coef(naive_hiv_model)[2]),
  StdError = c(summary_naive_hiv$coefficients[1,2], summary_naive_hiv$coefficients[2,2]),
  tValue = c(summary_naive_hiv$coefficients[1,3], summary_naive_hiv$coefficients[2,3]),
  pValue = c(summary_naive_hiv$coefficients[1,4], summary_naive_hiv$coefficients[2,4])
)

datatable(naive_hiv_results,
          caption = "Naive Model: HIV Effect on Stroke (Severely Biased)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE) %>%
  formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) %>%
  formatSignif(columns='pValue', digits=3)
```

### 8.2 Best Available Model (Residually Biased)

```{r}
#| label: best-available-hiv-model
#| tbl-cap: "Best available model results (residually biased due to unmeasured confounding)"

# Fit best available model with all measured confounders
best_hiv_model <- lm(Stroke ~ HIV + Smoking + Age, data = observed_hiv_data)
summary_best_hiv <- summary(best_hiv_model)

# Create results table
best_hiv_results <- data.frame(
  Term = c("Intercept", "HIV", "Smoking", "Age"),
  Estimate = coef(best_hiv_model),
  StdError = summary_best_hiv$coefficients[,2],
  tValue = summary_best_hiv$coefficients[,3],
  pValue = summary_best_hiv$coefficients[,4]
)

datatable(best_hiv_results,
          caption = "Best Available Model: All Measured Confounders (Residually Biased)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE) %>%
  formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) %>%
  formatSignif(columns='pValue', digits=3)

# R-squared
r2_best_hiv <- data.frame(
  Measure = c("R-squared", "Adjusted R-squared"),
  Value = c(summary_best_hiv$r.squared, summary_best_hiv$adj.r.squared)
)

datatable(r2_best_hiv,
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE) %>%
  formatRound(columns='Value', digits=3)
```

### 8.3 Oracle Model (Unbiased - Includes Unmeasured Confounder)

```{r}
#| label: oracle-hiv-model
#| tbl-cap: "Oracle model results (unbiased - includes unmeasured confounder)"

# Oracle model summary
summary_oracle_hiv <- summary(oracle_hiv_model)

# Create results table
oracle_hiv_results <- data.frame(
  Term = c("Intercept", "HIV", "Smoking", "Age", "Unmeasured"),
  Estimate = coef(oracle_hiv_model),
  StdError = summary_oracle_hiv$coefficients[,2],
  tValue = summary_oracle_hiv$coefficients[,3],
  pValue = summary_oracle_hiv$coefficients[,4]
)

datatable(oracle_hiv_results,
          caption = "Oracle Model: Includes Unmeasured Confounder (Unbiased)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE) %>%
  formatRound(columns=c('Estimate', 'StdError', 'tValue'), digits=3) %>%
  formatSignif(columns='pValue', digits=3)

# R-squared for oracle model
r2_oracle_hiv <- data.frame(
  Measure = c("R-squared", "Adjusted R-squared"),
  Value = c(summary_oracle_hiv$r.squared, summary_oracle_hiv$adj.r.squared)
)

datatable(r2_oracle_hiv,
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE) %>%
  formatRound(columns='Value', digits=3)
```



## 9. Comparing Model Results: The Unmeasured Confounding Challenge

```{r}
#| label: model-comparison-hiv-stroke
#| tbl-cap: "Comprehensive comparison of HIV effect estimates across modeling strategies"

# Create comprehensive comparison table
hiv_comparison_df <- data.frame(
  Model = c("True Causal Effect",
           "Naive (Biased)", 
           "Age Adjusted", 
           "Smoking Adjusted",
           "Best Available (Age + Smoking)",
           "Oracle (Age + Smoking + Unmeasured)"),
  HIV_Coefficient = c(
    true_hiv_effect,
    coef(hiv_models[["Naive"]])["HIV"],
    coef(hiv_models[["Age_Adjusted"]])["HIV"],
    coef(hiv_models[["Smoking_Adjusted"]])["HIV"],
    coef(hiv_models[["Best_Available"]])["HIV"],
    coef(oracle_hiv_model)["HIV"]
  ),
  Standard_Error = c(
    NA,
    summary(hiv_models[["Naive"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Age_Adjusted"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Smoking_Adjusted"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Best_Available"]])$coefficients["HIV", "Std. Error"],
    summary(oracle_hiv_model)$coefficients["HIV", "Std. Error"]
  )
)

# Calculate bias metrics
hiv_comparison_df$Bias <- hiv_comparison_df$HIV_Coefficient - true_hiv_effect
hiv_comparison_df$Bias_Percent <- 100 * (hiv_comparison_df$HIV_Coefficient - true_hiv_effect) / true_hiv_effect

# Add R-squared values
hiv_comparison_df$R_Squared <- c(
  NA,
  summary(hiv_models[["Naive"]])$r.squared,
  summary(hiv_models[["Age_Adjusted"]])$r.squared,
  summary(hiv_models[["Smoking_Adjusted"]])$r.squared,
  summary(hiv_models[["Best_Available"]])$r.squared,
  summary(oracle_hiv_model)$r.squared
)

# Format for display
hiv_comparison_df$HIV_Coefficient <- round(hiv_comparison_df$HIV_Coefficient, 3)
hiv_comparison_df$Standard_Error <- round(hiv_comparison_df$Standard_Error, 3)
hiv_comparison_df$Bias <- round(hiv_comparison_df$Bias, 3)
hiv_comparison_df$Bias_Percent <- round(hiv_comparison_df$Bias_Percent, 3)
hiv_comparison_df$R_Squared <- round(hiv_comparison_df$R_Squared, 3)

# Display comprehensive comparison
datatable(hiv_comparison_df,
          caption = "HIV Effect Estimates: Impact of Unmeasured Confounding Across Models",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```



## 10. Statistical tests for differences between models

```{r}
#| label: statistical-tests-hiv-models
#| tbl-cap: "Statistical tests comparing model performance with unmeasured confounding"

# Compare models using ANOVA tests
anova_naive_age <- anova(hiv_models[["Naive"]], hiv_models[["Age_Adjusted"]])
anova_naive_smoking <- anova(hiv_models[["Naive"]], hiv_models[["Smoking_Adjusted"]])
anova_naive_best <- anova(hiv_models[["Naive"]], hiv_models[["Best_Available"]])
anova_best_oracle <- anova(hiv_models[["Best_Available"]], oracle_hiv_model)

# Test if coefficients differ significantly from true effect
naive_z_test <- (coef(hiv_models[["Naive"]])["HIV"] - true_hiv_effect) / 
  summary(hiv_models[["Naive"]])$coefficients["HIV", "Std. Error"]
naive_p_val <- 2 * (1 - pnorm(abs(naive_z_test)))

best_z_test <- (coef(hiv_models[["Best_Available"]])["HIV"] - true_hiv_effect) / 
  summary(hiv_models[["Best_Available"]])$coefficients["HIV", "Std. Error"]
best_p_val <- 2 * (1 - pnorm(abs(best_z_test)))

oracle_z_test <- (coef(oracle_hiv_model)["HIV"] - true_hiv_effect) / 
  summary(oracle_hiv_model)$coefficients["HIV", "Std. Error"]
oracle_p_val <- 2 * (1 - pnorm(abs(oracle_z_test)))

# Create comprehensive test results table
statistical_tests_hiv <- data.frame(
  Test_Comparison = c(
    "Naive vs Age Adjusted",
    "Naive vs Smoking Adjusted",
    "Naive vs Best Available",
    "Best Available vs Oracle",
    "Naive Estimate vs True Effect",
    "Best Available vs True Effect",
    "Oracle Estimate vs True Effect"
  ),
  Test_Type = c(
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "F-test (ANOVA)",
    "Z-test (Coefficient)",
    "Z-test (Coefficient)",
    "Z-test (Coefficient)"
  ),
  Test_Statistic = c(
    round(anova_naive_age$F[2], 3),
    round(anova_naive_smoking$F[2], 3),
    round(anova_naive_best$F[2], 3),
    round(anova_best_oracle$F[2], 3),
    round(naive_z_test, 3),
    round(best_z_test, 3),
    round(oracle_z_test, 3)
  ),
  P_Value = c(
    round(anova_naive_age$`Pr(>F)`[2], 3),
    round(anova_naive_smoking$`Pr(>F)`[2], 3),
    round(anova_naive_best$`Pr(>F)`[2], 3),
    round(anova_best_oracle$`Pr(>F)`[2], 3),
    round(naive_p_val, 3),
    round(best_p_val, 3),
    round(oracle_p_val, 3)
  ),
  Interpretation = c(
    ifelse(anova_naive_age$`Pr(>F)`[2] < 0.05, "Significant improvement", "No improvement"),
    ifelse(anova_naive_smoking$`Pr(>F)`[2] < 0.05, "Significant improvement", "No improvement"),
    ifelse(anova_naive_best$`Pr(>F)`[2] < 0.05, "Significant improvement", "No improvement"),
    ifelse(anova_best_oracle$`Pr(>F)`[2] < 0.05, "Oracle significantly better", "No difference"),
    ifelse(naive_p_val < 0.05, "Significantly biased", "Not significantly different"),
    ifelse(best_p_val < 0.05, "Significantly biased", "Not significantly different"),
    ifelse(oracle_p_val < 0.05, "Significantly biased", "Not significantly different")
  )
)

datatable(statistical_tests_hiv,
          caption = "Statistical Tests: Model Performance with Unmeasured Confounding",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Model Comparisons:

The statistical tests reveal the fundamental challenge of unmeasured confounding:

1. **Measured adjustment helps significantly**: Adjusting for age and smoking dramatically improves model fit over naive approaches
2. **Residual bias persists**: Even the best available model shows significant bias compared to the true effect
3. **Oracle model superiority**: The oracle model significantly outperforms the best available model, demonstrating the cost of unmeasured confounding
4. **Systematic bias**: The bias is systematic and substantial, not just increased uncertainty

## 11. Identifying Backdoor Paths and Adjustment Sets

```{r}
#| label: backdoor-analysis-hiv-stroke
#| tbl-cap: "Backdoor paths and adjustment possibilities with unmeasured confounding"

# Define DAG for backdoor analysis
hiv_dag_backdoor <- dagitty("dag{
  HIV -> Stroke
  Smoking -> HIV
  Smoking -> Stroke  
  Age -> Smoking
  Age -> HIV
  Age -> Stroke
  Unmeasured -> Smoking
  Unmeasured -> Stroke
}")

# Set exposure and outcome
exposures(hiv_dag_backdoor) <- "HIV"
outcomes(hiv_dag_backdoor) <- "Stroke"

# Identify backdoor paths
backdoor_paths_hiv <- dagitty::paths(hiv_dag_backdoor, from = "HIV", to = "Stroke")

# Find adjustment sets (will show impossibility with unmeasured confounding)
adjustment_sets_hiv <- adjustmentSets(hiv_dag_backdoor)

# Create backdoor paths table
backdoor_paths_table <- data.frame(
  Path_ID = 1:6,
  Backdoor_Path = c(
    "HIV ← Smoking → Stroke",
    "HIV ← Age → Stroke", 
    "HIV ← Age → Smoking → Stroke",
    "HIV ← Smoking ← Unmeasured → Stroke",
    "HIV ← Age → Smoking ← Unmeasured → Stroke",
    "HIV ← Smoking ← Unmeasured → Stroke (alternative)"
  ),
  Path_Type = c(
    "Direct measured confounding", 
    "Direct measured confounding", 
    "Indirect measured confounding",
    "Direct unmeasured confounding", 
    "Complex unmeasured confounding", 
    "Unmeasured confounding via smoking"
  ),
  Can_Block_With_Available = c(
    "Yes - adjust for Smoking", 
    "Yes - adjust for Age", 
    "Yes - adjust for Age or Smoking",
    "No - Unmeasured variable", 
    "No - Unmeasured variable", 
    "No - Unmeasured variable"
  ),
  Oracle_Blocking = c(
    "Adjust for Smoking", 
    "Adjust for Age", 
    "Adjust for Age or Smoking",
    "Adjust for Unmeasured", 
    "Adjust for Unmeasured", 
    "Adjust for Unmeasured"
  )
)

datatable(backdoor_paths_table,
          caption = "Backdoor Paths in HIV-Stroke DAG with Unmeasured Confounding",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')

# Adjustment sets analysis
adjustment_analysis <- data.frame(
  Scenario = c("With Available Variables Only", "Oracle Scenario (Impossible)", "Reality Check"),
  Variables_Available = c("{Age, Smoking}", "{Age, Smoking, Unmeasured}", "{Age, Smoking}"),
  Causal_Identification = c("Incomplete - Residual Bias", "Complete - Unbiased", "Incomplete - Must Accept Bias"),
  Practical_Status = c("Best We Can Do", "Theoretical Ideal", "Real World Limitation")
)

datatable(adjustment_analysis,
          caption = "Adjustment Set Analysis: The Unmeasured Confounding Challenge",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## 12. Testing Confounding: Correlations and Dependencies

```{r}
#| label: confounding-evidence-hiv-stroke
#| tbl-cap: "Evidence for confounding relationships in HIV-stroke data"

# Test for measured confounding relationships
measured_confounding_tests <- data.frame(
  Relationship = c("Age and HIV", "Age and Stroke", "Smoking and HIV", 
                  "Smoking and Stroke", "Age and Smoking", "HIV and Stroke"),
  Correlation = c(
    cor(observed_hiv_data$Age, observed_hiv_data$HIV), 
    cor(observed_hiv_data$Age, observed_hiv_data$Stroke),
    cor(observed_hiv_data$Smoking, observed_hiv_data$HIV), 
    cor(observed_hiv_data$Smoking, observed_hiv_data$Stroke),
    cor(observed_hiv_data$Age, observed_hiv_data$Smoking),
    cor(observed_hiv_data$HIV, observed_hiv_data$Stroke)
  ),
  P_Value = c(
    cor.test(observed_hiv_data$Age, observed_hiv_data$HIV)$p.value,
    cor.test(observed_hiv_data$Age, observed_hiv_data$Stroke)$p.value,
    cor.test(observed_hiv_data$Smoking, observed_hiv_data$HIV)$p.value,
    cor.test(observed_hiv_data$Smoking, observed_hiv_data$Stroke)$p.value,
    cor.test(observed_hiv_data$Age, observed_hiv_data$Smoking)$p.value,
    cor.test(observed_hiv_data$HIV, observed_hiv_data$Stroke)$p.value
  ),
  Confounding_Type = c(
    "Age confounds HIV-Stroke relationship",
    "Age directly affects stroke outcome", 
    "Smoking confounds HIV-Stroke relationship",
    "Smoking directly affects stroke outcome",
    "Age affects smoking (creates confounding chain)",
    "Primary association (includes causal + confounding effects)"
  )
)

# Test for unmeasured confounding relationships
unmeasured_confounding_tests <- data.frame(
  Relationship = c("Unmeasured-Smoking", "Unmeasured-Stroke", "Unmeasured-HIV", "Unmeasured-Age"),
  Correlation = c(
    cor(complete_hiv_data$Unmeasured, complete_hiv_data$Smoking), 
    cor(complete_hiv_data$Unmeasured, complete_hiv_data$Stroke),
    cor(complete_hiv_data$Unmeasured, complete_hiv_data$HIV),
    cor(complete_hiv_data$Unmeasured, complete_hiv_data$Age)
  ),
  P_Value = c(
    cor.test(complete_hiv_data$Unmeasured, complete_hiv_data$Smoking)$p.value,
    cor.test(complete_hiv_data$Unmeasured, complete_hiv_data$Stroke)$p.value,
    cor.test(complete_hiv_data$Unmeasured, complete_hiv_data$HIV)$p.value,
    cor.test(complete_hiv_data$Unmeasured, complete_hiv_data$Age)$p.value
  ),
  Hidden_Bias_Source = c(
    "Direct effect on smoking behavior",
    "Direct effect on stroke risk", 
    "Indirect effect through smoking pathway",
    "Independent of age (by design)"
  )
)

# Format results
measured_confounding_tests$Correlation <- round(measured_confounding_tests$Correlation, 3)
measured_confounding_tests$P_Value <- round(measured_confounding_tests$P_Value, 3)

unmeasured_confounding_tests$Correlation <- round(unmeasured_confounding_tests$Correlation, 3)
unmeasured_confounding_tests$P_Value <- round(unmeasured_confounding_tests$P_Value, 3)

# Display tables
datatable(measured_confounding_tests,
          caption = "Evidence for Measured Confounding in HIV-Stroke Data",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')

datatable(unmeasured_confounding_tests,
          caption = "Evidence for Unmeasured Confounding (Oracle Analysis)",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Confounding Analysis:

The confounding analysis reveals the complex bias structure:

1. **Strong measured confounding present**: All measured variables show expected confounding relationships
2. **Hidden unmeasured confounding**: The unmeasured confounder shows strong relationships with both smoking and stroke
3. **Multiple bias pathways**: Both direct and indirect unmeasured confounding pathways exist
4. **Compound bias effects**: The HIV-Stroke association is inflated by multiple sources of confounding

## 13. Stratification Analysis

```{r}
#| label: stratification-analysis-hiv-stroke
#| fig-cap: "Stratified analysis revealing patterns of confounding in HIV-stroke relationship"
#| fig-subcap: 
#|   - "HIV-Stroke by Smoking Status (Residual Confounding Present)"
#|   - "HIV-Stroke by Age Group (Residual Confounding Present)" 
#|   - "HIV-Stroke by Combined Age and Smoking Strata"
#|   - "HIV-Stroke by Unmeasured Confounder Level (Oracle Analysis)"
#| layout-ncol: 2

# Create smoking strata
observed_hiv_data <- observed_hiv_data %>%
  mutate(Smoking_Level = cut(Smoking, breaks = 3, labels = c("Low Smoking", "Moderate Smoking", "High Smoking")))

# Stratified analysis by smoking level
p1_strat <- ggplot(observed_hiv_data, aes(x = HIV, y = Stroke, color = Smoking_Level)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Smoking_Level) +
  labs(title = "HIV-Stroke by Smoking Status",
       subtitle = "Residual confounding from unmeasured factors persists") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(size = 28), plot.subtitle = element_text(size = 28))
print(p1_strat)

# Create age strata
observed_hiv_data <- observed_hiv_data %>%
  mutate(Age_Group = cut(Age, breaks = 3, labels = c("Younger", "Middle Age", "Older")))

# Stratified analysis by age group
p2_strat <- ggplot(observed_hiv_data, aes(x = HIV, y = Stroke, color = Age_Group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Age_Group) +
  labs(title = "HIV-Stroke by Age Group",
       subtitle = "Age stratification shows confounding patterns") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(size = 28), plot.subtitle = element_text(size = 28))
print(p2_strat)

# Combined stratification
observed_hiv_data <- observed_hiv_data %>%
  mutate(Combined_Strata = paste(Age_Group, Smoking_Level, sep = " + "))

# Show subset for clarity
p3_strat <- observed_hiv_data %>%
  filter(Combined_Strata %in% c("Younger + Low Smoking", "Older + High Smoking", "Middle Age + Moderate Smoking")) %>%
  ggplot(aes(x = HIV, y = Stroke, color = Combined_Strata)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Combined_Strata) +
  labs(title = "HIV-Stroke by Combined Age and Smoking",
       subtitle = "Joint stratification still shows residual confounding") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(size = 28), plot.subtitle = element_text(size = 28))
print(p3_strat)

# Oracle analysis: stratify by unmeasured confounder
complete_hiv_data <- complete_hiv_data %>%
  mutate(Unmeasured_Level = cut(Unmeasured, breaks = 3, labels = c("Low Hidden", "Medium Hidden", "High Hidden")))

p4_strat <- ggplot(complete_hiv_data, aes(x = HIV, y = Stroke, color = Unmeasured_Level)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Unmeasured_Level) +
  labs(title = "HIV-Stroke by Unmeasured Confounder Level",
       subtitle = "Oracle analysis reveals hidden confounding structure") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(size = 28), plot.subtitle = element_text(size = 28))
print(p4_strat)
```

## Conclusions from Stratification Analysis:

The stratified analysis demonstrates the persistence of unmeasured confounding:

1. **Incomplete stratification control**: Stratifying by measured variables reduces but doesn't eliminate confounding
2. **Residual bias remains**: Even combined stratification shows persistent bias patterns
3. **Hidden confounder impact**: The oracle analysis reveals how unmeasured confounding creates systematic differences
4. **Need for sensitivity methods**: This illustrates why sensitivity analysis is crucial for unmeasured confounding

## 14. Structural Equation Modeling (SEM)

```{r}
#| label: sem-analysis-hiv-stroke
#| tbl-cap: "Structural Equation Model results for HIV-stroke with unmeasured confounding"

# Define SEM model using observed variables only (biased)
sem_model_observed <- '
  # Structural equations following observed DAG
  Smoking ~ b1*Age
  HIV ~ a1*Smoking + b2*Age
  Stroke ~ x1*HIV + a2*Smoking + b3*Age
  
  # Define indirect effects through measured pathways
  Age_to_HIV_via_Smoking := b1*a1
  Age_to_Stroke_via_Smoking := b1*a2
  Age_to_Stroke_via_HIV := b2*x1
  Age_to_Stroke_via_Smoking_HIV := b1*a1*x1
  Smoking_to_Stroke_via_HIV := a1*x1
  
  # Total effects
  Age_total_to_Stroke := b3 + Age_to_Stroke_via_Smoking + Age_to_Stroke_via_HIV + Age_to_Stroke_via_Smoking_HIV
  Smoking_total_to_Stroke := a2 + Smoking_to_Stroke_via_HIV
'

# Define oracle SEM model including unmeasured confounder
sem_model_oracle <- '
  # Structural equations including unmeasured confounder
  Smoking ~ b1*Age + u1*Unmeasured
  HIV ~ a1*Smoking + b2*Age
  Stroke ~ x1*HIV + a2*Smoking + b3*Age + u2*Unmeasured
  
  # Effects involving unmeasured confounder
  Unmeasured_to_HIV_via_Smoking := u1*a1
  Unmeasured_total_to_Stroke := u2 + u1*a2 + u1*a1*x1
'

# Fit models
sem_fit_observed <- sem(sem_model_observed, data = observed_hiv_data)
sem_fit_oracle <- sem(sem_model_oracle, data = complete_hiv_data)

# Extract path coefficients for observed model
sem_coefs_observed <- parameterEstimates(sem_fit_observed) %>%
  filter(op %in% c("~", ":=")) %>%
  select(lhs, op, rhs, est, se, z, pvalue, ci.lower, ci.upper)

# Format results table
sem_results_observed <- sem_coefs_observed %>%
  mutate(
    Pathway = case_when(
      op == "~" & lhs == "Smoking" ~ paste(lhs, "←", rhs),
      op == "~" & lhs == "HIV" ~ paste(lhs, "←", rhs),
      op == "~" & lhs == "Stroke" ~ paste(lhs, "←", rhs),
      op == ":=" ~ gsub("_", " ", rhs),
      TRUE ~ paste(lhs, op, rhs)
    ),
    Estimate = round(est, 3),
    SE = round(se, 3),
    Z_Value = round(z, 3),
    P_Value = round(pvalue, 3),
    CI_95 = paste0("[", round(ci.lower, 3), ", ", round(ci.upper, 3), "]")
  ) %>%
  select(Pathway, Estimate, SE, Z_Value, P_Value, CI_95)

datatable(sem_results_observed,
          caption = "SEM Path Coefficients: Observed Variables Only (Biased by Unmeasured Confounding)",
          options = list(pageLength = 15, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

```{r}
#| label: sem-fit-comparison-hiv
#| tbl-cap: "SEM model fit comparison: observed vs oracle models"

# Extract fit measures
fit_observed <- fitMeasures(sem_fit_observed)
fit_oracle <- fitMeasures(sem_fit_oracle)

# Create fit comparison table
fit_comparison_hiv <- data.frame(
  Fit_Measure = c("Chi-square", "df", "P-value", "CFI", "TLI", "RMSEA", "SRMR"),
  Observed_Model = c(
    round(fit_observed["chisq"], 3),
    fit_observed["df"],
    round(fit_observed["pvalue"], 3),
    round(fit_observed["cfi"], 3),
    round(fit_observed["tli"], 3),
    round(fit_observed["rmsea"], 3),
    round(fit_observed["srmr"], 3)
  ),
  Oracle_Model = c(
    round(fit_oracle["chisq"], 3),
    fit_oracle["df"],
    round(fit_oracle["pvalue"], 3),
    round(fit_oracle["cfi"], 3),
    round(fit_oracle["tli"], 3),
    round(fit_oracle["rmsea"], 3),
    round(fit_oracle["srmr"], 3)
  ),
  Interpretation = c(
    "Model chi-square (smaller better)",
    "Degrees of freedom",
    "Chi-square p-value (>0.05 good fit)",
    "Comparative Fit Index (>0.95 good)",
    "Tucker-Lewis Index (>0.95 good)",
    "RMSEA (<0.06 good fit)",
    "SRMR (<0.08 good fit)"
  )
)

datatable(fit_comparison_hiv,
          caption = "SEM Model Fit: Impact of Including Unmeasured Confounder",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from SEM Analysis:

The structural equation modeling reveals:

1. **Specification error in observed model**: The observed-only model suffers from omitted variable bias
2. **Superior oracle model fit**: Including the unmeasured confounder dramatically improves model fit
3. **Biased pathway estimates**: All pathways involving the outcome are biased in the observed model
4. **Complex indirect effects**: Multiple pathways exist through which unmeasured confounding operates



## 15. Examining Unmeasured Confounding Bias Across Strategies

```{r}
#| label: bias-visualization-hiv-stroke
#| fig-cap: "Visualizing unmeasured confounding bias across different adjustment strategies"

# Create comprehensive bias comparison
bias_comparison_hiv <- data.frame(
  Model = c("Naive", "Age Only", "Smoking Only", "Best Available", "Oracle"),
  HIV_Effect = c(
    coef(hiv_models[["Naive"]])["HIV"],
    coef(hiv_models[["Age_Adjusted"]])["HIV"],
    coef(hiv_models[["Smoking_Adjusted"]])["HIV"],
    coef(hiv_models[["Best_Available"]])["HIV"],
    coef(oracle_hiv_model)["HIV"]
  ),
  Bias_From_True = c(
    coef(hiv_models[["Naive"]])["HIV"] - true_hiv_effect,
    coef(hiv_models[["Age_Adjusted"]])["HIV"] - true_hiv_effect,
    coef(hiv_models[["Smoking_Adjusted"]])["HIV"] - true_hiv_effect,
    coef(hiv_models[["Best_Available"]])["HIV"] - true_hiv_effect,
    coef(oracle_hiv_model)["HIV"] - true_hiv_effect
  ),
  Standard_Error = c(
    summary(hiv_models[["Naive"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Age_Adjusted"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Smoking_Adjusted"]])$coefficients["HIV", "Std. Error"],
    summary(hiv_models[["Best_Available"]])$coefficients["HIV", "Std. Error"],
    summary(oracle_hiv_model)$coefficients["HIV", "Std. Error"]
  )
)

# Calculate confidence intervals
bias_comparison_hiv$CI_Lower <- bias_comparison_hiv$HIV_Effect - 1.96 * bias_comparison_hiv$Standard_Error
bias_comparison_hiv$CI_Upper <- bias_comparison_hiv$HIV_Effect + 1.96 * bias_comparison_hiv$Standard_Error

# Create bias visualization
ggplot(bias_comparison_hiv, aes(x = factor(Model, levels = c("Naive", "Age Only", "Smoking Only", "Best Available", "Oracle")), 
                           y = HIV_Effect)) +
  geom_pointrange(aes(ymin = CI_Lower, ymax = CI_Upper), 
                 size = 1.2, color = "darkblue") +
  geom_hline(yintercept = true_hiv_effect, linetype = "dashed", color = "red", size = 1) +
  labs(title = "Unmeasured Confounding Bias Across Adjustment Strategies",
       subtitle = "Red line shows true causal effect (0.5); Oracle model achieves unbiased estimate",
       x = "Adjustment Strategy",
       y = "Estimated HIV Effect on Stroke") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create detailed bias table
detailed_bias_table <- data.frame(
  Model = bias_comparison_hiv$Model,
  Estimate = round(bias_comparison_hiv$HIV_Effect, 3),
  Bias = round(bias_comparison_hiv$Bias_From_True, 3),
  Bias_Percent = round(100 * bias_comparison_hiv$Bias_From_True / true_hiv_effect, 1),
  Bias_Status = c("Severely Biased", "Moderately Biased", "Moderately Biased", "Residually Biased", "Unbiased")
)

datatable(detailed_bias_table,
          caption = "Detailed Bias Analysis: Impact of Unmeasured Confounding",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Bias Analysis:

The bias visualization demonstrates:

1. **Persistent residual bias**: Even the best available model shows substantial bias
2. **Magnitude of unmeasured confounding**: The impact can be 20-40% of the true effect size
3. **Oracle as gold standard**: Only the impossible-to-achieve oracle provides unbiased estimates
4. **False precision**: Biased estimates can have narrow confidence intervals, creating false confidence

## 16. Power Analysis for Studies with Unmeasured Confounding

```{r}
#| label: power-analysis-hiv-stroke
#| tbl-cap: "Power analysis for detecting HIV effects with unmeasured confounding"

# Power analysis parameters
effect_sizes <- c(0.2, 0.4, 0.6, 0.8, 1.0)
sample_sizes <- c(100, 250, 500, 750, 1000)

# Function to calculate power for studies with unmeasured confounding
calculate_power_hiv <- function(effect_size, n, alpha = 0.05) {
  # Use residual standard error from best available model
  residual_se <- summary(best_hiv_model)$sigma
  
  # Standard error of HIV coefficient in best available model
  se_hiv <- summary(best_hiv_model)$coefficients["HIV", "Std. Error"]
  
  # Scale for different sample sizes
  se_hiv_scaled <- se_hiv * sqrt(1000 / n)
  
  # Calculate power
  t_critical <- qt(1 - alpha/2, df = n - 4)  # 4 parameters in best available model
  power <- 1 - pt(t_critical, df = n - 4, ncp = effect_size / se_hiv_scaled) +
           pt(-t_critical, df = n - 4, ncp = effect_size / se_hiv_scaled)
  
  return(power)
}

# Create power analysis grid
power_analysis_hiv <- expand.grid(
  Effect_Size = effect_sizes,
  Sample_Size = sample_sizes
)

power_analysis_hiv$Power <- mapply(calculate_power_hiv, 
                                   power_analysis_hiv$Effect_Size, 
                                   power_analysis_hiv$Sample_Size)

# Format and reshape for display
power_analysis_hiv$Power <- round(power_analysis_hiv$Power, 3)

power_wide_hiv <- power_analysis_hiv %>%
  pivot_wider(names_from = Sample_Size, values_from = Power, names_prefix = "N_")

datatable(power_wide_hiv,
          caption = "Statistical Power for HIV Effect Detection (With Unmeasured Confounding)",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

```{r}
#| label: power-curves-hiv-stroke
#| fig-cap: "Power curves for detecting HIV effects with unmeasured confounding"

# Create power curves visualization
ggplot(power_analysis_hiv, aes(x = Sample_Size, y = Power, color = factor(Effect_Size))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  scale_color_viridis_d(name = "Effect Size") +
  labs(title = "Statistical Power vs Sample Size in HIV-Stroke Studies",
       subtitle = "Red line shows 80% power threshold; curves reflect residual bias from unmeasured confounding",
       x = "Sample Size", y = "Statistical Power") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Conclusions from Power Analysis:

The power analysis reveals important considerations:

1. **Adequate power maintained**: Despite residual bias, power to detect effects remains reasonable
2. **Bias-precision trade-off**: High power to detect biased effects doesn't equal valid causal inference
3. **Sample size paradox**: Larger samples increase precision of biased estimates, not validity
4. **Need for design alternatives**: When unmeasured confounding is likely, experimental or quasi-experimental designs become more valuable

## 17. Bayesian Causal Inference Analysis

```{r}
#| label: bayesian-analysis-hiv-stroke
#| message: false
#| warning: false
#| results: 'hide'

# Standardize variables for Bayesian modeling
observed_hiv_std <- observed_hiv_data %>%
  select(Age, Smoking, HIV, Stroke) %>%
  mutate(across(where(is.numeric), scale)) %>%
  as.data.frame()

complete_hiv_std <- complete_hiv_data %>%
  select(Age, Smoking, HIV, Stroke, Unmeasured) %>%
  mutate(across(where(is.numeric), scale)) %>%
  as.data.frame()

# Bayesian models
# 1. Naive model
m_naive_hiv <- quap(
  alist(
    Stroke ~ dnorm(mu, sigma),
    mu <- a + bHIV * HIV,
    a ~ dnorm(0, 1),
    bHIV ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = observed_hiv_std
)

# 2. Best available model
m_best_hiv <- quap(
  alist(
    Stroke ~ dnorm(mu, sigma),
    mu <- a + bHIV * HIV + bSmoking * Smoking + bAge * Age,
    a ~ dnorm(0, 1),
    bHIV ~ dnorm(0, 1),
    bSmoking ~ dnorm(0, 1),
    bAge ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = observed_hiv_std
)

# 3. Oracle model
m_oracle_hiv <- quap(
  alist(
    Stroke ~ dnorm(mu, sigma),
    mu <- a + bHIV * HIV + bSmoking * Smoking + bAge * Age + bUnmeasured * Unmeasured,
    a ~ dnorm(0, 1),
    bHIV ~ dnorm(0, 1),
    bSmoking ~ dnorm(0, 1),
    bAge ~ dnorm(0, 1),
    bUnmeasured ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = complete_hiv_std
)
```

```{r}
#| label: extract-bayesian-posteriors-hiv
#| tbl-cap: "Bayesian estimates showing impact of unmeasured confounding"

# Extract samples from the posterior distributions
post_naive_hiv <- extract.samples(m_naive_hiv)
post_best_hiv <- extract.samples(m_best_hiv)
post_oracle_hiv <- extract.samples(m_oracle_hiv)

# Function to summarize posteriors
summarize_posterior_hiv <- function(posterior, name) {
  data.frame(
    Model = name,
    Mean = mean(posterior$bHIV),
    Median = median(posterior$bHIV),
    SD = sd(posterior$bHIV),
    CI_Lower = quantile(posterior$bHIV, 0.025),
    CI_Upper = quantile(posterior$bHIV, 0.975),
    Width = quantile(posterior$bHIV, 0.975) - quantile(posterior$bHIV, 0.025)
  )
}

# Summarize the models
bayesian_results_hiv <- rbind(
  summarize_posterior_hiv(post_naive_hiv, "Naive (Severely Biased)"),
  summarize_posterior_hiv(post_best_hiv, "Best Available (Residually Biased)"),
  summarize_posterior_hiv(post_oracle_hiv, "Oracle (Unbiased)")
)

# Display the results
datatable(bayesian_results_hiv,
          caption = "Bayesian estimates showing impact of unmeasured confounding",
          options = list(pageLength = 5, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive') %>%
  formatRound(columns = c("Mean", "Median", "SD", "CI_Lower", "CI_Upper", "Width"), digits = 3)
```

```{r}
#| label: plot-bayesian-posteriors-hiv
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Posterior distributions showing impact of unmeasured confounding"

# Create a data frame with the posterior samples
all_posteriors_hiv <- data.frame(
  `Naive (Severely Biased)` = post_naive_hiv$bHIV,
  `Best Available (Residually Biased)` = post_best_hiv$bHIV,
  `Oracle (Unbiased)` = post_oracle_hiv$bHIV,
  check.names = FALSE
)

# Convert to long format for plotting
long_posteriors_hiv <- all_posteriors_hiv %>%
  pivot_longer(cols = everything(),
               names_to = "Model",
               values_to = "Effect_Estimate")

# Set factor levels for consistent ordering
long_posteriors_hiv$Model <- factor(long_posteriors_hiv$Model,
                                levels = c("Naive (Severely Biased)", "Best Available (Residually Biased)", 
                                          "Oracle (Unbiased)"))

# Plot density curves for all models
ggplot(long_posteriors_hiv, aes(x = Effect_Estimate, fill = Model)) +
  geom_density(alpha = 0.6) +
  geom_vline(data = bayesian_results_hiv,
             aes(xintercept = Mean, color = Model),
             linetype = "dashed", size = 1) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Posterior Distributions Showing Impact of Unmeasured Confounding",
    subtitle = "Oracle model shows unbiased estimate; best available model shows residual bias",
    x = "Treatment Effect (Standardized)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Interpretation of Bayesian Analysis:

The Bayesian analysis clearly demonstrates the impact of unmeasured confounding:

1. **Residual bias persists**: Even the best available model shows biased posterior distributions
2. **Oracle model superiority**: Only the oracle model provides unbiased posterior estimates
3. **Uncertainty quantification**: Bayesian analysis shows that even biased estimates can have narrow credible intervals
4. **Systematic bias**: The bias is systematic, not just increased uncertainty

## 18. Counterfactual Analysis: Treatment Effect Estimation with Unmeasured Confounding

```{r}
#| label: counterfactual-analysis-hiv
#| fig-cap: "Counterfactual predictions under different treatment levels with unmeasured confounding"

# Function to predict Stroke based on different HIV levels (using best available model)
predict_counterfactual_hiv <- function(hiv_values, data = observed_hiv_data) {
  # Use the best available model (biased due to unmeasured confounding)
  model <- lm(Stroke ~ HIV + Smoking + Age, data = data)
  
  # Use mean values for confounders
  mean_smoking <- mean(data$Smoking)
  mean_age <- mean(data$Age)
  
  # Predict Stroke for different values of HIV
  stroke_pred <- predict(model, newdata = data.frame(
    HIV = hiv_values,
    Smoking = mean_smoking,
    Age = mean_age
  ))
  
  return(stroke_pred)
}

# Function for oracle predictions
predict_counterfactual_oracle <- function(hiv_values, data = complete_hiv_data) {
  # Use the oracle model (unbiased)
  model <- lm(Stroke ~ HIV + Smoking + Age + Unmeasured, data = data)
  
  # Use mean values for all confounders
  mean_smoking <- mean(data$Smoking)
  mean_age <- mean(data$Age)
  mean_unmeasured <- mean(data$Unmeasured)
  
  # Predict Stroke for different values of HIV
  stroke_pred <- predict(model, newdata = data.frame(
    HIV = hiv_values,
    Smoking = mean_smoking,
    Age = mean_age,
    Unmeasured = mean_unmeasured
  ))
  
  return(stroke_pred)
}

# Create a range of HIV values for intervention
hiv_range <- seq(min(observed_hiv_data$HIV), max(observed_hiv_data$HIV), length.out = 100)

# Predict Stroke for different HIV levels
stroke_pred_best <- predict_counterfactual_hiv(hiv_range)
stroke_pred_oracle <- predict_counterfactual_oracle(hiv_range)

# Create data frames for plotting
counterfactual_best_df <- data.frame(HIV = hiv_range, Stroke = stroke_pred_best)
counterfactual_oracle_df <- data.frame(HIV = hiv_range, Stroke = stroke_pred_oracle)

# Also show naive prediction for comparison
naive_pred <- predict(hiv_models[["Naive"]], 
                     newdata = data.frame(HIV = hiv_range))
counterfactual_naive_df <- data.frame(HIV = hiv_range, Stroke = naive_pred)

# Plot the counterfactual predictions
ggplot() +
  # Add actual data points
  geom_point(data = observed_hiv_data, aes(x = HIV, y = Stroke), alpha = 0.2, color = "gray") +
  # Add naive prediction (severely biased)
  geom_line(data = counterfactual_naive_df, aes(x = HIV, y = Stroke),
            color = "red", size = 1.5, linetype = "dotted") +
  # Add best available prediction (residually biased)
  geom_line(data = counterfactual_best_df, aes(x = HIV, y = Stroke),
            color = "orange", size = 1.5, linetype = "dashed") +
  # Add oracle counterfactual prediction (unbiased)
  geom_line(data = counterfactual_oracle_df, aes(x = HIV, y = Stroke),
            color = "blue", size = 1.5) +
  labs(
    title = "Counterfactual Predictions: Impact of Unmeasured Confounding",
    subtitle = "Blue=Oracle (unbiased), Orange=Best Available (biased), Red=Naive (severely biased)",
    x = "HIV Status",
    y = "Stroke Risk"
  ) +
  theme_minimal()
```

```{r}
#| label: treatment-effect-table-hiv
#| tbl-cap: "Estimated treatment effects under different interventions with unmeasured confounding"

# Calculate treatment effects for specific interventions
intervention_effects_hiv <- data.frame(
  Intervention = c(
    "Increase HIV exposure by 1 unit",
    "Increase HIV exposure by 2 units",
    "Move from 25th to 75th percentile",
    "Move from minimum to maximum HIV exposure"
  ),
  HIV_Change = c(
    1,
    2,
    quantile(observed_hiv_data$HIV, 0.75) - quantile(observed_hiv_data$HIV, 0.25),
    max(observed_hiv_data$HIV) - min(observed_hiv_data$HIV)
  ),
  Naive_Effect = c(
    1 * coef(hiv_models[["Naive"]])["HIV"],
    2 * coef(hiv_models[["Naive"]])["HIV"],
    (quantile(observed_hiv_data$HIV, 0.75) - quantile(observed_hiv_data$HIV, 0.25)) * coef(hiv_models[["Naive"]])["HIV"],
    (max(observed_hiv_data$HIV) - min(observed_hiv_data$HIV)) * coef(hiv_models[["Naive"]])["HIV"]
  ),
  Best_Available_Effect = c(
    1 * coef(hiv_models[["Best_Available"]])["HIV"],
    2 * coef(hiv_models[["Best_Available"]])["HIV"],
    (quantile(observed_hiv_data$HIV, 0.75) - quantile(observed_hiv_data$HIV, 0.25)) * coef(hiv_models[["Best_Available"]])["HIV"],
    (max(observed_hiv_data$HIV) - min(observed_hiv_data$HIV)) * coef(hiv_models[["Best_Available"]])["HIV"]
  ),
  Oracle_Effect = c(
    1 * coef(oracle_hiv_model)["HIV"],
    2 * coef(oracle_hiv_model)["HIV"],
    (quantile(observed_hiv_data$HIV, 0.75) - quantile(observed_hiv_data$HIV, 0.25)) * coef(oracle_hiv_model)["HIV"],
    (max(observed_hiv_data$HIV) - min(observed_hiv_data$HIV)) * coef(oracle_hiv_model)["HIV"]
  )
)

# Format the results
intervention_effects_hiv$HIV_Change <- round(intervention_effects_hiv$HIV_Change, 3)
intervention_effects_hiv$Naive_Effect <- round(intervention_effects_hiv$Naive_Effect, 3)
intervention_effects_hiv$Best_Available_Effect <- round(intervention_effects_hiv$Best_Available_Effect, 3)
intervention_effects_hiv$Oracle_Effect <- round(intervention_effects_hiv$Oracle_Effect, 3)

datatable(intervention_effects_hiv,
          caption = "Expected outcomes under different HIV interventions (naive vs best available vs oracle)",
          options = list(pageLength = 10, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Counterfactual Analysis:

The counterfactual analysis demonstrates the practical implications of unmeasured confounding:

1. **Biased intervention predictions**: Both naive and best available analyses provide incorrect intervention effect estimates
2. **Oracle benchmark**: Only the oracle model provides accurate counterfactual predictions
3. **Policy implications**: Unmeasured confounding leads to systematically incorrect policy recommendations
4. **Residual bias significance**: Even careful adjustment for measured confounders leaves substantial bias

## 19. Sensitivity Analysis: Quantifying Unmeasured Confounding Impact

```{r}
#| label: sensitivity-analysis-detailed-hiv
#| fig-cap: "Detailed sensitivity analysis for unmeasured confounding"

# Function to simulate varying strengths of unmeasured confounding
simulate_varying_unmeasured_confounding <- function(u_smoking_strength, u_stroke_strength) {
  set.seed(123)
  n <- 1000
  
  # Generate unmeasured confounder U with varying strength
  U <- rnorm(n, mean = 0, sd = 1)
  B_age <- rnorm(n, mean = 0, sd = 1)
  
  # Generate variables with varying unmeasured confounding strength
  A_smoking <- 0.3 * B_age + u_smoking_strength * U + rnorm(n, mean = 0, sd = 0.8)
  X_hiv <- 0.4 * A_smoking + 0.2 * B_age + rnorm(n, mean = 0, sd = 0.7)
  Y_stroke <- 0.5 * X_hiv + 0.7 * A_smoking + 0.3 * B_age + u_stroke_strength * U + rnorm(n, mean = 0, sd = 0.6)
  
  # Estimate the apparent effect with measured confounders only
  temp_data <- data.frame(Stroke = Y_stroke, HIV = X_hiv, Smoking = A_smoking, Age = B_age)
  apparent_effect <- coef(lm(Stroke ~ HIV + Smoking + Age, data = temp_data))[2]
  
  return(apparent_effect)
}

# Test different levels of unmeasured confounding
u_strengths <- seq(0, 1.0, by = 0.1)
sensitivity_grid_detailed <- expand.grid(
  U_Smoking_Strength = u_strengths,
  U_Stroke_Strength = u_strengths
)

sensitivity_grid_detailed$Apparent_Effect <- mapply(simulate_varying_unmeasured_confounding, 
                                          sensitivity_grid_detailed$U_Smoking_Strength,
                                          sensitivity_grid_detailed$U_Stroke_Strength)

# Create sensitivity results table for key scenarios
sensitivity_scenarios_detailed <- data.frame(
  Scenario = c("No unmeasured confounding", "Weak unmeasured confounding", 
               "Moderate unmeasured confounding", "Strong unmeasured confounding",
               "Very strong unmeasured confounding"),
  U_Smoking_Effect = c(0.0, 0.2, 0.5, 0.8, 1.0),
  U_Stroke_Effect = c(0.0, 0.2, 0.5, 0.8, 1.0),
  Apparent_Effect = c(
    simulate_varying_unmeasured_confounding(0.0, 0.0),
    simulate_varying_unmeasured_confounding(0.2, 0.2),
    simulate_varying_unmeasured_confounding(0.5, 0.5),
    simulate_varying_unmeasured_confounding(0.8, 0.8),
    simulate_varying_unmeasured_confounding(1.0, 1.0)
  ),
  Bias_from_True = c(
    simulate_varying_unmeasured_confounding(0.0, 0.0) - true_hiv_effect,
    simulate_varying_unmeasured_confounding(0.2, 0.2) - true_hiv_effect,
    simulate_varying_unmeasured_confounding(0.5, 0.5) - true_hiv_effect,
    simulate_varying_unmeasured_confounding(0.8, 0.8) - true_hiv_effect,
    simulate_varying_unmeasured_confounding(1.0, 1.0) - true_hiv_effect
  )
)

# Format results
sensitivity_scenarios_detailed$Apparent_Effect <- round(sensitivity_scenarios_detailed$Apparent_Effect, 3)
sensitivity_scenarios_detailed$Bias_from_True <- round(sensitivity_scenarios_detailed$Bias_from_True, 3)

datatable(sensitivity_scenarios_detailed,
          caption = "Detailed Sensitivity Analysis: Impact of Varying Unmeasured Confounding Strength",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

```{r}
#| label: sensitivity-heatmap-detailed-hiv
#| fig-cap: "Detailed sensitivity heatmap showing bias from unmeasured confounding"

# Create heatmap of detailed sensitivity analysis
sensitivity_wide_detailed <- sensitivity_grid_detailed %>%
  mutate(Bias = Apparent_Effect - true_hiv_effect) %>%
  select(U_Smoking_Strength, U_Stroke_Strength, Bias)

ggplot(sensitivity_wide_detailed, aes(x = U_Smoking_Strength, y = U_Stroke_Strength, fill = Bias)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Detailed Sensitivity to Unmeasured Confounding",
       subtitle = "Bias in HIV effect estimate across different confounding strengths",
       x = "Unmeasured Confounder Effect on Smoking",
       y = "Unmeasured Confounder Effect on Stroke",
       fill = "Bias") +
  theme_minimal()
```

## Conclusions from Detailed Sensitivity Analysis:

The detailed sensitivity analysis reveals the vulnerability to unmeasured confounding:

1. **Non-linear bias increases**: Bias increases non-linearly with unmeasured confounding strength
2. **Both pathways matter**: Effects on both smoking and stroke contribute to bias
3. **Substantial bias possible**: Even moderate unmeasured confounding can create substantial bias
4. **Robustness assessment**: This analysis helps assess how robust conclusions are to unmeasured confounding



## 20. Mediation Analysis: Decomposing Effects with Unmeasured Confounding

```{r}
#| label: mediation-analysis-hiv
#| tbl-cap: "Mediation analysis with unmeasured confounding (observed pathways only)"

# Test mediation pathways using observed variables only (biased by unmeasured confounding)
# Path 1: Age → HIV → Stroke
age_hiv_model_hiv <- lm(HIV ~ Age, data = observed_hiv_data)
age_hiv_stroke_model_hiv <- lm(Stroke ~ HIV + Age, data = observed_hiv_data)

# Path 2: Age → Smoking → Stroke  
age_smoking_model_hiv <- lm(Smoking ~ Age, data = observed_hiv_data)
age_smoking_stroke_model_hiv <- lm(Stroke ~ Smoking + Age, data = observed_hiv_data)

# Path 3: Age → Smoking → HIV → Stroke
smoking_hiv_model_hiv <- lm(HIV ~ Smoking, data = observed_hiv_data)
smoking_hiv_stroke_model_hiv <- lm(Stroke ~ HIV + Smoking, data = observed_hiv_data)

# Calculate mediation effects (biased due to unmeasured confounding)
mediation_results_hiv <- data.frame(
  Pathway = c(
    "Age → HIV (a path)",
    "HIV → Stroke|Age (b path)", 
    "Age → HIV → Stroke (indirect)",
    "Age → Stroke (direct)",
    "Age → Smoking (a path)",
    "Smoking → Stroke|Age (b path)",
    "Age → Smoking → Stroke (indirect)",
    "Smoking → HIV (a path)",
    "HIV → Stroke|Smoking (b path)",
    "Smoking → HIV → Stroke (indirect)"
  ),
  Effect_Observed = c(
    coef(age_hiv_model_hiv)["Age"],
    coef(age_hiv_stroke_model_hiv)["HIV"],
    coef(age_hiv_model_hiv)["Age"] * coef(age_hiv_stroke_model_hiv)["HIV"],
    coef(age_hiv_stroke_model_hiv)["Age"],
    coef(age_smoking_model_hiv)["Age"],
    coef(age_smoking_stroke_model_hiv)["Smoking"],
    coef(age_smoking_model_hiv)["Age"] * coef(age_smoking_stroke_model_hiv)["Smoking"],
    coef(smoking_hiv_model_hiv)["Smoking"],
    coef(smoking_hiv_stroke_model_hiv)["HIV"],
    coef(smoking_hiv_model_hiv)["Smoking"] * coef(smoking_hiv_stroke_model_hiv)["HIV"]
  ),
  Bias_Status = c(
    "Unbiased (direct effect of age)",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding affecting smoking",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding affecting smoking",
    "Biased by unmeasured confounding",
    "Biased by unmeasured confounding"
  )
)

# Format results
mediation_results_hiv$Effect_Observed <- round(mediation_results_hiv$Effect_Observed, 3)

datatable(mediation_results_hiv,
          caption = "Mediation Analysis with Unmeasured Confounding (Observed Pathways Only)",
          options = list(pageLength = 12, scrollX = TRUE),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Mediation Analysis:

The mediation analysis reveals how unmeasured confounding affects pathway decomposition:

1. **Biased pathway estimates**: Most mediation effects are biased due to unmeasured confounding
2. **Complex bias propagation**: Bias propagates through multiple pathways creating compound effects
3. **Direct effects least biased**: Some direct effects (like age on HIV) may be less affected by unmeasured confounding
4. **Interpretation challenges**: Mediation analysis becomes much more challenging with unmeasured confounding

## 21. Forest Plot Visualization of Treatment Effects

```{r}
#| label: forest-plot-hiv
#| fig-cap: "Forest plot of HIV effect estimates across different modeling approaches with unmeasured confounding"

# Create forest plot data from our comparison table
forest_data_hiv <- hiv_comparison_df %>%
  filter(Model != "True Causal Effect") %>%
  mutate(
    Model = factor(Model, levels = rev(c(
      "Naive (Biased)",
      "Age Adjusted",
      "Smoking Adjusted", 
      "Best Available (Age + Smoking)",
      "Oracle (Age + Smoking + Unmeasured)"
    ))),
    CI_Lower = HIV_Coefficient - 1.96 * Standard_Error,
    CI_Upper = HIV_Coefficient + 1.96 * Standard_Error
  )

# Plot forest plot
ggplot(forest_data_hiv, aes(x = HIV_Coefficient, y = Model,
                           xmin = CI_Lower, xmax = CI_Upper)) +
  geom_pointrange(size = 0.8, color = "darkgreen") +
  geom_vline(xintercept = true_hiv_effect, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "HIV Effect Estimates: Impact of Unmeasured Confounding",
    subtitle = "Dashed line represents the true causal effect (0.5)",
    x = "Estimated HIV Effect on Stroke",
    y = "Modeling Approach"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))
```

## Conclusions from Forest Plot:

The forest plot illustrates the challenge of unmeasured confounding:

1. **Persistent bias**: Even the best available model shows bias compared to the oracle model
2. **Oracle superiority**: Only the oracle model provides unbiased estimates close to the true effect
3. **False precision**: Biased estimates can have narrow confidence intervals, providing false confidence
4. **Need for sensitivity analysis**: The gap between best available and oracle estimates highlights the need for sensitivity analysis

## 22. Simpson's Paradox Detection Analysis

Simpson's Paradox occurs when the direction of an association reverses when data is aggregated versus when examined within subgroups. With unmeasured confounding present, we need to test whether this phenomenon occurs in our data.

### 22.1 Testing for Simpson's Paradox by Smoking Status

```{r}
#| label: simpsons-paradox-smoking-hiv
#| tbl-cap: "Testing for Simpson's Paradox: HIV-Stroke relationship by Smoking strata with unmeasured confounding"

# Create smoking categories
observed_hiv_data <- observed_hiv_data %>%
  mutate(
    Smoking_category = case_when(
      Smoking <= quantile(Smoking, 0.33) ~ "Low Smoker",
      Smoking <= quantile(Smoking, 0.67) ~ "Moderate Smoker", 
      TRUE ~ "Heavy Smoker"
    )
  )

# Calculate correlations overall and within smoking strata
overall_corr_hiv <- cor(observed_hiv_data$HIV, observed_hiv_data$Stroke)

smoking_strata_analysis_hiv <- observed_hiv_data %>%
  group_by(Smoking_category) %>%
  summarise(
    n = n(),
    HIV_mean = round(mean(HIV), 3),
    Stroke_mean = round(mean(Stroke), 3),
    HIV_Stroke_correlation = round(cor(HIV, Stroke), 3),
    HIV_coefficient = round(coef(lm(Stroke ~ HIV))[2], 3),
    HIV_pvalue = round(summary(lm(Stroke ~ HIV))$coefficients[2,4], 3),
    .groups = 'drop'
  )

# Add overall results for comparison
overall_results_hiv <- data.frame(
  Smoking_category = "Overall (All)",
  n = nrow(observed_hiv_data),
  HIV_mean = round(mean(observed_hiv_data$HIV), 3),
  Stroke_mean = round(mean(observed_hiv_data$Stroke), 3),
  HIV_Stroke_correlation = round(overall_corr_hiv, 3),
  HIV_coefficient = round(coef(lm(Stroke ~ HIV, data = observed_hiv_data))[2], 3),
  HIV_pvalue = round(summary(lm(Stroke ~ HIV, data = observed_hiv_data))$coefficients[2,4], 3)
)

# Combine results
simpsons_smoking_df_hiv <- rbind(overall_results_hiv, smoking_strata_analysis_hiv)

datatable(simpsons_smoking_df_hiv,
          caption = "Simpson's Paradox Test: HIV-Stroke Relationship by Smoking Status (Unmeasured Confounding Present)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')

# Test for paradox direction
smoking_paradox_test_hiv <- data.frame(
  Comparison = c("Overall vs Low Smoker", "Overall vs Moderate Smoker", "Overall vs Heavy Smoker"),
  Overall_Direction = c("Positive", "Positive", "Positive"),
  Strata_Direction = c(
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[1] > 0, "Positive", "Negative"),
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[2] > 0, "Positive", "Negative"),
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[3] > 0, "Positive", "Negative")
  ),
  Paradox_Present = c(
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[1] < 0, "YES", "NO"),
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[2] < 0, "YES", "NO"),
    ifelse(smoking_strata_analysis_hiv$HIV_coefficient[3] < 0, "YES", "NO")
  )
)

datatable(smoking_paradox_test_hiv,
          caption = "Simpson's Paradox Direction Test by Smoking (With Unmeasured Confounding)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

### 22.2 Testing for Simpson's Paradox by Age Groups

```{r}
#| label: simpsons-paradox-age-hiv
#| tbl-cap: "Testing for Simpson's Paradox: HIV-Stroke relationship by Age strata with unmeasured confounding"

# Calculate correlations overall and within age strata
age_strata_analysis_hiv <- observed_hiv_data %>%
  group_by(Age_Group) %>%
  summarise(
    n = n(),
    HIV_mean = round(mean(HIV), 3),
    Stroke_mean = round(mean(Stroke), 3),
    HIV_Stroke_correlation = round(cor(HIV, Stroke), 3),
    HIV_coefficient = round(coef(lm(Stroke ~ HIV))[2], 3),
    HIV_pvalue = round(summary(lm(Stroke ~ HIV))$coefficients[2,4], 3),
    .groups = 'drop'
  )

# Combine with overall results
overall_results_age_hiv <- data.frame(
  Age_Group = "Overall (All)",
  n = nrow(observed_hiv_data),
  HIV_mean = round(mean(observed_hiv_data$HIV), 3),
  Stroke_mean = round(mean(observed_hiv_data$Stroke), 3),
  HIV_Stroke_correlation = round(overall_corr_hiv, 3),
  HIV_coefficient = round(coef(lm(Stroke ~ HIV, data = observed_hiv_data))[2], 3),
  HIV_pvalue = round(summary(lm(Stroke ~ HIV, data = observed_hiv_data))$coefficients[2,4], 3)
)

simpsons_age_df_hiv <- rbind(overall_results_age_hiv, age_strata_analysis_hiv)

datatable(simpsons_age_df_hiv,
          caption = "Simpson's Paradox Test: HIV-Stroke Relationship by Age Group (Unmeasured Confounding Present)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')

# Test for paradox direction
age_paradox_test_hiv <- data.frame(
  Comparison = c("Overall vs Younger", "Overall vs Middle Age", "Overall vs Older"),
  Overall_Direction = c("Positive", "Positive", "Positive"),
  Strata_Direction = c(
    ifelse(age_strata_analysis_hiv$HIV_coefficient[1] > 0, "Positive", "Negative"),
    ifelse(age_strata_analysis_hiv$HIV_coefficient[2] > 0, "Positive", "Negative"),
    ifelse(age_strata_analysis_hiv$HIV_coefficient[3] > 0, "Positive", "Negative")
  ),
  Paradox_Present = c(
    ifelse(age_strata_analysis_hiv$HIV_coefficient[1] < 0, "YES", "NO"),
    ifelse(age_strata_analysis_hiv$HIV_coefficient[2] < 0, "YES", "NO"),
    ifelse(age_strata_analysis_hiv$HIV_coefficient[3] < 0, "YES", "NO")
  )
)

datatable(age_paradox_test_hiv,
          caption = "Simpson's Paradox Direction Test by Age (With Unmeasured Confounding)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

### 22.3 Testing for Simpson's Paradox by Unmeasured Confounder (Oracle Analysis)

```{r}
#| label: simpsons-paradox-unmeasured-oracle-hiv
#| tbl-cap: "Testing for Simpson's Paradox: HIV-Stroke relationship by Unmeasured Confounder strata"

# Calculate correlations overall and within unmeasured confounder strata
overall_corr_complete_hiv <- cor(complete_hiv_data$HIV, complete_hiv_data$Stroke)

unmeasured_strata_analysis_hiv <- complete_hiv_data %>%
  group_by(Unmeasured_Level) %>%
  summarise(
    n = n(),
    HIV_mean = round(mean(HIV), 3),
    Stroke_mean = round(mean(Stroke), 3),
    HIV_Stroke_correlation = round(cor(HIV, Stroke), 3),
    HIV_coefficient = round(coef(lm(Stroke ~ HIV))[2], 3),
    HIV_pvalue = round(summary(lm(Stroke ~ HIV))$coefficients[2,4], 3),
    .groups = 'drop'
  )

# Add overall results for comparison
overall_results_complete_hiv <- data.frame(
  Unmeasured_Level = "Overall (All)",
  n = nrow(complete_hiv_data),
  HIV_mean = round(mean(complete_hiv_data$HIV), 3),
  Stroke_mean = round(mean(complete_hiv_data$Stroke), 3),
  HIV_Stroke_correlation = round(overall_corr_complete_hiv, 3),
  HIV_coefficient = round(coef(lm(Stroke ~ HIV, data = complete_hiv_data))[2], 3),
  HIV_pvalue = round(summary(lm(Stroke ~ HIV, data = complete_hiv_data))$coefficients[2,4], 3)
)

# Combine results
simpsons_unmeasured_df_hiv <- rbind(overall_results_complete_hiv, unmeasured_strata_analysis_hiv)

datatable(simpsons_unmeasured_df_hiv,
          caption = "Simpson's Paradox Test: HIV-Stroke Relationship by Unmeasured Confounder Strata (Oracle Analysis)",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')

# Test for paradox direction
unmeasured_paradox_test_hiv <- data.frame(
  Comparison = c("Overall vs Low Hidden", "Overall vs Medium Hidden", "Overall vs High Hidden"),
  Overall_Direction = c("Positive", "Positive", "Positive"),
  Strata_Direction = c(
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[1] > 0, "Positive", "Negative"),
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[2] > 0, "Positive", "Negative"),
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[3] > 0, "Positive", "Negative")
  ),
  Paradox_Present = c(
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[1] < 0, "YES", "NO"),
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[2] < 0, "YES", "NO"),
    ifelse(unmeasured_strata_analysis_hiv$HIV_coefficient[3] < 0, "YES", "NO")
  )
)

datatable(unmeasured_paradox_test_hiv,
          caption = "Simpson's Paradox Direction Test by Unmeasured Confounder",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

### 22.4 Visual Detection of Simpson's Paradox

```{r}
#| label: simpsons-paradox-visualization-hiv
#| fig-cap: "Visual detection of Simpson's Paradox with unmeasured confounding"
#| fig-subcap: 
#|   - "Overall vs Smoking Strata Relationships (Unmeasured Confounding Present)"
#|   - "Overall vs Age Strata Relationships (Unmeasured Confounding Present)"
#|   - "Overall vs Unmeasured Confounder Strata (Oracle Analysis)"
#| layout-ncol: 1

# Plot 1: Simpson's Paradox by Smoking (with unmeasured confounding)
p1_simpson_hiv <- ggplot(observed_hiv_data, aes(x = HIV, y = Stroke)) +
  # Overall relationship
  geom_smooth(aes(color = "Overall"), method = "lm", se = TRUE, size = 2) +
  # Stratified relationships
  geom_smooth(aes(color = Smoking_category), method = "lm", se = FALSE, size = 1) +
  geom_point(aes(color = Smoking_category), alpha = 0.3) +
  scale_color_manual(values = c("Overall" = "black", "Low Smoker" = "blue", 
                               "Moderate Smoker" = "green", "Heavy Smoker" = "red")) +
  labs(title = "Testing Simpson's Paradox: HIV-Stroke by Smoking Status",
       subtitle = "Black line shows overall relationship (biased by unmeasured confounding)",
       x = "HIV Status", y = "Stroke Risk", color = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p1_simpson_hiv)

# Plot 2: Simpson's Paradox by Age (with unmeasured confounding)
p2_simpson_hiv <- ggplot(observed_hiv_data, aes(x = HIV, y = Stroke)) +
  # Overall relationship
  geom_smooth(aes(color = "Overall"), method = "lm", se = TRUE, size = 2) +
  # Stratified relationships
  geom_smooth(aes(color = Age_Group), method = "lm", se = FALSE, size = 1) +
  geom_point(aes(color = Age_Group), alpha = 0.3) +
  scale_color_manual(values = c("Overall" = "black", "Younger" = "purple", 
                               "Middle Age" = "orange", "Older" = "brown")) +
  labs(title = "Testing Simpson's Paradox: HIV-Stroke by Age Group",
       subtitle = "Black line shows overall relationship (biased by unmeasured confounding)",
       x = "HIV Status", y = "Stroke Risk", color = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p2_simpson_hiv)

# Plot 3: Simpson's Paradox by Unmeasured Confounder (Oracle)
p3_simpson_hiv <- ggplot(complete_hiv_data, aes(x = HIV, y = Stroke)) +
  # Overall relationship
  geom_smooth(aes(color = "Overall"), method = "lm", se = TRUE, size = 2) +
  # Stratified relationships
  geom_smooth(aes(color = Unmeasured_Level), method = "lm", se = FALSE, size = 1) +
  geom_point(aes(color = Unmeasured_Level), alpha = 0.3) +
  scale_color_manual(values = c("Overall" = "black", "Low Hidden" = "cyan", 
                               "Medium Hidden" = "magenta", "High Hidden" = "yellow")) +
  labs(title = "Testing Simpson's Paradox: HIV-Stroke by Unmeasured Confounder",
       subtitle = "Oracle analysis showing impact of unmeasured confounding on relationships",
       x = "HIV Status", y = "Stroke Risk", color = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p3_simpson_hiv)
```

### 22.5 Formal Statistical Test for Simpson's Paradox

```{r}
#| label: simpsons-paradox-formal-test-hiv
#| tbl-cap: "Formal statistical test for Simpson's Paradox with unmeasured confounding"

# Function to test for Simpson's Paradox
test_simpsons_paradox_hiv <- function(data, strata_var, exposure_var, outcome_var) {
  # Overall correlation
  overall_cor <- cor(data[[exposure_var]], data[[outcome_var]])
  overall_coef <- coef(lm(data[[outcome_var]] ~ data[[exposure_var]]))[2]
  
  # Within-strata correlations and coefficients
  strata_stats <- data %>%
    group_by(!!sym(strata_var)) %>%
    summarise(
      n = n(),
      correlation = cor(!!sym(exposure_var), !!sym(outcome_var)),
      coefficient = coef(lm(!!sym(outcome_var) ~ !!sym(exposure_var)))[2],
      .groups = 'drop'
    )
  
  # Test for direction reversal
  overall_direction <- sign(overall_coef)
  strata_directions <- sign(strata_stats$coefficient)
  
  # Count reversals
  reversals <- sum(strata_directions != overall_direction)
  total_strata <- nrow(strata_stats)
  
  # Test for significant difference in slopes
  # Using interaction test
  interaction_model <- lm(paste(outcome_var, "~", exposure_var, "*", strata_var), data = data)
  interaction_p <- summary(interaction_model)$coefficients[grep(":", rownames(summary(interaction_model)$coefficients)), 4]
  
  return(list(
    overall_coefficient = overall_coef,
    strata_coefficients = strata_stats$coefficient,
    reversals = reversals,
    total_strata = total_strata,
    proportion_reversed = reversals / total_strata,
    interaction_p_values = interaction_p,
    paradox_detected = any(interaction_p < 0.05) && reversals > 0
  ))
}

# Test for Simpson's Paradox by Smoking (with unmeasured confounding)
smoking_paradox_result_hiv <- test_simpsons_paradox_hiv(observed_hiv_data, "Smoking_category", "HIV", "Stroke")

# Test for Simpson's Paradox by Age (with unmeasured confounding)
age_paradox_result_hiv <- test_simpsons_paradox_hiv(observed_hiv_data, "Age_Group", "HIV", "Stroke")

# Test for Simpson's Paradox by Unmeasured Confounder (Oracle)
unmeasured_paradox_result_hiv <- test_simpsons_paradox_hiv(complete_hiv_data, "Unmeasured_Level", "HIV", "Stroke")

# Create summary table
paradox_summary_hiv <- data.frame(
  Stratification_Variable = c("Smoking Status", "Age Group", "Unmeasured Confounder"),
  Overall_Coefficient = c(smoking_paradox_result_hiv$overall_coefficient, 
                         age_paradox_result_hiv$overall_coefficient,
                         unmeasured_paradox_result_hiv$overall_coefficient),
  Direction_Reversals = c(smoking_paradox_result_hiv$reversals, 
                         age_paradox_result_hiv$reversals,
                         unmeasured_paradox_result_hiv$reversals),
  Total_Strata = c(smoking_paradox_result_hiv$total_strata, 
                  age_paradox_result_hiv$total_strata,
                  unmeasured_paradox_result_hiv$total_strata),
  Proportion_Reversed = c(smoking_paradox_result_hiv$proportion_reversed, 
                         age_paradox_result_hiv$proportion_reversed,
                         unmeasured_paradox_result_hiv$proportion_reversed),
  Interaction_Significant = c(
    any(smoking_paradox_result_hiv$interaction_p_values < 0.05),
    any(age_paradox_result_hiv$interaction_p_values < 0.05),
    any(unmeasured_paradox_result_hiv$interaction_p_values < 0.05)
  ),
  Simpsons_Paradox_Detected = c(smoking_paradox_result_hiv$paradox_detected, 
                                age_paradox_result_hiv$paradox_detected,
                                unmeasured_paradox_result_hiv$paradox_detected)
)

# Format results
paradox_summary_hiv$Overall_Coefficient <- round(paradox_summary_hiv$Overall_Coefficient, 3)
paradox_summary_hiv$Proportion_Reversed <- round(paradox_summary_hiv$Proportion_Reversed, 3)

datatable(paradox_summary_hiv,
          caption = "Formal Test Results for Simpson's Paradox with Unmeasured Confounding",
          options = list(pageLength = 10, dom = 't'),
          rownames = FALSE,
          class = 'cell-border stripe compact responsive')
```

## Conclusions from Simpson's Paradox Analysis:

The Simpson's Paradox analysis with unmeasured confounding reveals important insights:

1. **No complete paradox detected**: While variation exists across strata, we don't observe complete direction reversal in our data structure
2. **Unmeasured confounding complicates interpretation**: The presence of unmeasured confounding makes it difficult to determine whether observed differences reflect true Simpson's Paradox or bias
3. **Oracle analysis provides clarity**: Stratification by the unmeasured confounder shows how hidden variables can create apparent paradoxes
4. **Methodological implications**: This highlights why unmeasured confounding makes causal inference so challenging in observational studies


## 23. Practical Implications and Conclusions

### 23.1 Summary of Key Findings

Our comprehensive analysis of the HIV-Stroke causal structure with unmeasured confounding has revealed critical insights about the limitations of observational studies:

1. **Unmeasured confounding creates persistent bias**: Even careful adjustment for all measured confounders leaves substantial residual bias
2. **Oracle superiority demonstrates the problem**: Only the impossible-to-achieve oracle model provides unbiased estimates
3. **Multiple analytical approaches confirm**: Frequentist, Bayesian, SEM, and sensitivity analyses all show consistent bias patterns
4. **Magnitude of bias can be substantial**: Unmeasured confounding can create bias of 20-50% or more of the true effect

### 23.2 The Fundamental Challenge of Observational Studies

This analysis illustrates why observational studies face inherent limitations that cannot be solved through better statistical methods alone:

**The Unmeasured Confounding Problem:**
- Creates backdoor paths that cannot be blocked by measured variables
- Leads to systematically biased estimates regardless of sample size
- Cannot be eliminated through standard analytical approaches
- Requires strong assumptions about what variables are unmeasured

**Critical Requirements for Valid Inference:**
- Complete knowledge of all relevant confounders (often impossible)
- Accurate measurement of all confounders (measurement error adds bias)
- Correct specification of causal relationships (DAG must be correct)
- Strong exchangeability assumptions (often untestable)

### 23.3 Real-World Applications and Examples

This framework applies to numerous observational study contexts where unmeasured confounding is likely:

- **Medical effectiveness studies**: Comparing treatments in routine clinical care where patient selection factors are incompletely measured
- **Environmental health**: Studying pollution effects where socioeconomic and lifestyle factors are imperfectly measured
- **Social interventions**: Evaluating programs where selection into treatment involves unmeasured motivational factors
- **Economic policy analysis**: Studying interventions where implementation factors are hidden or unmeasured

Think of trying to measure the effect of exercise on heart health. You can measure age, diet, smoking, and medical history, but there are invisible factors you might miss: genetic predisposition, stress levels, sleep quality, environmental exposures, or personality traits that affect both exercise habits and heart health. These unmeasured factors create a "shadow bias" that traditional statistical methods cannot eliminate, like trying to measure the shadow of an object you cannot see.

### 23.4 Comparison with Experimental Studies

Unlike experimental studies where randomization eliminates both measured and unmeasured confounding:

- **Unmeasured confounders remain active** in observational studies
- **Causal identification becomes impossible** without strong assumptions
- **Residual bias persists** regardless of analytical sophistication
- **Sensitivity analysis becomes critical** for assessing robustness

### 23.5 Methodological Recommendations

Based on this analysis, we recommend a multi-pronged approach to observational studies:

1. **Explicit causal modeling**: Use DAGs to make assumptions transparent and identify potential sources of bias
2. **Comprehensive confounder measurement**: Measure as many relevant confounders as possible, but acknowledge limitations
3. **Sensitivity analysis**: Always conduct sensitivity analyses for unmeasured confounding
4. **Multiple analytical approaches**: Use different methods to assess robustness of findings
5. **Transparent uncertainty**: Clearly communicate the limitations and potential biases
6. **External validation**: Seek convergent evidence from multiple study designs when possible

### 23.6 The Role of Sensitivity Analysis

Sensitivity analysis becomes crucial when unmeasured confounding is possible:

- **Quantifies vulnerability**: Shows how much bias different levels of unmeasured confounding would create
- **Guides interpretation**: Helps determine whether conclusions are robust to plausible confounding
- **Informs design**: Identifies critical unmeasured variables to target in future studies
- **Supports transparency**: Makes explicit the assumptions required for causal interpretation

### 23.7 Limitations and Future Directions

Even sophisticated analyses of unmeasured confounding have limitations:

1. **Assumption dependent**: Sensitivity analyses require assumptions about unmeasured confounder strength
2. **Single confounder focus**: Most methods consider one unmeasured confounder, but multiple may exist
3. **Linear relationships**: Most approaches assume linear relationships between variables
4. **Static confounding**: Don't account for time-varying unmeasured confounding

**Future directions include:**
- Methods for multiple unmeasured confounders
- Nonlinear sensitivity analysis approaches
- Integration with quasi-experimental methods
- Machine learning approaches to confounder detection

### 23.8 The Value of DAG Thinking with Unmeasured Confounding

DAG thinking becomes even more critical when unmeasured confounding is possible:

- **Makes assumptions explicit**: Clearly shows what variables are assumed to be unmeasured
- **Identifies bias sources**: Pinpoints pathways through which unmeasured confounding operates
- **Guides sensitivity analysis**: Helps parameterize sensitivity analyses appropriately
- **Facilitates communication**: Provides clear visualization of the assumed causal structure and its limitations

### 23.9 Practical Decision Making Under Uncertainty

When unmeasured confounding is likely, decision makers must:

1. **Acknowledge uncertainty**: Accept that causal estimates may be biased by unknown amounts
2. **Consider bias direction**: Use subject matter knowledge to assess likely direction of bias
3. **Seek convergent evidence**: Look for consistent patterns across multiple study designs
4. **Make explicit trade-offs**: Balance the need for action with uncertainty about causal effects
5. **Plan for learning**: Design systems to gather additional evidence over time

### 23.10 The Analogy of Hidden Forces

Understanding unmeasured confounding is like trying to understand the motion of planets before discovering gravity. Early astronomers could see the planets move and could even predict their positions with complex models, but they couldn't explain why the predictions sometimes failed. The unseen force of gravity was creating patterns they couldn't account for. 

Similarly, in observational studies, we can see relationships between exposures and outcomes, and we can build sophisticated models that seem to work well. But there are often "invisible forces" - unmeasured confounders - that create bias in ways we cannot fully account for. Just as astronomy was revolutionized by recognizing these hidden forces, causal inference is revolutionized by acknowledging the limits imposed by unmeasured confounding.

### 23.11 The Clinical Translation Challenge

In the specific context of HIV and stroke research, our analysis highlights several critical considerations:

**For HIV Care Providers:**
- Recognize that observational studies of HIV-stroke relationships may overestimate or underestimate true risks
- Consider unmeasured factors like adherence patterns, social determinants, or genetic susceptibility
- Use multiple sources of evidence when making treatment decisions
- Acknowledge uncertainty when counseling patients about stroke risk

**For Public Health Policy:**
- Develop policies that are robust to different plausible effect sizes
- Invest in better measurement of previously unmeasured confounders
- Support experimental or quasi-experimental studies when possible
- Create monitoring systems to detect unexpected effects

**For Future Research:**
- Prioritize measurement of previously unmeasured variables (genetic factors, detailed behavioral measures, environmental exposures)
- Develop better methods for handling multiple unmeasured confounders
- Integrate observational findings with experimental evidence when available
- Improve reporting of uncertainty and potential biases




### 23.12 The Broader Implications for Evidence-Based Practice

This analysis has implications beyond HIV and stroke research:

**For Evidence Synthesis:**
- Systematic reviews should explicitly address unmeasured confounding
- Meta-analyses of observational studies should include sensitivity analyses
- Evidence grading systems should account for unmeasured confounding potential
- Clinical guidelines should acknowledge observational study limitations

**For Medical Education:**
- Teach critical appraisal skills that include assessment of unmeasured confounding
- Emphasize the difference between association and causation in complex systems
- Provide training in DAG construction and causal inference methods
- Foster intellectual humility about causal claims from observational data

**For Healthcare Systems:**
- Develop decision-making frameworks that account for uncertainty
- Create systems for updating practices as new evidence emerges
- Balance the need for action with acknowledgment of uncertainty
- Invest in data infrastructure that captures previously unmeasured factors

## Conclusions:

This analysis demonstrates that unmeasured confounding represents a fundamental challenge for causal inference from observational data. While statistical methods can control for measured confounders, they cannot eliminate bias from unmeasured confounders. This reality requires:

- **Humility about causal claims**: Acknowledging the limits of what observational studies can definitively establish
- **Transparency about assumptions**: Making explicit what must be assumed for causal interpretation
- **Sensitivity to bias**: Conducting analyses to assess robustness to unmeasured confounding
- **Multiple evidence sources**: Seeking convergent evidence across different study designs and contexts

The presence of unmeasured confounding doesn't invalidate observational studies, but it does require careful interpretation, explicit acknowledgment of limitations, and appropriate caution in drawing causal conclusions. Understanding these limitations is essential for both researchers conducting studies and decision makers using their results.

The HIV-Stroke relationship examined here, with its complex web of measured and unmeasured confounding, illustrates why observational epidemiology requires both sophisticated analytical methods and intellectual humility. The gap between what we can measure and what truly affects the relationships we study represents one of the most important challenges in modern causal inference.

Only by acknowledging these challenges and developing appropriate methods to address them can we hope to extract valid causal insights from the observational data that dominates much of social science, epidemiology, and policy research. The future of causal inference lies not in pretending we can eliminate all sources of bias, but in becoming more sophisticated about characterizing, quantifying, and communicating the uncertainty that unmeasured confounding creates.

In the end, the goal is not perfect causal knowledge—which may be impossible in complex systems—but rather the best possible evidence to guide decisions under uncertainty. This requires a combination of methodological rigor, transparent communication, and intellectual honesty about what we know, what we don't know, and what we may never be able to know with certainty.

## Session Information for Reproducibility

```{r}
#| label: session-info-hiv-stroke-final
#| echo: false

# Session information for reproducibility
# sessionInfo()

```




